### 2018-03-12 outline.tex
trygve

% !TEX program = pdflatex
\input{preamble/preamble.tex}
\input{preamble/preamble_math.tex}
\input{preamble/preamble_acronyms.tex}

% \linenumbers

\title{Trygve's Problem}
\author{Jackson Loper}

\usepackage{amsthm}
\newtheorem{thm}{Theorem}

\begin{document}
\maketitle

\section{Overview}

When many different measurement tools are available, it can be difficult to understand
how to combine information from different tools.  This challenge is compounded when the measurement
tools are destructive.  For example, after processing a neuron with a given single-cell rna
sequencing technique, we cannot generally go back and process that same neuron with a different
technique.  This is because the measurement technique destroys the original neuron.  This means
we can never see \emph{both measurements} for the same cell, side by side.  How, then, can we ever
hope to understand how the different measurements relate to each other?

In this paper, we hope to shed some light on this this basic mathematical question.  It has many
applications in the hard sciences.  For example:

\begin{itemize}

\item Different tools may teach us different things, and we need to combine tools
to learn everything we can.  Let $A,B$ denote two different quantities of interest about cells (e.g. morphology and gene expression).   Let us say one tool enables us to measure $A$ and another tool to measure $B$.  We would like
to be able to say something about what morphologies are associated with what gene expressions.

\item Two labs may used different methods, but may wish to pool results.  Each lab looks at the other
lab's data and asks themselves "what would their data have looked like if they had used our method?"  When
they find they don't know the answer, it greatly complicates collaboration.

\item One may attempt to cluster the data.  It is simple to cluster the data from one technique, and cluster the data
from another technique -- but how can we know how those clusters correspond to one another?

\end{itemize}

How could we learn the relationship between two measurement techniques, unless we can observe both measurements 
applied to the same members of a population (dual measurement)?  In this work we show that it is indeed 
sometimes possible to get useful and statistically rigorous bounds, even without these dual measurements.

In particular, we consider the case that we can obtain samples from a variety of subpopulations.  
If we can assume that the \emph{relationship} between the measurements is the same in each subpopulation, 
we can use these samples to learn something about that relationship -- even though we never see both 
measurements for any single member of the population.

We then apply our method to single cell RNA data.  We show how different clusterings from different single 
cell RNA modalities can be connected.  Our method yields results which are generally consistent with the scientist's
understanding of the data and clusters, but also reveal potential gaps which could be important for further study.

\section{Mathematical formulation}

Mathematically, we formulate our model as follows:

\begin{itemize}
\item Let us say we have $n+m$ individual members of a population.  These could be cells, humans, or whatever
the smallest sample unit may be for a given problem.
\item For each individual member, we have some readily observable features by which we can group the members,
such as the size of a cell, where an individual lives, or some other basic information.  We designate these
features as $\ell_i$ for the $i$th individual.
\item For the first $n$ members, we have made an observation using the first technique.  This gives us
the values $X_1 \cdots X_n$.
\item The last $m$ members were observed using the second technique.  This gives us the values
$Y_{n+1} \cdots Y_{n+m}$.
\end{itemize}

For simplicity, we here assume that $\ell_i,X_i,Y_i$ take values in finite sets, $\Omega_\ell,\Omega_X,\Omega_Y$.  Extensions to more general
cases should be straightforward, but we must leave them to future work.

We further posit a set of \emph{counterfactual} variables -- variables which could not ever be obtained in practice,
but which help us organize our thinking.  These are sometimes referred to as "potential outcomes"  (cf. \cite{rubin2005causal}).

\begin{itemize}
\item Let $Y_1 \cdots Y_n$ denote the observations we \emph{would have gotten} if we had applied the second technique
to the first $n$ members.
\item Let $X_{n+1} \cdots X_n$ denote the observations we would have gotten if we had applied the first technique
to the last $m$ members. 
\end{itemize}

Our main assumption is that

\begin{equation}
\mathbb{P}(X_i=x,Y_i=y|L_i=\ell)=p^*(x_i|\ell_i)q^*(y_i|x_i)
\end{equation}

for some distributions $p^*,q^*$.  Moreover, we assume that the vectors $\{(p^*(x_1|\ell)\cdots p^*(x_n|\ell))\}_\ell$ 
are linearly independent.  If this seems unlikely, it may make sense to collapse similar values of $\ell$ together.

The validity of these assumptions for a given situation should be closely contemplated.  
There are several key questions to answer
when deciding whether this assumption is applicable.  Does the process by which samples are gathered
gathered depend only upon $\ell$?  In particular, is it not at all statistically related to which measurement technique was 
applied, except through $\ell$?  Are the particular measurement biases of both techniques the same for every value of $\ell$?  
Is the statistical relationship between the quantities being measured by the 
two techniques the same for every value of $\ell$?\footnote{Note that this is 
automatically true if both techniques are measuring the same thing.  More generally, if two techniques have something that
they both measure, we can restrict attention to that one common phenomenon.}  If the answer to all of these questions is yes, the assumption that $(X_i,Y_i)$ is drawn from $p(x_i|\ell_i)p(y_i|x_i)$ may apply.

We can now articulate our goal.  Informally, our goal is to see what we can understand about the relationship between
the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$.  Ideally, we would
like to learn the distribution $p^*(y_i|x_i)$.  Unfortunately, given the limitations of the data, this may not be 
possible.  Instead, we articulate a slightly less amibtious goal: to put \emph{bounds} on $q(y_i|x_i)$.  Specifically,
we will seek to produce functions $\hat u_1(y|x),\hat u_2(y|x)$ such that

\begin{equation}\label{eq:ubound}
\hat u_1(y|x) \leq q^*(y|x) \leq \hat u_2(y|x)
\end{equation}

In particular, we define

\begin{align*}
\Theta^* &\triangleq\left\{q:\ \sum_x p^*(x|\ell) q^*(y|x) =\sum_x p^*(x|\ell)q^*(y|x), \forall \ell,y\right\}\\ 
u_1^*(y|x) &= \inf_{q\in\Theta^*} \{q(y|x)\} \\
u_2^*(y|x) &= \sup_{q\in\Theta^*} \{q(y|x)\}
\end{align*}

Then obviously $u_1^*(y|x) \leq q^*(y|x)\leq u_2^*(y|x)$ is a valid bound.  Moreover, given the data and assumptions
we have, it is as tight a bound as is possible to make (indeed, we could not hope to
distinguish between any two $q_1,q_2 \in \Theta^*$, since they yield the same distributions on what we can
observe).

Thus, our mathematical task to to attempt to find estimators for $u_1^*,u_2^*$, which will allow us to bound
the values of $q^*(y|x)$, and thus understand something about the relationship between the two measurements, $X$ and
$Y$.

\section{Mathematical solution}

To produce useful bounds on $q^*(y|x)$, we first estimate

\begin{itemize}
\item $p^*$ using the empirical distribtion of the observations $\{X_i\}$, $\hat p(x|\ell)$
\item $h^*(y|\ell)\triangleq \sum_x p^*(x|\ell)q^*(y|\ell)$ using the empirical distribtion of $\{Y_i\}$, $\hat h(y|\ell)$
\end{itemize}

These empirical distributions may not be quite consistent with the original assumption that the distribution
of $X,Y|\ell$ may be written as $p^*(x|\ell)q^*(y|x)$.  In particular, it may be that there is \emph{no} value of $q$ such
that $\sum_x \hat p(x|\ell) q(y|x) = \hat h(y|\ell)$.  There may also be many such values.  We satisfy ourselves
with finding a somewhat good fit for $q$.  Let 
$N_{X,\ell}=\#\{i\leq n:\ \ell_i=\ell\},N_{Y,\ell}=\#\{i> n:\ \ell_i=\ell\}$.  We take

\begin{equation}
\hat q = \argmax_q \sum_\ell N_{Y,\ell}\sum_{y}\hat h(y|\ell)\log\left(\sum_{x}\hat p(x|\ell)q(y|x)\right) + \kappa \sum_{xy} \log q(y|x)
\end{equation}

Note in the second term we apply an entropic penalty to $q$.  This is both necessary to ensure 
that $\hat q$ is uniquely defined in terms of $\hat q,\hat h$, as well as a useful regularization.
This regularization pushes us slightly towards uniform distributions, which makes sense if we believe that $q(y|x)>0$ 
for each $x,y$.  Other regularizations could also be used and would potentially
enjoy similar theoretical guarantees to the ones we will show below.

This method gives us an estimate for $\hat q$, but we emphasize this estimate may be \emph{inconsistent}
for the true value of $q^*$.  As we have said, this is simply a limitation of the data we have available -- attempting to deduce
$q^*$ may be too much to ask for.  However, we can get bounds.

The key is to search over all possible values of $q$ which agree with $\hat q$ in terms of what \emph{is} identifiable.
That is, let

\begin{align*}
\hat\Theta &\triangleq\left\{q:\ \sum_x \hat p(x|\ell)\hat q(y|x) =\sum_x \hat p(x|\ell)q(y|x), \forall \ell,y\right\}\\ 
\hat u_1(y|x) &= \inf_{q\in\hat\Theta} \left\{q(y|x) \right\} \\
\hat u_2(y|x) &= \sup_{q\in\hat\Theta} \left\{q(y|x) \right\}
\end{align*}

These will serve as consistent estimators for our quantities of interest, $u_1^*,u_2^*$, as the following theorem shows.\vspace{.1in}

\begin{thm}
Fix any $\kappa>0$.  Let $N_{X,\ell},N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.  Then $\hat u_1(y|x)\rightarrow u^*_1(y|x)$ in probability, and likewise for $u_2$.
\end{thm}

\begin{proof}

The logic is in perfect symmetry for $u_1,u_2$, so we focus on the $u_1$ case.

We begin by seeing how our process would have worked if we were able to know the exact values of $p^*$ and $h^*$,
instead of the empirical distributions $\hat p,\hat h$.  That is, we define
%
\begin{align*}
L_N(q,p,h) &\triangleq \sum_\ell N_{Y,\ell}\sum_{y}h(y|\ell)\log\left(\sum_{x}p(x|\ell)q(y|x)\right) + \kappa \sum_{xy} \log q(y|x) \\
\Theta(p,\tilde q) &\triangleq\left\{q:\ \sum_x p(x|\ell) \tilde q(y|x) =\sum_x p(x|\ell) q(y|x), \forall \ell,y\right\}
\end{align*}
%
and take
%
\begin{align*}
\tilde q^* &= \argmax_q L_N(q,p^*,h^*)  \\
\tilde u_1^*(y|x) &= \inf_{q\in \Theta(p^*,\tilde q^*)} q(y|x) \\
\end{align*}
%
Our first task is to show that the map $p^*,h^* \mapsto \tilde u_1^*$ is continuous.  This follows in two steps:

%%%%% two steps
\begin{enumerate}
% STEP I
\item The map $p^*,h^* \mapsto \argmax_q L(q,p^*,h^*)$ is continuous.  Note that $q\mapsto L(q,p^*,h^*)$
is \emph{uniformly} strictly concave for all $p^*,h^*$.  Indeed, the hessian is given by
%
\begin{align*}
\frac{\partial L_N(q,p^*,h^*)}{\partial q(x_1,y_1) \partial q(x_2,y_2)} =& 
    - \delta_{y_1=y_2}\sum_\ell N_{Y,\ell} h^*(y_1|\ell) 
              \frac{p^*(x_1|\ell)p^*(x_2|\ell)}{\left(\sum_x p(x|\ell) q(y_1|x)\right)^2} \\
    & \qquad - 2\kappa \delta_{(x_1,y_1)=(x_2,y_2)} \\
\end{align*}
%
Since $q$ lives in a compact space and $L$ is continuous with respect to $p^*,h^*$, we thus 
know that $p^*,h^* \mapsto \argmax_q L_N(q,p^*,h^*)$ is also continuous.  
% STEP II
\item The map $p,\tilde q \mapsto \inf_{q\in\Theta(p,\tilde q)} q(y|x)$ is continuous, as long as 
$\{(p(x_1|\ell)\cdots p(x_n|\ell))\}_\ell$ are linearly independent (remember we have assumed the independence). 
It would suffice to show the following: 
for any fixed $p_1 \approx p_2,\tilde q_1 \approx \tilde q_2,x,y$ where the $\approx$ is sufficiently tight
and $q_1\in \Theta(p_1,\tilde q_1)$, 
we can find $q_2\in \Theta(p_2,\tilde q_2)$ with $q_1(y|x)\approx q_2(y|x)$.  We will solve
this problem first by looking at what happens if $\tilde q$ changes, and then at what happens if $p$ changes.
\begin{itemize}
\item Let us say $p_1=p_2=p$.  Then, in the absence of constraints, we could solve
our problem by simply taking $q_2 = q_1 + \tilde q_2 - \tilde q_1$.
However, if $q_1(y|x)=0$ for any $x,y$, we cannot guarantee that the resulting $q_2$ will be a positive measure. 
To resolve this, we note that the entropic penalty ensures that $\tilde q_2,\tilde q_1$ lie on the interior
of the probability simplex.  This enables us to resolve the issue by constructing ``smoothed'' version of $q_1$, namely
$(1-\alpha) q_1 + \alpha (\tilde q_1)$ which also lies on the interior of the probability simplex.   Let's do this
rigorously.

Fix any $\tilde q_1,\delta>0$ and $q_1 \in \Theta(p,\tilde q_1)$. Let $c = \min_{x,y} \tilde q_1(y|x)$.  Now pick any $\tilde q_2$ such that $|\tilde q_2-\tilde q_1|< \delta c/2$ in the uniform norm.  Take $q_2 = (1-\delta)q_1 + \delta \tilde q_1 + \tilde q_2 - \tilde q_1$.  Observe that $q_2 (y|x) \geq \delta c - \delta c/2$ for every $x,y$, i.e. it is a valid
probability density.  Note also that $q_2 \in \Theta(p, \tilde q_2)$.  Finally, note that $|q_2 - q_1|<\delta$ in
the uniform norm, as desired.

\item Let us now say $\tilde q_1 = \tilde q_2 = \tilde q$.  We are given $q_1,p_1,p_2$ and asked to
produce $q_2\in \Theta(p_2,\tilde q)$ such that $q_1\approx q_2$.  Let 
$A_{1,\ell,x} = p_1^*(x|\ell)$ and $A_{2,\ell,x} = p_2^*(x|\ell)$.  We have assumed that the rows of $A_1$ are linearly independent, so $A_1 A_1^T$ is invertible.  There is therefore an open set around $A_1$ so that as long as $A_2 \approx A_1$ we have that the rows of $A_2$ are also linearly independent.  Therefore to find $q_2 \in \Theta(p_2, \tilde q)$
we can simply take $q_2 = q_1 - A_2^T (A_2 A_2^T)^{-1} A_2 (q_1 -\tilde q)$.  Note that $A_1 (q_1 - \tilde q) =0$, so
the norm of $A_2 (q_1 -\tilde q)$ can be bounded.  Likewise the strength of $(A_2 A_2^T)^{-1}$
can be bounded by the smallest absolute eigenvalue of $A_2$, which, in turn, is bounded by its proximity to $A_1$.  
Finally, the entries of $A_2^T$ are uniformly bounded between zero and one.  In conclusion, 
for any $p_1$ we can always find a $p_2$ close enough so that we can completely control $|q_1-q_2|$ in the Euclidean
norm.  Since we are working in finite dimensions, this is sufficient to control it in the uniform norm.

Finally, we note that it is possible this method will result in $q_2$ being an invalid probability distribution, i.e.
$q_2(y|x)<0$.  This could be true in two senses.  First, it might be that $\sum_y q_2(y|x)\neq 1$.  Fortunately,
this is impossible.  Indeed, $(q_1 - \tilde q)1 =0$ so certainly $(q_2-q_1)1=(\cdots)(q_1-\tilde q)1=0$ as well.   
Second, we might have that $q_2(y|x)<0$.  To remedy this
second point, we take recourse to the same method explained in detail for the case $p_1=p_2$ above. 
That is, we work with a slightly ``smoothed'' version of $q_1$, and find something which is very close to the smoothed version,
and thereby ensure that the resulting $q_2$ is valid.  This $q_2$ is close to the slightly smoothed $q_1$ and thus also
close to $q_1$.

\end{itemize}

\end{enumerate}

It is well-known that the empirical distributions $\hat p, \hat h$ converge uniformly in probability to the true values, $p^*,h^*$.  Since the map $p,q \rightarrow u_1$ is continuous in a neighborhood of $p^*,q^*$, it follows that $\hat u_1$ also converge uniformly to $\tilde u_1^*$ in probability.

It thus suffices to show that $\tilde u_1^* \rightarrow u_1^*$.  To show this, let $\tilde h^* =p^*\tilde q^*$,
i.e. $h^*(y|\ell)=\sum_x p^*(x|\ell)\tilde q^*(y|x)$.  The key will be to control $\tilde h^*$ and see how that control
enables us to control $\tilde u^*$.

\begin{enumerate}
\item $\tilde h^* \rightarrow h^*$ in the uniform norm as 
$N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.

To show this, we express $L$ in terms of 
Kullback-Leibler divergences, $D(h^*_\ell||\tilde h^*_\ell)$ and entropies 
$\mathcal H(h^*_\ell)$.  Applying Pinsker's inequality, we obtain
%
\begin{align*}
L(\tilde q^*,p^*,h^*) &= \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - D(h^*_\ell||\tilde h^*_\ell)) + \kappa \sum_{xy} \log q^*(y|x)\\
&\leq \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - 2|h^*_\ell-\tilde h^*_\ell|^2_{TV} + \kappa \sum_{xy} \log q^*(y|x)\\
&\leq \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - \frac{1}{2}(\sup_y |h^*(y|\ell)-\tilde h^*(y|\ell)|)^2 + \kappa \sum_{xy} \log q^*(y|x)\\
\end{align*}

Since $\tilde q^*,\tilde h^*$ are found by taking $L$ as large as possible, it would thus seem that we could use
this inequality to argue that $h^* \approx \tilde h^*$ when $N_{Y,\ell}$ are large.  There is, however, once again 
a bit of trickiness in the case that $q^*(y|x)=0$ for some $x,y$.  Indeed, it may be that $L(q^*,p^*,h^*)=-\infty$,
and this complicates some of the arguments that we want to make.   To remedy this, we first design a smoothed
version of $q^*$.  Then everything goes through as we might like.  Let's do it rigorously.

Fix $0<\epsilon<1$. We take
$q^*_{SM}= (1-\epsilon) q^* + \epsilon \tilde q^*$ as our smoothed version of $q^*$.  
The convexity of the KL divergence then gives that
%
\begin{align*}
D(h^* || p^*q^*_{SM})&=D(h^*_{\ell} || (1-\epsilon) h^*_{\ell} + \epsilon \tilde h^*) \leq (1-\epsilon)0 + \epsilon D(h^*_{\ell} || \tilde h^*) 
\end{align*}
%
Thus, applying Pinsker (and the fact that $\tilde q^* = \argmax _q L_N(\tilde q,p^*,h^*)$)
%
\begin{align*}
0 &\leq L_N(\tilde q^*,p^*,h^*) - L_N(q^*_{SM},p^*,h^*)\\
 &= \sum_\ell N_{y,\ell} (\epsilon-1) D(h^*_\ell||\tilde h^*_\ell)+ \kappa \sum_{xy} \log \frac{\tilde q^*(y|x)}{q^*_{SM}(y|x)}\\
&\leq \sum_\ell N_{y,\ell} (\epsilon-1) D(h^*_\ell||\tilde h^*_\ell)- \kappa |\Omega_X||\Omega_Y|\log \epsilon \\
&\leq \sum_\ell N_{y,\ell} (\epsilon-1) 2|h^*_\ell-\tilde h^*_\ell|^2_{TV}- \kappa |\Omega_X||\Omega_Y|\log \epsilon \\
&\leq \frac{\epsilon-1}{2}\sum_\ell N_{y,\ell} \left(\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)|\right)^2-\kappa |\Omega_X||\Omega_Y|\log \epsilon \\
\end{align*}
%
One consequence of that inequality is that 
%
\begin{align*}
\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)| &\leq \sqrt{\frac{-2\kappa |\Omega_X||\Omega_Y|\log \epsilon}{N_{Y,\ell}} } \\
\end{align*}
%
With this inequality in hand, we are now prepard to prove that $\tilde h^* \rightarrow h^*$.  Pick any $0<\delta<1$.  
Choose $N_{\ell,y}>-2\kappa |\Omega_X||\Omega_Y|\log \delta/\delta$.  Take $\epsilon=\delta$.  Then
%
\begin{align*}
\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)| &\leq \sqrt{\delta} \\
\end{align*}

\item We now show that if $\tilde h^*\approx h^*$, then $\tilde u_1^* \approx u_1^*$.  To do so, let
%
\[
\tilde \Theta^* = \left\{q: \sum_x p^*(x|\ell)q(y|x)=\sum_x p^*(x|\ell)\tilde q^*(y|x) = \tilde h^*(y|\ell)\right\} 
\]
%
To complete our proof, it then suffices to show that we can find $\tilde h^*$ close enough to $h^*$ so that
we can find a $q \in \tilde \Theta^*$ with $q \approx q^*$.

\end{enumerate}

\end{proof}

Then certainly $u_1^*(y|x) \leq q(y|x)\leq u_2^*(y|x)$.  What's more, we can

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}

### 2018-03-12 trygve/outline.tex
trygve

--- 2018-03-12 outline.tex
@@
 \newtheorem{thm}{Theorem}
+\newtheorem{conj}{Conjecture}
 
@@
 
-When many different measurement tools are available, it can be difficult to understand
-how to combine information from different tools.  This challenge is compounded when the measurement
-tools are destructive.  For example, after processing a neuron with a given single-cell rna
-sequencing technique, we cannot generally go back and process that same neuron with a different
-technique.  This is because the measurement technique destroys the original neuron.  This means
-we can never see \emph{both measurements} for the same cell, side by side.  How, then, can we ever
-hope to understand how the different measurements relate to each other?
+When many different measurement tools are applied, how can we best combine information from the different different tools?
+If we can apply various measurement tools to the same specimens, we can at least begin to understand how the tools they are related.
+We call this ``joint measurement.''  However, the challenge is compounded when joint measurement is unavailable.
+For example, consider the case that the measurement tools are destructive.
+After processing a neuron with a given single-cell rna sequencing technique, we cannot generally go back and process that same neuron with a different technique.
+This is because the measurement technique destroys the original neuron.
+This means we can never see \emph{both measurements} for the same cell, side by side.
+How, then, can we ever hope to understand how the different measurements relate to each other?
 
-In this paper, we hope to shed some light on this this basic mathematical question.  It has many
-applications in the hard sciences.  For example:
+In this paper, we hope to shed some light on this this basic mathematical question.  It has many applications in the sciences.  For example:
 
@@
 
-How could we learn the relationship between two measurement techniques, unless we can observe both measurements 
-applied to the same members of a population (dual measurement)?  In this work we show that it is indeed 
-sometimes possible to get useful and statistically rigorous bounds, even without these dual measurements.
+In this work we show that it is indeed sometimes possible to get useful and statistically rigorous bounds for these kinds of questions, even when joint measurements are unavailable.
 
 In particular, we consider the case that we can obtain samples from a variety of subpopulations.
-If we can assume that the \emph{relationship} between the measurements is the same in each subpopulation, 
-we can use these samples to learn something about that relationship -- even though we never see both 
-measurements for any single member of the population.
+If we can assume that the \emph{relationship} between the measurements is the same in each subpopulation, we can use these samples to learn something about that relationship -- even though we never see both measurements for any single member of the population.
 
-We then apply our method to single cell RNA data.  We show how different clusterings from different single 
-cell RNA modalities can be connected.  Our method yields results which are generally consistent with the scientist's
-understanding of the data and clusters, but also reveal potential gaps which could be important for further study.
+We apply our method to single cell RNA data.
+We show how different clusterings from different single cell RNA modalities can be connected.
+Our method yields results which are generally consistent with the scientist's understanding of the data and clusters, but also reveal potential gaps which could be important for further study.
 
@@
 \begin{itemize}
-\item Let us say we have $n+m$ individual members of a population.  These could be cells, humans, or whatever
-the smallest sample unit may be for a given problem.
-\item For each individual member, we have some readily observable features by which we can group the members,
-such as the size of a cell, where an individual lives, or some other basic information.  We designate these
-features as $\ell_i$ for the $i$th individual.
-\item For the first $n$ members, we have made an observation using the first technique.  This gives us
-the values $X_1 \cdots X_n$.
-\item The last $m$ members were observed using the second technique.  This gives us the values
-$Y_{n+1} \cdots Y_{n+m}$.
+\item Let us say we have $n+m$ individual members of a population.  These could be cells, humans, or whatever the smallest sample unit may be for a given problem.
+\item For each individual member, we have some readily observable features by which we can group the members, such as the size of a cell, where an individual lives, or some other basic information.
+We designate these features as $\ell_i$ for the $i$th individual.
+\item For the first $n$ members, we have made an observation using the first technique.  This gives us the values $X_1 \cdots X_n$.
+\item The last $m$ members were observed using the second technique.  This gives us the values $Y_{n+1} \cdots Y_{n+m}$.
 \end{itemize}
 
-For simplicity, we here assume that $\ell_i,X_i,Y_i$ take values in finite sets, $\Omega_\ell,\Omega_X,\Omega_Y$.  Extensions to more general
-cases should be straightforward, but we must leave them to future work.
+For simplicity, we here assume that $\ell_i,X_i,Y_i$ take values in finite sets, $\Omega_\ell,\Omega_X,\Omega_Y$.
+Extensions to more general cases should be straightforward, but we must leave them to future work.
 
-We further posit a set of \emph{counterfactual} variables -- variables which could not ever be obtained in practice,
-but which help us organize our thinking.  These are sometimes referred to as "potential outcomes"  (cf. \cite{rubin2005causal}).
+We further posit a set of \emph{counterfactual} variables -- variables which could not ever be obtained in practice, but which help us organize our thinking.
+These are sometimes referred to as "potential outcomes"  (cf.
+\cite{rubin2005causal}).
 
 \begin{itemize}
-\item Let $Y_1 \cdots Y_n$ denote the observations we \emph{would have gotten} if we had applied the second technique
-to the first $n$ members.
-\item Let $X_{n+1} \cdots X_n$ denote the observations we would have gotten if we had applied the first technique
-to the last $m$ members. 
+\item Let $Y_1 \cdots Y_n$ denote the observations we \emph{would have gotten} if we had applied the second technique to the first $n$ members.
+\item Let $X_{n+1} \cdots X_n$ denote the observations we would have gotten if we had applied the first technique to the last $m$ members. 
 \end{itemize}
@@
 
-for some distributions $p^*,q^*$.  Moreover, we assume that the vectors $\{(p^*(x_1|\ell)\cdots p^*(x_n|\ell))\}_\ell$ 
-are linearly independent.  If this seems unlikely, it may make sense to collapse similar values of $\ell$ together.
+for some distributions $p^*,q^*$.
+Moreover, we assume that the vectors $\{(p^*(x_1|\ell)\cdots p^*(x_n|\ell))\}_\ell$ are linearly independent.
+If this seems unlikely, it may make sense to collapse similar values of $\ell$ together.
 
 The validity of these assumptions for a given situation should be closely contemplated.
-There are several key questions to answer
-when deciding whether this assumption is applicable.  Does the process by which samples are gathered
-gathered depend only upon $\ell$?  In particular, is it not at all statistically related to which measurement technique was 
-applied, except through $\ell$?  Are the particular measurement biases of both techniques the same for every value of $\ell$?  
-Is the statistical relationship between the quantities being measured by the 
-two techniques the same for every value of $\ell$?\footnote{Note that this is 
-automatically true if both techniques are measuring the same thing.  More generally, if two techniques have something that
-they both measure, we can restrict attention to that one common phenomenon.}  If the answer to all of these questions is yes, the assumption that $(X_i,Y_i)$ is drawn from $p(x_i|\ell_i)p(y_i|x_i)$ may apply.
+There are several key questions to answer when deciding whether this assumption is applicable.
+Does the process by which samples are gathered gathered depend only upon $\ell$?
+In particular, is it not at all statistically related to which measurement technique was  applied, except through $\ell$?
+Are the particular measurement biases of both techniques the same for every value of $\ell$?
+Is the statistical relationship between the quantities being measured by the two techniques the same for every value of $\ell$?\footnote{Note that this is automatically true if both techniques are measuring the same thing.
+More generally, if two techniques have something that they both measure, we can restrict attention to that one common phenomenon.}  If the answer to all of these questions is yes, the assumption that $(X_i,Y_i)$ is drawn from $p(x_i|\ell_i)p(y_i|x_i)$ may apply.
 
-We can now articulate our goal.  Informally, our goal is to see what we can understand about the relationship between
-the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$.  Ideally, we would
-like to learn the distribution $p^*(y_i|x_i)$.  Unfortunately, given the limitations of the data, this may not be 
-possible.  Instead, we articulate a slightly less amibtious goal: to put \emph{bounds} on $q(y_i|x_i)$.  Specifically,
-we will seek to produce functions $\hat u_1(y|x),\hat u_2(y|x)$ such that
+We can now articulate our goal.
+Informally, our goal is to see what we can understand about the relationship between the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$.
+Ideally, we would like to learn the distribution $p^*(y_i|x_i)$.
+Unfortunately, given the limitations of the data, this may not be possible.
+Instead, we articulate a slightly less amibtious goal: to put \emph{bounds} on $q(y_i|x_i)$.
+Specifically, we will seek to produce $\hat u_1(y|x),\hat u_2(y|x)$ such that
 
@@
 
-In particular, we define
+with high probability.  In particular, we define
 
@@
 
-Then obviously $u_1^*(y|x) \leq q^*(y|x)\leq u_2^*(y|x)$ is a valid bound.  Moreover, given the data and assumptions
-we have, it is as tight a bound as is possible to make (indeed, we could not hope to
-distinguish between any two $q_1,q_2 \in \Theta^*$, since they yield the same distributions on what we can
-observe).
+Then obviously $u_1^*(y|x) \leq q^*(y|x)\leq u_2^*(y|x)$ is a valid bound.
+Moreover, given the data and assumptions we have, it is as tight a bound as is possible to make (indeed, we could not hope to distinguish between any two $q_1,q_2 \in \Theta^*$, since they yield the same distributions on what we can observe).
 
-Thus, our mathematical task to to attempt to find estimators for $u_1^*,u_2^*$, which will allow us to bound
-the values of $q^*(y|x)$, and thus understand something about the relationship between the two measurements, $X$ and
-$Y$.
+Thus, our mathematical task to to attempt to find estimators for $u_1^*,u_2^*$, which will allow us to bound the values of $q^*(y|x)$, and thus understand something about the relationship between the two measurements, $X$ and $Y$.
 
@@
 
-These empirical distributions may not be quite consistent with the original assumption that the distribution
-of $X,Y|\ell$ may be written as $p^*(x|\ell)q^*(y|x)$.  In particular, it may be that there is \emph{no} value of $q$ such
-that $\sum_x \hat p(x|\ell) q(y|x) = \hat h(y|\ell)$.  There may also be many such values.  We satisfy ourselves
-with finding a somewhat good fit for $q$.  Let 
-$N_{X,\ell}=\#\{i\leq n:\ \ell_i=\ell\},N_{Y,\ell}=\#\{i> n:\ \ell_i=\ell\}$.  We take
+These empirical distributions may not be quite consistent with the original assumption that the distribution of $X,Y|\ell$ may be written as $p^*(x|\ell)q^*(y|x)$.
+In particular, it may be that there is \emph{no} value of $q$ such that $\sum_x \hat p(x|\ell) q(y|x) = \hat h(y|\ell)$.
+There may also be many such values.
+We satisfy ourselves with finding a somewhat good fit for $q$.
+Let $N_{X,\ell}=\#\{i\leq n:\ \ell_i=\ell\},N_{Y,\ell}=\#\{i> n:\ \ell_i=\ell\}$.
+We take
 
@@
 
-Note in the second term we apply an entropic penalty to $q$.  This is both necessary to ensure 
-that $\hat q$ is uniquely defined in terms of $\hat q,\hat h$, as well as a useful regularization.
-This regularization pushes us slightly towards uniform distributions, which makes sense if we believe that $q(y|x)>0$ 
-for each $x,y$.  Other regularizations could also be used and would potentially
-enjoy similar theoretical guarantees to the ones we will show below.
+Note in the second term we apply an entropic penalty to $q$.
+This is both necessary to ensure that $\hat q$ is uniquely defined in terms of $\hat q,\hat h$, as well as a useful regularization.
+This regularization pushes us slightly towards uniform distributions, which makes sense if we believe that $q(y|x)>0$  for each $x,y$.
+Other regularizations could also be used and would potentially enjoy similar theoretical guarantees to the ones we will show below.
 
-This method gives us an estimate for $\hat q$, but we emphasize this estimate may be \emph{inconsistent}
-for the true value of $q^*$.  As we have said, this is simply a limitation of the data we have available -- attempting to deduce
-$q^*$ may be too much to ask for.  However, we can get bounds.
+This method gives us an estimate for $\hat q$, but we emphasize this estimate may be \emph{inconsistent} for the true value of $q^*$.
+As we have said, this is simply a limitation of the data we have available -- attempting to deduce $q^*$ may be too much to ask for.
+However, we can get bounds.
 
-The key is to search over all possible values of $q$ which agree with $\hat q$ in terms of what \emph{is} identifiable.
-That is, let
+The key is to search over all possible values of $q$ which agree with $\hat q$ in terms of what \emph{is} identifiable. That is, let
 
@@
 
-These will serve as consistent estimators for our quantities of interest, $u_1^*,u_2^*$, as the following theorem shows.\vspace{.1in}
+We conjecture that these will serve as consistent estimators for our quantities of interest, $u_1^*,u_2^*$.\vspace{.1in}
 
-\begin{thm}
-Fix any $\kappa>0$.  Let $N_{X,\ell},N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.  Then $\hat u_1(y|x)\rightarrow u^*_1(y|x)$ in probability, and likewise for $u_2$.
-\end{thm}
+\vspace{.1in}
+\begin{conj}
+Fix any $\kappa>0$.  Let $N_{X,\ell},N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.  Then $\hat u_1(y|x)\rightarrow u^*_1(y|x)$ in probability, and likewise for $\hat u_2,u^*_2$.
+\end{conj}
 
-\begin{proof}
+\section{Empirical results}
 
-The logic is in perfect symmetry for $u_1,u_2$, so we focus on the $u_1$ case.
+...look reasonable, and more or less checked out with the Allen Institute people.
+That said, they caution that some cell types tend to die more often in one experimental modality than another, which would violate our assumptions.
+However, this doesn't bother them, because they see this technique as a way to \emph{validate} their assignments of clusters.
+As such, if the result of our algorithm doesn't make sense, that's great because it may help them to figure out \emph{which cell types are dying more} and then they can upweight certain groups correspondingly.
+Or it may help them figure out that their cluster assignment was all wrong.
+This is how they see my method fitting into their pipelines: as a way of sanity checking their cluster assignments.
+Which seems cool.
 
-We begin by seeing how our process would have worked if we were able to know the exact values of $p^*$ and $h^*$,
-instead of the empirical distributions $\hat p,\hat h$.  That is, we define
-%
+\section{The conjecture}
+
+Here we give an outline for how the conjecture might be proved.
+
+For any transition matrices $p,h$,\footnote{Here by transition matrix we just mean that the rows sum to 1 and the entries are nonnegative.} let $\Theta(p,h)$ denote the subset of transition matrices satisfying
+
+\[
+\Theta(p,h) \triangleq \{q:\ pq=h\}
+\]
+
+Now let us say we have some $p^*,h^*$, and some estimators $\hat p,\hat h$.
+If the estimators were consistent, we might hope that $\Theta(\hat p,\hat h)$ would yield a consistent estimator for $\Theta(p^*,h^*)$.
+In particular, we would ask that the Hausdorff distance would converge to zero in probability, $d_H(\Theta(\hat p,\hat h),\Theta(p^*,h^*))\rightarrow 0$.
+Unfortunately, this is simply not the case.
+Here's an example:
+
 \begin{align*}
-L_N(q,p,h) &\triangleq \sum_\ell N_{Y,\ell}\sum_{y}h(y|\ell)\log\left(\sum_{x}p(x|\ell)q(y|x)\right) + \kappa \sum_{xy} \log q(y|x) \\
-\Theta(p,\tilde q) &\triangleq\left\{q:\ \sum_x p(x|\ell) \tilde q(y|x) =\sum_x p(x|\ell) q(y|x), \forall \ell,y\right\}
-\end{align*}
-%
-and take
-%
-\begin{align*}
-\tilde q^* &= \argmax_q L_N(q,p^*,h^*)  \\
-\tilde u_1^*(y|x) &= \inf_{q\in \Theta(p^*,\tilde q^*)} q(y|x) \\
-\end{align*}
-%
-Our first task is to show that the map $p^*,h^* \mapsto \tilde u_1^*$ is continuous.  This follows in two steps:
-
-%%%%% two steps
-\begin{enumerate}
-% STEP I
-\item The map $p^*,h^* \mapsto \argmax_q L(q,p^*,h^*)$ is continuous.  Note that $q\mapsto L(q,p^*,h^*)$
-is \emph{uniformly} strictly concave for all $p^*,h^*$.  Indeed, the hessian is given by
-%
-\begin{align*}
-\frac{\partial L_N(q,p^*,h^*)}{\partial q(x_1,y_1) \partial q(x_2,y_2)} =& 
-    - \delta_{y_1=y_2}\sum_\ell N_{Y,\ell} h^*(y_1|\ell) 
-              \frac{p^*(x_1|\ell)p^*(x_2|\ell)}{\left(\sum_x p(x|\ell) q(y_1|x)\right)^2} \\
-    & \qquad - 2\kappa \delta_{(x_1,y_1)=(x_2,y_2)} \\
-\end{align*}
-%
-Since $q$ lives in a compact space and $L$ is continuous with respect to $p^*,h^*$, we thus 
-know that $p^*,h^* \mapsto \argmax_q L_N(q,p^*,h^*)$ is also continuous.  
-% STEP II
-\item The map $p,\tilde q \mapsto \inf_{q\in\Theta(p,\tilde q)} q(y|x)$ is continuous, as long as 
-$\{(p(x_1|\ell)\cdots p(x_n|\ell))\}_\ell$ are linearly independent (remember we have assumed the independence). 
-It would suffice to show the following: 
-for any fixed $p_1 \approx p_2,\tilde q_1 \approx \tilde q_2,x,y$ where the $\approx$ is sufficiently tight
-and $q_1\in \Theta(p_1,\tilde q_1)$, 
-we can find $q_2\in \Theta(p_2,\tilde q_2)$ with $q_1(y|x)\approx q_2(y|x)$.  We will solve
-this problem first by looking at what happens if $\tilde q$ changes, and then at what happens if $p$ changes.
-\begin{itemize}
-\item Let us say $p_1=p_2=p$.  Then, in the absence of constraints, we could solve
-our problem by simply taking $q_2 = q_1 + \tilde q_2 - \tilde q_1$.
-However, if $q_1(y|x)=0$ for any $x,y$, we cannot guarantee that the resulting $q_2$ will be a positive measure. 
-To resolve this, we note that the entropic penalty ensures that $\tilde q_2,\tilde q_1$ lie on the interior
-of the probability simplex.  This enables us to resolve the issue by constructing ``smoothed'' version of $q_1$, namely
-$(1-\alpha) q_1 + \alpha (\tilde q_1)$ which also lies on the interior of the probability simplex.   Let's do this
-rigorously.
-
-Fix any $\tilde q_1,\delta>0$ and $q_1 \in \Theta(p,\tilde q_1)$.
-Let $c = \min_{x,y} \tilde q_1(y|x)$.
-Now pick any $\tilde q_2$ such that $|\tilde q_2-\tilde q_1|< \delta c/2$ in the uniform norm.
-Take $q_2 = (1-\delta)q_1 + \delta \tilde q_1 + \tilde q_2 - \tilde q_1$.
-Observe that $q_2 (y|x) \geq \delta c - \delta c/2$ for every $x,y$, i.e.
-it is a valid
-probability density.  Note also that $q_2 \in \Theta(p, \tilde q_2)$.  Finally, note that $|q_2 - q_1|<\delta$ in
-the uniform norm, as desired.
-
-\item Let us now say $\tilde q_1 = \tilde q_2 = \tilde q$.  We are given $q_1,p_1,p_2$ and asked to
-produce $q_2\in \Theta(p_2,\tilde q)$ such that $q_1\approx q_2$.  Let 
-$A_{1,\ell,x} = p_1^*(x|\ell)$ and $A_{2,\ell,x} = p_2^*(x|\ell)$.
-We have assumed that the rows of $A_1$ are linearly independent, so $A_1 A_1^T$ is invertible.
-There is therefore an open set around $A_1$ so that as long as $A_2 \approx A_1$ we have that the rows of $A_2$ are also linearly independent.
-Therefore to find $q_2 \in \Theta(p_2, \tilde q)$
-we can simply take $q_2 = q_1 - A_2^T (A_2 A_2^T)^{-1} A_2 (q_1 -\tilde q)$.  Note that $A_1 (q_1 - \tilde q) =0$, so
-the norm of $A_2 (q_1 -\tilde q)$ can be bounded.  Likewise the strength of $(A_2 A_2^T)^{-1}$
-can be bounded by the smallest absolute eigenvalue of $A_2$, which, in turn, is bounded by its proximity to $A_1$.  
-Finally, the entries of $A_2^T$ are uniformly bounded between zero and one.  In conclusion, 
-for any $p_1$ we can always find a $p_2$ close enough so that we can completely control $|q_1-q_2|$ in the Euclidean
-norm.  Since we are working in finite dimensions, this is sufficient to control it in the uniform norm.
-
-Finally, we note that it is possible this method will result in $q_2$ being an invalid probability distribution, i.e.
-$q_2(y|x)<0$.  This could be true in two senses.  First, it might be that $\sum_y q_2(y|x)\neq 1$.  Fortunately,
-this is impossible.  Indeed, $(q_1 - \tilde q)1 =0$ so certainly $(q_2-q_1)1=(\cdots)(q_1-\tilde q)1=0$ as well.   
-Second, we might have that $q_2(y|x)<0$.  To remedy this
-second point, we take recourse to the same method explained in detail for the case $p_1=p_2$ above. 
-That is, we work with a slightly ``smoothed'' version of $q_1$, and find something which is very close to the smoothed version,
-and thereby ensure that the resulting $q_2$ is valid.  This $q_2$ is close to the slightly smoothed $q_1$ and thus also
-close to $q_1$.
-
-\end{itemize}
-
-\end{enumerate}
-
-It is well-known that the empirical distributions $\hat p, \hat h$ converge uniformly in probability to the true values, $p^*,h^*$.
-Since the map $p,q \rightarrow u_1$ is continuous in a neighborhood of $p^*,q^*$, it follows that $\hat u_1$ also converge uniformly to $\tilde u_1^*$ in probability.
-
-It thus suffices to show that $\tilde u_1^* \rightarrow u_1^*$.  To show this, let $\tilde h^* =p^*\tilde q^*$,
-i.e. $h^*(y|\ell)=\sum_x p^*(x|\ell)\tilde q^*(y|x)$.  The key will be to control $\tilde h^*$ and see how that control
-enables us to control $\tilde u^*$.
-
-\begin{enumerate}
-\item $\tilde h^* \rightarrow h^*$ in the uniform norm as 
-$N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.
-
-To show this, we express $L$ in terms of 
-Kullback-Leibler divergences, $D(h^*_\ell||\tilde h^*_\ell)$ and entropies 
-$\mathcal H(h^*_\ell)$.  Applying Pinsker's inequality, we obtain
-%
-\begin{align*}
-L(\tilde q^*,p^*,h^*) &= \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - D(h^*_\ell||\tilde h^*_\ell)) + \kappa \sum_{xy} \log q^*(y|x)\\
-&\leq \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - 2|h^*_\ell-\tilde h^*_\ell|^2_{TV} + \kappa \sum_{xy} \log q^*(y|x)\\
-&\leq \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - \frac{1}{2}(\sup_y |h^*(y|\ell)-\tilde h^*(y|\ell)|)^2 + \kappa \sum_{xy} \log q^*(y|x)\\
+p^* &= (1, 0)\\
+h^* &= (1, 0)\\
+\Theta(p^*,h^*) &= 
+  \left\{q:\ q=\left(\begin{array}{cc}
+  1 & 0 \\
+  \alpha & 1-\alpha \\
+  \end{array}\right)\right\}
 \end{align*}
 
-Since $\tilde q^*,\tilde h^*$ are found by taking $L$ as large as possible, it would thus seem that we could use
-this inequality to argue that $h^* \approx \tilde h^*$ when $N_{Y,\ell}$ are large.  There is, however, once again 
-a bit of trickiness in the case that $q^*(y|x)=0$ for some $x,y$.  Indeed, it may be that $L(q^*,p^*,h^*)=-\infty$,
-and this complicates some of the arguments that we want to make.   To remedy this, we first design a smoothed
-version of $q^*$.  Then everything goes through as we might like.  Let's do it rigorously.
+Notice that the second row of $q$ is left \emph{totally} unspecified.
+This is because the equation $p^*q=h^*$ simply places no constraints on the second row of $q$.
+Now let us say we use our consistent estimator, and get something \emph{very} close to the truth.
+We might get a dreadfully wrong impression of $q$:
 
-Fix $0<\epsilon<1$. We take
-$q^*_{SM}= (1-\epsilon) q^* + \epsilon \tilde q^*$ as our smoothed version of $q^*$.  
-The convexity of the KL divergence then gives that
-%
 \begin{align*}
-D(h^* || p^*q^*_{SM})&=D(h^*_{\ell} || (1-\epsilon) h^*_{\ell} + \epsilon \tilde h^*) \leq (1-\epsilon)0 + \epsilon D(h^*_{\ell} || \tilde h^*) 
-\end{align*}
-%
-Thus, applying Pinsker (and the fact that $\tilde q^* = \argmax _q L_N(\tilde q,p^*,h^*)$)
-%
-\begin{align*}
-0 &\leq L_N(\tilde q^*,p^*,h^*) - L_N(q^*_{SM},p^*,h^*)\\
- &= \sum_\ell N_{y,\ell} (\epsilon-1) D(h^*_\ell||\tilde h^*_\ell)+ \kappa \sum_{xy} \log \frac{\tilde q^*(y|x)}{q^*_{SM}(y|x)}\\
-&\leq \sum_\ell N_{y,\ell} (\epsilon-1) D(h^*_\ell||\tilde h^*_\ell)- \kappa |\Omega_X||\Omega_Y|\log \epsilon \\
-&\leq \sum_\ell N_{y,\ell} (\epsilon-1) 2|h^*_\ell-\tilde h^*_\ell|^2_{TV}- \kappa |\Omega_X||\Omega_Y|\log \epsilon \\
-&\leq \frac{\epsilon-1}{2}\sum_\ell N_{y,\ell} \left(\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)|\right)^2-\kappa |\Omega_X||\Omega_Y|\log \epsilon \\
-\end{align*}
-%
-One consequence of that inequality is that 
-%
-\begin{align*}
-\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)| &\leq \sqrt{\frac{-2\kappa |\Omega_X||\Omega_Y|\log \epsilon}{N_{Y,\ell}} } \\
-\end{align*}
-%
-With this inequality in hand, we are now prepard to prove that $\tilde h^* \rightarrow h^*$.  Pick any $0<\delta<1$.  
-Choose $N_{\ell,y}>-2\kappa |\Omega_X||\Omega_Y|\log \delta/\delta$.  Take $\epsilon=\delta$.  Then
-%
-\begin{align*}
-\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)| &\leq \sqrt{\delta} \\
+\hat p &= \left(\frac{99}{100}, \frac{1}{100}\right)\\
+\hat h &= (1, 0)\\
+\Theta(\hat p,\hat h) &= 
+  \left\{q:\ q=\left(\begin{array}{cc}
+  1 & 0 \\
+  1 & 0 \\
+  \end{array}\right)\right\}
 \end{align*}
 
-\item We now show that if $\tilde h^*\approx h^*$, then $\tilde u_1^* \approx u_1^*$.  To do so, let
-%
-\[
-\tilde \Theta^* = \left\{q: \sum_x p^*(x|\ell)q(y|x)=\sum_x p^*(x|\ell)\tilde q^*(y|x) = \tilde h^*(y|\ell)\right\} 
-\]
-%
-To complete our proof, it then suffices to show that we can find $\tilde h^*$ close enough to $h^*$ so that
-we can find a $q \in \tilde \Theta^*$ with $q \approx q^*$.
+Suddenly $q$ is \emph{entirely constrained} by the equation $pq=h$.
+This dramatic change is possible because $q$ is constrained not only by the equation $pq=h$ but also by the fact that $q$ must be a transition matrix.
+This allows for the possibility that although $p^*\approx \hat p$ and $h^*=\hat h$, we have that $d_H(\Theta(\hat p,\hat h),\Theta(p^*,h^*))=1$.
+Not good.
 
-\end{enumerate}
+However, this troubling example relied crucially on the fact that a given entry of $p$ went from zero to nonzero.
+Let us therefore ask that the estimator $\hat p$ was consistent both in the usual Euclidean $\mathscr L^2$ sense, as well as in the sense that it eventually got the zeros right.
+That is, $\mathbb P(\hat p(x|\ell)=0\Leftrightarrow p^*(x|\ell)=0) \rightarrow 1$.
+Such an estimator is certainly easy to come by (the empirical distribution will do).
+To prove consistency, it would then be sufficient to show that
 
-\end{proof}
+\vspace{.1in}
+\begin{conj}
+For every $\epsilon>0$, we can find a $\delta>0$ so that if $|h_1 -h_2|^2<\delta$, $|p_1 -p_2|^2<\delta$, and $p_1(x|\ell)=0\Leftrightarrow p_2(x|\ell)=0$, then $d_H(\Theta(p_1,h_1),\Theta(p_1,h_1))<\epsilon$.
+\end{conj}
 
-Then certainly $u_1^*(y|x) \leq q(y|x)\leq u_2^*(y|x)$.  What's more, we can
+To prove this conjecture, the essential point is to understand how the constraint $pq=h$ related to the $k$-faces (vertices, facets, etc) of the polytope of transition matrices.
+If that constraint is somehow \emph{degenerate} with respect to one of the faces, then we get in trouble because a slight perturbation can change the degeneracy and thus create weird effects.
+However, as long as the nature of the degeneracy is the same for both $p_1,p_2$, I believe there is no problem.
+So to prove this conjecture, it remains to do three essential things
+
+\begin{itemize}
+    \item mathematically characterize the ``degeneracy'' of the constraint $pq=h$ with respect to faces of the transition matrix polytope
+    \item show that if $p_1,p_2$ share the same qualitative nature of degeneracy and $p_1\approx p_2$, then $\Theta(p_1,h_1)\approx\Theta(p_2,h_2)$.
+    \item show that the qualitative nature of this degeneracy is only changed when an entry of $p$ switches between zero and nonzero
+\end{itemize}
 

### 2018-03-12 generativemodelsforscrna/outline.tex
generatiev models

% !TEX program = pdflatex
\input{preamble/preamble.tex}
\input{preamble/preamble_math.tex}
\input{preamble/preamble_acronyms.tex}

% \linenumbers
\DeclareMathOperator*{\Categorical}{Categorical}

\title{Generative Models for scRNA}
\author{Jackson Loper}

\usepackage{amsthm}
\newtheorem{thm}{Theorem}
\newtheorem{conj}{Conjecture}

\begin{document}
\maketitle

\begin{description}
  \item[Status] Succesfully trained the new ``iterative model,'' looked at it
  \item[Todo] Compare clusters against Allen, compare held-out log likeloohd
\end{description}

\section{Overview}

The Allen Institute has developed a method for clustering, one that has apparently allowed them to distinguish 116 unique cell types using single cell RNA data.  The method follows the following iterative procedure:

\begin{algorithm}[H]
 \KwData{RNA expression for a collection of cells}
 \vspace{.1in}
  Run PCA on the data (e.g. take first 16 components)\;
  Run T-SNE on the resulting low-d representation\;
  Circle things that look like clusters\;
  \If{there is more than one cluster}{\For{each cluster $c$}{run itclust on the cells inside $c$\;}}
 \caption{itclust}
\end{algorithm}

This method is nice because the important low-dimensional representations \emph{within} a cluster may be quite different from the ones that help you distinguish \emph{between} two subclusters within that cluster.  This is what allows them to achieve the level of granularity that they have.

Unfortunately, it is a bit difficult to grasp the uncertainty involved in the process.  There is certainly noise in the data, and different practitioners of this basic procedure may have different understandings of what this noise is.  The result is that when Bosilijka Tasic (one of the lead experimentalists at Allen) gives two data scientists the same data, they each come back with different answers.  And they have no way to compare or evaluate these two different answers.  One essential question is ``what level of granularity does the signal-to-noise ratio support?''  The algorithm above, as stated, runs iteratively until subclusters cannot be distinguished via T-SNE.  But this is obviously a very subjective matter.  What to do?

We propose to help clarify some of these issues by building a fully generative model of gene expression, one that rigorously measures some kinds of uncertainty in every step of the process.  To be most useful to the Allen Institute, however, we also want this generative model to fit into the scientific paradigm under which they are operating.  In particular, the treatment of clusters should be hierarchical in nature, and it is desirable that the model have latent representations that roughly correspond to the low-dimensional representations found in the Allen algorithm.

Towards this end, we propose the following generative model for gene expression of a single cell.  Let $H$ denote a tree of cell types.

\begin{enumerate}
\item $T \sim \Categorical(\pi)$, one of the leaves of $H$.  Let $t_0,t_1(T),t_2(T),\cdots ,t_\ell(T)$, indicate the nodes in the unique path in $H$ which leads from the root to the leaf $T$.  Here $t_0$ is the root of the tree, $\ell$ is the depth of the leaf $T$, and $t_\ell(T)=T$.   To consolidate notation, we take $T_i \triangleq t_i(T)$.
\item For each $i<\ell$ let $Z_{T_i} \sim \cN(\mu_{T_{i+1}},\Sigma_{T_{i+1}})$, and take $Z_{T_\ell} \sim \cN(0,I)$
\item For each gene $g$, let $\lambda_g \in \mathbb R^{p}$ denote a set of parameters.  These will be a deterministic function of the $Z$s:
\[
\lambda_g = f_g \left(\sum_{i=0}^{\ell} h_{T_i}Z_{T_i}\right)
\]
\item Conditioned on $Z$, the  $g$th gene of the cell be distributed according to some $p(x_g ; \lambda_g)$.
\end{enumerate}

Let us unpack some of this:

\begin{itemize}
\item The latent representation $Z_i$ corresponds to features of that cell which are specific to all the cells which are descendents of $T_i$.
\item The the cell's subcluster $T_{i+1}$ effects the distribution of its latent features at the $i$th level, $Z_i$.  That is, the latent distribution on the $i$th latent representation is governed by what subcluster it is part of within $T_i$.
\item The distribution of the final latent representation, $Z_T$, is simply a standard normal.
\item The functions $h_{T_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
\end{itemize}

\section{Empirical results}

At the current moment, we have only actually implemented this model in the case the $H$ is a tree with depth 1, i.e. the hierarchy is trivial.  We use amortized variational methods for training and inference.  It looks fairly promising.  For a simple qualitative assessment, consider the following.

Scatterplot of the aggregate posterior distribution of $Z_{T_0}$, colored by posterior mean of the cluster assigment:

\includegraphics[width=0.4\textwidth]{pics/Z1}

Notice that there are only so many clusters that are really clearly distinct in this latent space.  However, let's focus on that cluster on the right-hand side.  What happens if we look at the $Z_{T_1}$ variables for those cells?  We get substantial subclustering apparent in the $Z_{T_2}$ space:

\includegraphics[width=0.4\textwidth]{pics/Z2}

% \bibliographystyle{unsrt}
% \bibliography{refs}

\end{document}

### 2018-03-15 generativemodelsforscrna/outline.tex
put in some numbers, commit some code

@@
 \begin{description}
-  \item[Status] Succesfully trained the new ``iterative model,'' looked at it
-  \item[Todo] Compare clusters against Allen, compare held-out log likeloohd
+  \item[Status] Succesfully trained the new ``iterative model,'' evaluated it to some extent
+  \item[Todo] Compare clusters against Allen, try unamortized inference
 \end{description}
@@
 
+\section{Model}
+
 Towards this end, we propose the following generative model for gene expression of a single cell.  Let $H$ denote a tree of cell types.
@@
 
+For inference, we use amortized variational inference.
+In particular, we take the variational family that $T$ and all the $Z_{T_i}$ are independent, $T$ is some categorical, and $Z$s are isotropic gaussians.
+We build a neural network to estimate the variational parameters from gene expressions.
+Note that this network must estimate the parameters for a gaussian at \emph{each node} in the tree $H$.
+
 \section{Empirical results}
@@
 the hierarchy is trivial.
-We use amortized variational methods for training and inference.
+We take the data model to be a zero-inflated negative binomial.
 It looks fairly promising.
@@
 
+\subsection{Held out log likelihood}
+
+To quantitatively analyse our method, we used a held out log likelihood:
+
+\begin{itemize}
+\item Train the method on 75\% of the cells
+\item For each of the remaining 25\% of the cells, select 75\% of the genes (``observed''):
+  \begin{itemize}
+  \item Compute the posterior mode of Z given the expression of those genes
+  \item Compute the log probability of the observed genes given the posterior mode
+  \item Compute the log probability of the unobserved genes given the posterior mode
+  \end{itemize}
+\item Average values over those 25\% of the cells
+\end{itemize}
+
+Here are the results for this ``Posterior mode predictive (PMP)'' approach:
+
+\begin{tabular}{r|cc}
+Method & PMP, observed & PMP, unobserved \\
+Factor analysis $k=16$ & -4.2 & -4.2 \\
+Factor analysis $k=64$ & -4.1 & -4.3 \\
+HNBPF $k=2$ & -5.7 & -5.7 \\
+HNBPF $k=16$ & -3.8 & -3.9 \\
+HNBPF $k=64$ & -3.8 & -5.7 \\
+ZINB-Vae $k=2$ & -3.6 & -3.6 \\
+Smooth ZINB-Vae $k=2$ & -3.6 & -3.6 \\
+Smooth Hierarchical ZINB $k=2+2$ & -3.5 & -3.5 \\
+\hline
+\end{tabular}
+
+Here HNBPF is a negative-binomial-poisson-factorization approach, ZINB-Vae is based on a Michael Jordan paper, $k$ is the number of latent dimensions.
+The results are certainly not mind-blowing, but its nice to see that we're at least heading in a reasonable direction.
+There may be an upper limit on the PMP, due to very real technical noise in the system, so it is hard to evaluate whether this is a trivial improvement or not.
+
+To arrive at our figures for our model, we use a truly ridiculous method of variational inference.
+In particular, we simply \emph{zero out} all of the ``unobserved'' entries, and apply our inference network unchanged to get estimates for the variational parameters.
+This is not ``cheating'' in the sense that it is a practical method for imputation in the case where certain genes are unobserved.
+However, it might well do less than optimally.
+I was talking to Scott, and it sounds like the question of amortized inference with unpredictably missing data is a big deal right now.
+Worth thinking about more.
+
+\subsection{Denoising}
+
 % \bibliographystyle{unsrt}

### 2018-03-15 tryg/batteries.py

@@ -1,0 +1,682 @@
+import tensorflow as tf
+
+
+get_variable2=tf.get_variable
+
+
+def range_noisy(*args):
+    for i in iterate_noisy(range(*args)):
+        yield i
+
+
+import time
+import numpy as np
+import sys
+
+
+import numpy as np
+import scipy as sp
+import scipy.stats
+import scipy.special
+
+def iterate_noisy(lst):
+
+    times=[]
+
+    lst=list(lst)
+    n=len(lst)
+
+    assert n%1==0    
+    n=int(n)
+    assert n>0
+    regularity=n//20
+    if regularity<=0:
+        regularity=1
+    
+    for i in range(n):
+        yield lst[i]
+        times.append(time.time())
+        
+        if i<5 or i%regularity==0:
+            if len(times)>1:
+                elapsed=times[-1]-times[0]
+                meantime=np.mean(np.diff(times))
+
+                completion_expected=(elapsed/i)*n
+                time_remaining=(elapsed/i)*(n-i)
+
+                # pnn("(%d⸗%.1f)"%(i,time_remaining),space=False)
+                pnn("(%d⸗%.1f)"%(i,time_remaining/60.0),space=False)
+                # pnn("(%d %.2f %.2f)"%(i,elapsed/60.0, (n-i)*meantime/60.0))
+                # pnn("(%d:%.1f)"%(i,(n-i)*meantime/60.0))
+
+                # print("%f elapsed, on iteration %d, completion expected %f, time remaining %f"%(
+                    # elapsed,i,completion_expected,time_remaining))
+
+        if i==n-1:
+            pnn("(UTL⸗%1f)"%((times[-1]-times[0])/60.0))
+
+def pnn(x,space=True): #{{{1
+    ''' print no-newline '''
+    sys.stdout.write(str(x) + (" " if space else ''))
+    sys.stdout.flush()
+
+'''
+ _        _                               
+| |_ __ _| |__   ___ ___  _ __ ___  _ __  
+| __/ _` | '_ \ / __/ _ \| '_ ` _ \| '_ \ 
+| || (_| | |_) | (_| (_) | | | | | | |_) |
+ \__\__,_|_.__/ \___\___/|_| |_| |_| .__/ 
+                                   |_|    
+'''
+
+class TabCompleteGraph:
+    def __init__(self,gr,scope=''):
+        self.gr=gr
+        self.scope='/'+scope
+        self.lenscope=len(self.scope)
+
+    def __repr__(self):
[…]

### 2018-03-20 generativemodelsforscrna/outline.tex
updates for mtg

@@
   \item[Status] Succesfully trained the new ``iterative model,'' evaluated it to some extent
-  \item[Todo] Compare clusters against Allen, try unamortized inference
+  \item[Todo] Compare clusters against Allen, consider linear model, do more layers
 \end{description}
@@
 Here HNBPF is a negative-binomial-poisson-factorization approach, ZINB-Vae is based on a Michael Jordan paper, $k$ is the number of latent dimensions.
-The results are certainly not mind-blowing, but its nice to see that we're at least heading in a reasonable direction.
-There may be an upper limit on the PMP, due to very real technical noise in the system, so it is hard to evaluate whether this is a trivial improvement or not.
+The results are certainly not mind-blowing, but its nice to see that we're at least not doing worse.
 
-To arrive at our figures for our model, we use a truly ridiculous method of variational inference.
-In particular, we simply \emph{zero out} all of the ``unobserved'' entries, and apply our inference network unchanged to get estimates for the variational parameters.
-This is not ``cheating'' in the sense that it is a practical method for imputation in the case where certain genes are unobserved.
-However, it might well do less than optimally.
-I was talking to Scott, and it sounds like the question of amortized inference with unpredictably missing data is a big deal right now.
-Worth thinking about more.
+To do posterior inference on held out likelihood, our approach is to first set all of the ``unobserved'' entries to the average value for the corresponding genes, and apply our inference network unchanged to get estimates for the variational parameters.
+We then refine these estimates with gradient descent.
+We note that if we use completely random guesses to initialize the gradient descent, we often converge to inferior local minima; this shows the benefit of the amortized inference.
 
-\subsection{Denoising}
+\subsection{Posterior predictive samples}
+
+One limitation of the method as we have implemented it is that the posterior predictive distribution seems to be substantially underfitting.  To see what we mean, consider the following figure:
+
+\includegraphics[width=0.8\textwidth]{pics/Z1gn}
+
+Here we four rows and three columns of plots.
+Each row corresponds to a different gene.
+Each dot in each plot corresponds to a cell.
+The position of the dot indicates the latent representation of the cell, as inferred by posterior inference.
+The color indicates the amount of expression for the particular gene for that cell, measured three ways:
+
+\begin{enumerate}
+\item The first column shows the true expression of the appropriate gene for each cell (i.e the color of each dot is the gene expression for that cell)
+\item The second column shows a posterior predictive sample of the gene expression for each cell
+\item The third column shows the mean of such samples for each cell
+\end{enumerate}
+
+We see that the true expression doesn't look much like the posterior predictive expression for the Vip and Sst genes.
+Something better is needed.
+Possibly more layers.
+Possibly more training.
+Worth thinking about.
 

### 2018-03-20 trygve/outline.tex
generatiev models
updates for mtg

@@
 \maketitle
+
+\begin{description}
+  \item[Status] Designed estimator, showed result to Allen people, they seem happy
+  \item[Todo] Prove consistency, try on other Allen datasets, sample uniformly from possible set
+\end{description}
 
@@
 
-...look reasonable, and more or less checked out with the Allen Institute people.
-That said, they caution that some cell types tend to die more often in one experimental modality than another, which would violate our assumptions.
+We looked at Allen Institute cell-type assignment of cells, performed using two different modalities: facs and patchseq.
+Our input was two table: cre lines by facs cluster, and cre lines by patch cluster (Figure \ref{fig:alleninput}).
+Our output was facs cluster by patch cluster (Figure \ref{fig:allenoutput}).
+
+\begin{figure}
+\includegraphics[width=.9\textheight,angle=270]{pics/alleninput}
+\caption{Allen institute data: Our input \label{fig:alleninput}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allen}
+\caption{Allen institute data: Our output \label{fig:allenoutput}}
+\end{figure}
+
+These looked reasonable to the scientists, and also pointed out some possible issues.
+In particular, some cell types tend to die more often in one experimental modality than another, which would violate our assumptions.
 However, this doesn't bother them, because they see this technique as a way to \emph{validate} their assignments of clusters.
@@
 Which seems cool.
+
+There is one possibly confusing aspect of our output that might be nice to address.
+The upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.
+For example, the rows of the upper bounds generally sum to more than 1.
+This is because we have taken \emph{elementwise bounds} on each entry.
+However, it would be nice to sample uniformly from the set of possible values of $q$ which are consistent with our observations.
+MCMC or rejection sampling might be good options to investigate.
+These would help give the user another way to understand what possible values $q$ might take on.
 

### 2018-03-21 generativemodelsforscrna/outline.tex
further training improved ppc plots
added denoising plots to scrna

@@
 
-We see that the true expression doesn't look much like the posterior predictive expression for the Vip and Sst genes.
+We see that the true expression doesn't match the posterior predictive expression for the Vip and Sst genes as well as we might like.
 Something better is needed.
@@
 
+\subsection{Posterior predictive means}
+
+The posterior predictive mean can be interpreted as a ``denoised'' version of the original.  In this section we examine the Allen data from that point view.
+
+For example, here we have selected 1000 random entries from the gene by cell matrix.
+For each entry, we plot a point in which the horizontal location indicates the true value of that entry, and the vertical position indicates the posterior predictive expectation of that entry.
+
+\includegraphics[width=0.8\textwidth]{pics/resids2}
+
+We can compare this with a similar plot using posterior predictive samples instead of the true values.  This then forms a different kind of posterior predictive check.
+
+\includegraphics[width=0.8\textwidth]{pics/resids3}
+
+The fact that these two plots look the same is encouraging.
+
+We can also use heatmaps to try to give more structure to our understanding of this ``denoising'' process.
+Here we have three plots: one giving the original gene expression for the first 1000 cells, one giving the posterior expected gene expression for the same cells, and one giving the residual.
+In all cases we have transformed the data by taking $\log(1+x)$ and subtracting gene-level means.
+Each column corresponds to a gene, and each row corresponds to a cell.
+We have sorted the cells by one of the latent components, and sorted the genes by using a property of the gene decoder.
+
+\includegraphics[width=0.8\textwidth]{pics/resids}
+
+Note that there is a group of cells which are much more difficult to understand, which are found at the bottom of the plot.  These are mostly astrocytes.
+
+Since we have 20000 cells, it might be nice to make a similar plot for all the cells.
+However, such a plot would be very tall and perhaps difficult to look at.
+We can instead take the full 20000 x 1400 plot and average neighboring rows and columns together.
+This gives a plot like this:
+
+\includegraphics[width=0.6\textwidth]{pics/residsfull}
+
+If there was a lot of structure in the residual, we would feel that perhaps there was something important and obvious we were missing.  Since there isn't, we are reasonably encouraged.
+
 % \bibliographystyle{unsrt}

### 2018-03-21 generativemodelsforscrna/outline.tex
clarified aggregate posterior plots marginally

@@
 
-Scatterplot of the aggregate posterior distribution of $Z_{T_0}$, colored by posterior mean of the cluster assigment:
+Scatterplot of the aggregate posterior distribution of $Z_{T_0}$, colored by posterior mean of the cluster assigment.
+That is, each dot is a cell, its position is given by a posterior sample of $Z_{T_0}$, and its color has something to do with what cluster we expect it to be in.
 
@@
 What happens if we look at the $Z_{T_1}$ variables for those cells?
-We get substantial subclustering apparent in the $Z_{T_2}$ space:
+We get substantial subclustering apparent in the $Z_{T_1}$ space.
+In the following picture, each dot is a cell from the cluster on the right-hand-side of the previous figure, and the position is given by a posterior sample of $Z_{T_1}$.
 

### 2018-03-23 generativemodelsforscrna/outline.tex
Added cell indexing

@@
 
-Towards this end, we propose the following generative model for gene expression of a single cell.  Let $H$ denote a tree of cell types.
+Towards this end, we propose the following generative model for gene expression of a single cell.
+We will denote the expression of gene $g$ in cell $c$ by $X_g^c$.
+Let $H$ denote a tree of cell types.
 
 \begin{enumerate}
-\item $T \sim \Categorical(\pi)$, one of the leaves of $H$.
-Let $t_0,t_1(T),t_2(T),\cdots ,t_\ell(T)$, indicate the nodes in the unique path in $H$ which leads from the root to the leaf $T$.
-Here $t_0$ is the root of the tree, $\ell$ is the depth of the leaf $T$, and $t_\ell(T)=T$.
-To consolidate notation, we take $T_i \triangleq t_i(T)$.
-\item For each $i<\ell$ let $Z_{T_i} \sim \cN(\mu_{T_{i+1}},\Sigma_{T_{i+1}})$, and take $Z_{T_\ell} \sim \cN(0,I)$
-\item For each gene $g$, let $\lambda_g \in \mathbb R^{p}$ denote a set of parameters.  These will be a deterministic function of the $Z$s:
+\item $T^c \sim \Categorical(\pi)$, one of the leaves of $H$.
+Let $t_0,t_1(T^c),t_2(T^c),\cdots ,t_\ell^c(T^c)$, indicate the nodes in the unique path in $H$ which leads from the root to the leaf $T$.
+Here $t_0$ is the root of the tree, $\ell^c$ is the depth of the leaf $T^c$, and $t^c_\ell(T^c)=T^c$.
+To consolidate notation, we take $T^c_i \triangleq t_i(T^c)$.
+\item For each $i<\ell$ let $Z^c_{T^c_i} \sim \cN(\mu_{T^c_{i+1}},\Sigma_{T^c_{i+1}})$, and take $Z^c_{T^c_\ell} \sim \cN(0,I)$
+\item For each gene $g$, let $\lambda^c_g \in \mathbb R^{p}$ denote a set of parameters.  These will be a deterministic function of the $Z$s:
 \[
-\lambda_g = f_g \left(\sum_{i=0}^{\ell} h_{T_i}Z_{T_i}\right)
+\lambda^c_g = f_g \left(\sum_{i=0}^{\ell} h_{T^c_i}(Z^c_{T^c_i})\right)
 \]
-\item Conditioned on $Z$, the  $g$th gene of the cell be distributed according to some $p(x_g ; \lambda_g)$. 
+\item Conditioned on $Z^c$ we will have $X^c_g \sim p(x_g ; \lambda^c_g)$. 
 \end{enumerate}
@@
 \begin{itemize}
-\item The latent representation $Z_i$ corresponds to features of that cell which are specific to all the cells which are descendents of $T_i$.
-\item The the cell's subcluster $T_{i+1}$ effects the distribution of its latent features at the $i$th level, $Z_i$.
-That is, the latent distribution on the $i$th latent representation is governed by what subcluster it is part of within $T_i$.
-\item The distribution of the final latent representation, $Z_T$, is simply a standard normal.
-\item The functions $h_{T_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
+\item The latent representation $Z^c_i$ corresponds to features of that cell which are specific to all the cells which are descendents of $T^c_i$.
+\item The the cell's subcluster $T^c_{i+1}$ effects the distribution of its latent features at the $i$th level, $Z^c_i$.
+That is, the latent distribution on the $i$th latent representation is governed by what subcluster it is part of within $T^c_i$.
+\item The distribution of the final latent representation, $Z^c_T$, is simply a standard normal.
+\item The functions $h_{T^c_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
 \end{itemize}
@@
 
-At the current moment, we have only actually implemented this model in the case the $H$ is a tree with depth 1, i.e.
-the hierarchy is trivial.
+At the current moment, we have only actually implemented this model in the case the $H$ is a tree with depth 2 (i.e.
+one root node and some children).
 We take the data model to be a zero-inflated negative binomial.

### 2018-03-23 trygve/outline.tex
Added cell indexing

@@
 
-There is one possibly confusing aspect of our output that might be nice to address.
+There is one possibly confusing aspect of our output that we would like to to address.
 The upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.
@@
 This is because we have taken \emph{elementwise bounds} on each entry.
-However, it would be nice to sample uniformly from the set of possible values of $q$ which are consistent with our observations.
-MCMC or rejection sampling might be good options to investigate.
-These would help give the user another way to understand what possible values $q$ might take on.
+An alternative is to  sample uniformly from the set of possible values of $q$ which are consistent with our observations.
+We hope this may give the user another way to understand what possible values $q$ might take on.
 

### 2018-03-23 generativemodelsforscrna/outline.tex
Added plate diagram

@@
 
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/plate}
+\caption{Plate diagram for the model \label{fig:plate}}
+\end{figure}
+
 Towards this end, we propose the following generative model for gene expression of a single cell.
@@
 \[
-\lambda^c_g = f_g \left(\sum_{i=0}^{\ell} h_{T^c_i}(Z^c_{T^c_i})\right)
+\lambda^c_g = f_g \left(\sum_{i=0}^{\ell} h_{g,T^c_i}(Z^c_{T^c_i})\right)
 \]
@@
 
-Let us unpack some of this:
+This can be a bit confusing to take in.  The plate diagram in Figure \ref{fig:plate} may help.  We can also try to unpack some of this in words:
 
@@
 \item The distribution of the final latent representation, $Z^c_T$, is simply a standard normal.
-\item The functions $h_{T^c_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
+\item The functions $h_{g,T^c_i}$ indicates how these cell-type-specific features alter the distribution on the gene expressions for cells in that branch of the tree.  
 \end{itemize}

### 2018-03-23 generativemodelsforscrna/outline.tex
minor clarification
better plate diagram
Added sampler stuff

@@
 The Allen Institute has developed a method for clustering, one that has apparently allowed them to distinguish 116 unique cell types using single cell RNA data.
-The method follows the following iterative procedure:
+The method follows the following recursive procedure:
 
@@
 \includegraphics[width=0.8\textwidth]{pics/plate}
-\caption{Plate diagram for the model \label{fig:plate}}
+\caption{Plate diagram for the model\label{fig:plate}.  $i$ indexes layers of the hierarchy of cell types, $n$ indexes nodes in the hierarchy of cell types,  $c$ indexes cells, and $g$ indexes genes.}
 \end{figure}

### 2018-03-23 trygve/outline.tex
Added sampler stuff

@@
 
+To this end, we use a Dikin ellipsoid sampler to approximate uniform samples from the set of valid $q$s which are consistent with what we've observed.
+These are shown in Figure \ref{fig:allenavg}.
+Whether the MCMC method is mixing or not is of course up for debate.
+See Appendix \ref{sec:dikin} for some details.
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenavg}
+\caption{Allen institute data: sampling information.
+The first plot is the estimated $q(y|x)$, as we have already seen in other figures.
+The second plot is a single sample of $q(y|x)$,  The third plot is an average value of each entry of $q$ over many samples.
+The final plot indicates the standard deviation of each entry over the samples.
+\label{fig:allenavg}}
+\end{figure}
+
 \section{The conjecture}
@@
 
+\appendix
+
+\section{Dikin sampler}
+
+\label{sec:dikin}
+
+Consider a convex polytope $T=\{x:\ Ax\leq b\}$.
+We have implemented a method for sampling from this polytope, based on the paper \citep{kannan2012random}.
+This method makes use of the Dikin ellipsoids, $E(x)$.
+For any $x$, these are defined by
+
+\begin{itemize}
+    \item Computing the distance from $x$ to each facet of the polytope, i.e. $d_i = b_i - \sum_j A_{ij} X_j$.
+    \item Constructing $\tilde A$ as $\tilde A_{ij} = A_{ij} / d_i$.
+    \item Define $E(x) = \{y:\ |\tilde A (X-y)|\leq 1\}$.  
+\end{itemize}
+
+We can use these ellpsoids to efficiently sample the polytope $T$.
+At each step, we have some point $X\in T$, and we would to use this point to obtain a new sample $Y$, such that by iterating this process we asymptotically obtain samples which are uniform in $T$.
+Here is how we use $X$ to get $Y$:
+
+\begin{algorithm}[H]
+ \KwData{A point $X\in T$}
+ \KwResult{A point $Y\in T$}
+ \vspace{.1in}
+  Sample a proposal $\tilde Y$, uniformly from $E(X)$\;
+    \eIf{$X \in E(\tilde Y)$}{
+        Sample $U \sim \mathrm{Uniform}[0,1]$\;
+        \eIf{$U \leq \mathrm{Vol}(E(X))/\mathrm{Vol}(E(\tilde Y)) \leq 1$}{
+            Let $Y \gets \tilde Y$\;
+        }{
+            Let $Y \gets X$\;
+        }
+    }{
+        Let $Y\gets X$\;
+    }
+ \caption{Dikin sampler step}
+\end{algorithm}
+
+It is easy to show that the stationary distribution of the Markov chain found by iterating these Dikin sampler steps is indeed uniform on $T$.
+To ensure an numerically robust method in the face of high-dimensional and nearly degenerate matrices, we take the following approach to robustly sampling from the ellipsoid:
+
+\begin{algorithm}[H]
+ \KwData{An $n\times m$ matrix $\tilde A$}
+ \KwResult{A point $X$ sampled uniformly from $\{x:\ |Ax|\leq 1\}$}
+ \vspace{.1in}
+  Sample $Z$ as an $n$-dimensional normal variables vector\;
+  Let $X$ denote the solution to the least squares problem $\min_x|\tilde Ax - Z|$\;
+  Normalize $X$ by $X \gets X/|\tilde AX|$\;
+  Sample $U \sim \mathrm{Uniform}[0,1]$\;
+  Scale $X$ by $X \gets X \times U^{1/m}$\;
+ \caption{Ellipsoid sampler}
+\end{algorithm}
+
 \end{document}

### 2018-03-23 trygve/outline.tex
Added widths information

@@
 
-There is one possibly confusing aspect of our output that we would like to to address.
-The upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.
+Note that the upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.
 For example, the rows of the upper bounds generally sum to more than 1.
@@
 We hope this may give the user another way to understand what possible values $q$ might take on.
-
 To this end, we use a Dikin ellipsoid sampler to approximate uniform samples from the set of valid $q$s which are consistent with what we've observed.
@@
 See Appendix \ref{sec:dikin} for some details.
+
+Another way to look at the space of possible values of $q$ which are consistent with the observables is to look at the width of the polytope in various directions.
+That is, select a direction uniformly at random, and measure the width of the polytope along that direction.
+Note that this is not at all the same as selecting a random direction and seeing how far one can travel before hitting any boundary; obtaining the full width in a given direction requires solving a linear program.
+Doing this procedure many times and taking a histogram yields Figure \ref{fig:allenwidths}.
+This suggests that the polytope may be only about 7 units in length; distributed over 1624 entries of the matrix, each of which lies in $[0,1]$; this suggests that the polytope may indeed be  constrained to be close to the output we showed in Figure \ref{fig:allenoutput}.
 
@@
 \label{fig:allenavg}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenwidths}
+\caption{Allen institute data: the width of the polytope.\label{fig:allenwidths}}
 \end{figure}

### 2018-03-29 generativemodelsforscrna/outline.tex
new residual plots

@@
 
-\subsection{Posterior predictive means}
-
-The posterior predictive mean can be interpreted as a ``denoised'' version of the original.  In this section we examine the Allen data from that point view.
-
-For example, here we have selected 1000 random entries from the gene by cell matrix.
-For each entry, we plot a point in which the horizontal location indicates the true value of that entry, and the vertical position indicates the posterior predictive expectation of that entry.
-
-\includegraphics[width=0.8\textwidth]{pics/resids2}
-
-We can compare this with a similar plot using posterior predictive samples instead of the true values.  This then forms a different kind of posterior predictive check.
-
-\includegraphics[width=0.8\textwidth]{pics/resids3}
-
-The fact that these two plots look the same is encouraging.
-
-We can also use heatmaps to try to give more structure to our understanding of this ``denoising'' process.
-Here we have three plots: one giving the original gene expression for the first 1000 cells, one giving the posterior expected gene expression for the same cells, and one giving the residual.
-In all cases we have transformed the data by taking $\log(1+x)$ and subtracting gene-level means.
-Each column corresponds to a gene, and each row corresponds to a cell.
-We have sorted the cells by one of the latent components, and sorted the genes by using a property of the gene decoder.
+Another way to see that the model isn't quite right is to look at posterior predictive p values.
+That is, for each entry of the matrix, we can compute the cumulative distribution function of that entry in light of the parameters predicted by the posterior mode.
+Here we have done that for the first 1000 entries; the results are not exactly encouraging:
 
@@
 
-Note that there is a group of cells which are much more difficult to understand, which are found at the bottom of the plot.  These are mostly astrocytes.
+We see that for some genes the $p$ values (which should be distributed uniformly) are almost always near 1, and in other cases they are almost always near 0.
+We can dig deeper into this by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.
+There are clearly some systematic differences.
 
-Since we have 20000 cells, it might be nice to make a similar plot for all the cells.
-However, such a plot would be very tall and perhaps difficult to look at.
-We can instead take the full 20000 x 1400 plot and average neighboring rows and columns together.
-This gives a plot like this:
+\includegraphics[width=0.8\textwidth]{pics/residsdetail}
 
-\includegraphics[width=0.6\textwidth]{pics/residsfull}
+One potential reason for this might be that the negative binomial data distribution is inappropriate.
+To diagnose this, we fitted a zero-inflated negative binomial distribution to each collection of cells Allen Institute dubbed a ``cluster.''  We then measured the goodness-of-fit, in terms of whether various moments matched.
+In particular, we compared the empirical mean, standard deviation, and proportion of zeros to the corresponding statistics which would be predicted by the learned models.
 
-If there was a lot of structure in the residual, we would feel that perhaps there was something important and obvious we were missing.  Since there isn't, we are reasonably encouraged.
+\includegraphics[width=0.8\textwidth]{pics/zinbfitmatch}
+
+In terms of mean and zero proportion, these seem nearly okay, but in terms of variance we can see that we are not doing the greatest job.
+For a great many cell types we are predicting standard deviations which are much higher than the reality.
+This is perhaps not surprising, since the negative binomial model is inherently overdispersed, and simply cannot accomodate underdispersed data.
+
+Another possible source of trouble is that in the particular model we use we have made certain assumption about how $\alpha,\beta,h$ can vary.
+This could be a big problem.
+To see why, we looked at each the fitted parameters of the zero-inflated negative binomial distribution for each cell-type.
+This yields a kind of three-dimensional scatterplot, in which each dot is a cell type, and the location indicates the value of each of its three fitted parameters.
+Here we show three projections of that scatterplot:
+
+\includegraphics[width=0.8\textwidth]{pics/zinbfitdiversity}
+
+Clearly there is quite a lot of variety in terms of how these parameters may vary.
+They do not appear to lie on any kind of low-dimensional manifold.
+So we must be very careful in terms of what assumptions our model makes about how these parameters may be tied together.
 

### 2018-03-29 trygve/outline.tex
new residual plots

--- 2018-03-23 trygve/outline.tex
@@
 See Appendix \ref{sec:dikin} for some details.
+
+Another way to look at the space of possible values of $q$ which are consistent with the observables is to try to understand the scope of the polytope along various directions.
+In this effort, we selected a direction uniformly at random and found the two extremal vertices in this direction.
+We then computed the distance between those vertices, as well as the distance between those vertices projected to the direction.
+Doing this procedure many times and making a scatterplot yields Figure \ref{fig:allenwidths}.
+This figure suggests that the polytope may be only about 8 units in diameter, although we caution that general results on polytope diameter estimation are not encouraging, cf.
+\citep{brieden1998approximation}.
+However, considering that the magnitude of this diameter would be distributed over 1624 entries of the matrix, each of which lies in $[0,1]$, it would seem that the polytope may indeed be constrained to be very close to the output we showed in Figure \ref{fig:allenoutput}.
 

### 2018-03-29 generativemodelsforscrna/outline.tex
typo

@@
 That is, for each entry of the matrix, we can compute the cumulative distribution function of that entry in light of the parameters predicted by the posterior mode.
-Here we have done that for the first 1000 entries; the results are not exactly encouraging:
+Here we have done that for the first 1000 entries.
+The results are not exactly encouraging:
 
@@
 
-We see that for some genes the $p$ values (which should be distributed uniformly) are almost always near 1, and in other cases they are almost always near 0.
+We see that for some genes the $p$ values (which should be distributed uniformly) are almost always near 1.
 We can dig deeper into this by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.
-There are clearly some systematic differences.
+There are clearly some systematic differences:
 
@@
 
-One potential reason for this might be that the negative binomial data distribution is inappropriate.
+One potential reason for this systematic error might be that the negative binomial data distribution is inappropriate.
 To diagnose this, we fitted a zero-inflated negative binomial distribution to each collection of cells Allen Institute dubbed a ``cluster.''  We then measured the goodness-of-fit, in terms of whether various moments matched.

### 2018-04-02 generativemodelsforscrna/outline.tex
consistency proof, such as it is

@@
 
-One limitation of the method as we have implemented it is that the posterior predictive distribution seems to be substantially underfitting.  To see what we mean, consider the following figure:
+One limitation of the method as we have implemented it is that the posterior predictive distribution seems to be substantially underfitting.  To see what we mean, consider Figure \ref{fig:Z1gn}:
 
+\begin{figure}
 \includegraphics[width=0.8\textwidth]{pics/Z1gn}
+\caption{\label{fig:Z1gn}}
+\end{figure}
 
@@
 That is, for each entry of the matrix, we can compute the cumulative distribution function of that entry in light of the parameters predicted by the posterior mode.
-Here we have done that for the first 1000 entries.
-The results are not exactly encouraging:
+In Figure \ref{fig:resids} we have done that for the first 1000 entries.
 
+\begin{figure}
 \includegraphics[width=0.8\textwidth]{pics/resids}
+\caption{\label{fig:resids}}
+\end{figure}
 
-We see that for some genes the $p$ values (which should be distributed uniformly) are almost always near 1.
-We can dig deeper into this by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.
-There are clearly some systematic differences:
+These results are actually fairly encouraging.
+Mostly it looks basically like white noise.
+However, we do see that for some genes the $p$ values are almost always near 1: these are genes which express essentially never, and so have a high chance of turning off, and so their $p$ value is nearly $1$ whenever they are off, which is virtually always.
+If we restrict ourselves to those genes whose average probability of going off is less than 20\%, we obtain Figure \ref{fig:residslesszeros}.
 
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/residslesszeros}
+\caption{\label{fig:residslesszeros}}
+\end{figure}
+
+Another way to dig deeper into the underfitting question is by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.
+There are clearly some systematic differences, as we see in Figure \ref{fig:residsdetail}.
+
+\begin{figure}
 \includegraphics[width=0.8\textwidth]{pics/residsdetail}
+\caption{\label{fig:residsdetail}}
+\end{figure}
 
@@
 In particular, we compared the empirical mean, standard deviation, and proportion of zeros to the corresponding statistics which would be predicted by the learned models.
+We show this in Figure \ref{fig:zinbfitmatch}.
 
+\begin{figure}
 \includegraphics[width=0.8\textwidth]{pics/zinbfitmatch}
+\caption{\label{fig:zinbfitmatch}}
+\end{figure}
 
@@
 This yields a kind of three-dimensional scatterplot, in which each dot is a cell type, and the location indicates the value of each of its three fitted parameters.
-Here we show three projections of that scatterplot:
+Here we show three projections of that scatterplot in Figure \ref{fig:zinbfitdiversity}.
 
+\begin{figure}
 \includegraphics[width=0.8\textwidth]{pics/zinbfitdiversity}
+\caption{\label{fig:zinbfitdiversity}}
+\end{figure}
 

### 2018-04-02 generativemodelsforscrna/paper.tex
consistency proof, such as it is

--- 2018-03-29 generativemodelsforscrna/outline.tex
@@
 Here we have done that for the first 1000 entries.
-The results are not exactly encouraging:
 
@@
 
-We see that for some genes the $p$ values (which should be distributed uniformly) are almost always near 1.
-We can dig deeper into this by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.
+These results are actually fairly encouraging.
+Mostly it looks basically like white noise.
+However, we do see that for some genes the $p$ values are almost always near 1: these are genes which express essentially never, and so have a high chance of turning off, and so their $p$ value is nearly $1$ whenever they are off, which is virtually always.
+
+Anotehr way to dig deeper into the underfitting question is by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.
 There are clearly some systematic differences:

### 2018-04-02 trygve/paper.tex
consistency proof, such as it is

% !TEX program = pdflatex
\input{preamble/preamble.tex}
\input{preamble/preamble_math.tex}
\input{preamble/preamble_acronyms.tex}

\DeclareMathOperator*{\tr}{tr}

\newcommand{\UN}[1]{\ensuremath{\left|#1\right|_\infty}}
\newcommand{\EN}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\kldiv}[2]{\ensuremath{D\left(#1||#2\right)}}

\usepackage{cancel}

% \linenumbers

\title{Identifying the unidentifiable: combining knowledge from destructive measurements}
\author{Jackson Loper}

\usepackage{amsthm}
\newtheorem*{thm}{Theorem}
\newtheorem{lem}{Lemma}
\newtheorem{conj}{Conjecture}

\begin{document}
\maketitle

\begin{abstract}
When many different measurement tools are applied, how can we best combine information from the different different tools?  If we can apply multiple measurement tools to the same specimen, we can at least begin to understand how the tools they are related.  We call this ``joint measurement.''  What can we do if joint measurement is unavailable?  In this paper we investigate a simple assumption that can make it possible do obtain joint distributions even when joint measurement isn't possible.
\end{abstract}

The modern setting is rife with experimental methodologies, and it can be very frustrating to understand how the output of these methods relate to one another.  This is particularly difficult if joint measurement is unavailable, i.e.\ it is impossible or impractical to observe a single specimen with multiple methodologies.  For example, after processing a neuron with a given single-cell rna sequencing technique, we cannot generally go back and process that same neuron with a different technique.   This makes it seemingly impossible to calibrate the techniques against each other.  By contrast, it is easy to calibrate thermometers against each other, because we can simply measure the same water with two different thermometers and see how the measurements relate.  Without joint measurement, this isn't possible.

The absence joint measurement gives rise to a number of problems:

\begin{itemize}

\item Different tools may teach us different things, and we need to combine tools
to learn everything we can.  Let $A,B$ denote two different quantities of interest about cells (e.g. morphology and gene expression).   Let us say one tool enables us to measure $A$ and another tool to measure $B$.  We would like
to be able to say something about what morphologies are associated with what gene expressions.

\item Two labs may used different methods, but may wish to pool results.  Each lab looks at the other
lab's data and asks themselves "what would their data have looked like if they had used our method?"  When
they find they don't know the answer, it greatly complicates collaboration.

\item One may attempt to cluster the data.  It is simple to cluster the data from one technique, and cluster the data
from another technique -- but how can we know how those clusters correspond to one another?

\end{itemize}

In this work we show that it is indeed sometimes possible to get useful and statistically rigorous bounds for these kinds of questions, even when joint measurements are unavailable.   In particular, we consider the case that we can obtain samples from a variety of subpopulations.  If we can assume that the \emph{relationship} between the measurements is the same in each subpopulation, we can use these samples to learn something about that relationship -- even though we never see both measurements for any single member of the population.

We apply our method to single cell RNA data.  We show how different clusterings from different single cell RNA modalities can be connected.  Our method yields results which are generally consistent with the scientist's understanding of the data and clusters, but also reveal potential gaps which could be important for further study.

This work stands on the shoulders of a long history of relating probabilistic assumptions to probabilistic inequalities.  Much of this literature comes from research into causality.  For example, in \cite{bonet2001instrumentality} Bonet uses polytopes not unlike the ones seen here to explore whether a variable can be used as an instrument.  The famous Clauser-Horne-Shimony-Holt inequality was designed to help answer causality questions in quantum physics, but it also sheds light on what distributions are consistent with certain assumptions \cite{clauser1969proposed}.  Indeed the physics literature has contributed many key inequalities (cf. \cite{chaves2014inferring}, \cite{kela2017semidefinite}, and the references therein).  Perhaps the closest work to this one would be \cite{makarov1982estimates}, which used two marginal distributions to get bounds on a property of the joint distribution (namely the distribution of the sum).  We build on this approach, both by using subpopulations to considerably refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.

\section{Mathematical formulation}

\begin{figure}
\hfill{}\includegraphics[width=.3\textwidth]{pics/plate}\hfill{}
\caption{We assume that we have two datasets, both governed by the same process.  In the top dataset, the results of technique II, denoted by $Y_i$, are hidden from us (we indicate this by making the circle gray).  In the bottom dataset, the results of technique I, denoted by $X_i$, are hidden from us.  In each dataset, we know what subpopulation of specimens we are considering, and this is designated by $\ell_i$.  Thus, for any given specimen $i$ we can observe either $\ell_i,X_i$ or $\ell_i,Y_i$ but never $\ell_i,X_i,Y_i$.  Using observations of $X_i$ we can certainly learn the parameters $p$ which govern the relationship between $X_i$ and the subpopulation, $\ell_i$.  However, if we believe that the relationship between $X$ and $Y$ is the same regardless of the population value $\ell$, we can also learn something about the parameters $q$ which govern the relationship between $X$ and $Y$.\label{fig:plate}}
\end{figure}

Mathematically, we formulate our model as follows:

\begin{itemize}
\item Let us say we have $n+m$ individual members of a population.  These could be cells, humans, or whatever the smallest sample unit may be for a given problem.
\item For each individual member, we have some readily observable features by which we can group the members, such as the size of a cell, where an individual lives, or some other basic information.  We designate these features as $\ell_i$ for the $i$th individual.
\item For the first $n$ members, we have made an observation using the first technique.  This gives us the values $X_1 \cdots X_n$.
\item The last $m$ members were observed using the second technique.  This gives us the values $Y_{n+1} \cdots Y_{n+m}$.
\end{itemize}

For simplicity, we here assume that $\ell_i,X_i,Y_i$ take values in finite sets, $\Omega_\ell,\Omega_X,\Omega_Y$.  Extensions to more general cases should be straightforward, but we must leave them to future work.

We further posit a set of \emph{counterfactual} variables -- variables which could not ever be obtained in practice, but which help us organize our thinking.  These are sometimes referred to as ``potential outcomes''  (cf. \cite{rubin2005causal}).

\begin{itemize}
\item Let $Y_1 \cdots Y_n$ denote the observations we \emph{would have gotten} if we had applied the second technique to the first $n$ members.
\item Let $X_{n+1} \cdots X_n$ denote the observations we would have gotten if we had applied the first technique to the last $m$ members. 
\end{itemize}

Our main assumption is that

\begin{equation}\label{eq:mainassumption}
\mathbb{P}(X_i=x,Y_i=y|L_i=\ell)=p^*(x_i|\ell_i)q^*(y_i|x_i)
\end{equation}

for some distributions $p^*,q^*$.  Moreover, we assume that the vectors $\{(p^*(x_1|\ell)\cdots p^*(x_n|\ell))\}_\ell$ are linearly independent.  If this seems unlikely, it may make sense to collapse similar values of $\ell$ together.  This assumption can be articulated by the plate diagram found in Figure \ref{fig:plate}.

The validity of these assumptions for a given situation should be closely contemplated.  There are several key questions to answer when deciding whether this assumption is applicable.  Does the process by which samples are gathered gathered depend only upon $\ell$?  In particular, is it not at all statistically related to which measurement technique was  applied, except through $\ell$?  Are the particular measurement biases of both techniques the same for every value of $\ell$?  Is the statistical relationship between the quantities being measured by the two techniques the same for every value of $\ell$?\footnote{Note that this is automatically true if both techniques are measuring the same thing.  More generally, if two techniques have something that they both measure, we can restrict attention to that one common phenomenon.}  If the answer to all of these questions is yes, the assumption that $(X_i,Y_i)$ is drawn from $p(x_i|\ell_i)p(y_i|x_i)$ may apply.

We can now articulate our goal.  Informally, our goal is to see what we can understand about the relationship between the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$, i.e. the distribution $q^*$.  Unfortunately, given the limitations of the data, this may not be possible.  Indeed, it is easy to see that $q$ is dramatically unidentifiable in many cases.

For example, consider the case that $\ell$ lies in the set ${1,2}$, $X$ lies in the set ${1,\cdots 100}$ and $Y$ lies in the set ${1,2}$.  As long as $m,n$ are large enough, we can effectively estimate $p^*$ and $h^*(Y|\ell)=\sum_x p^*(x|\ell)q(Y|x)$.  However, this does not tell us very much about what $q$ might be.  Indeed, $q$ lies in a 100-dimensional space, and the restriction that $q$ must satisfy $h^*(Y|\ell)=\sum_x p^*(x|\ell)q(Y|x)$ for each $\ell,Y$ actually only introduces 2 new constraints.  Thus there is a 98-dimensional space that $q$ may lie in, and we simply cannot know where in that space $q$ may lie.  However, in practice, we find that the \emph{inequality constraints}, namely that $q(y|x)>0$ for every $x,y$, can actually force this apparently vast space to be tightly centered around a single point.

This suggests a somewhat less amibtious goal: to understand the \emph{set of values} of $q$ which are consistent with the data.  Specifically, we will seek to produce an estimator for the set

\begin{equation*}
\Theta^* \triangleq \left\{q:\ \sum_y p^*(x|\ell)q(y|x) = \sum_y p^*(x|\ell)q^*(y|x)\ \forall \ell,y\right\}
\end{equation*}

\section{Algorithm}

We begin by estimating

\begin{itemize}
\item $p^*$ using the empirical distribution of the observations $\{X_i\}$, $\hat p(x|\ell)$
\item $h^*(y|\ell)\triangleq \sum_x p^*(x|\ell)q^*(y|\ell)$ using the empirical distribution of $\{Y_i\}$, $\hat h(y|\ell)$
\end{itemize}

These empirical distributions may not be quite consistent with the original assumption that the distribution of $X,Y|\ell$ may be written as $p^*(x|\ell)q^*(y|x)$.  In particular, it may be that there is \emph{no} value of $q$ such that $\sum_x \hat p(x|\ell) q(y|x) = \hat h(y|\ell)$.  There may also be many such values.  To obtain a well-defined estimator, we take a value of $q$ which is decent, and use a slight regularization to ensure it is uniquely defined.  Let $N_{X,\ell}=\#\{i\leq n:\ \ell_i=\ell\},N_{Y,\ell}=\#\{i> n:\ \ell_i=\ell\}$.  We take

\begin{equation*}
\hat q = \argmax_q \sum_\ell N_{Y,\ell}\sum_{y}\hat h(y|\ell)\log\left(\sum_{x}\hat p(x|\ell)q(y|x)\right) + \kappa \sum_{xy} \log q(y|x)
\end{equation*}

Note in the second term we apply an entropic penalty to $q$.  This is both necessary to ensure that $\hat q$ is uniquely defined in terms of $\hat q,\hat h$, as well as a useful regularization. This regularization pushes us slightly towards uniform distributions, which makes sense if we believe that $q(y|x)>0$  for each $x,y$.  Other regularizations could also be used and would potentially enjoy similar theoretical guarantees to the ones we will show below.

This method gives us an estimate for $\hat q$, but we emphasize this estimate is very likely to be \emph{inconsistent} for the true value of $q^*$.  As we described in the previous section, this is simply a limitation of the data we have available -- attempting to deduce $q^*$ may be too much to ask for.  However, subject to suitable assumptions, the estimator
\begin{equation*}
\hat \Theta\triangleq \left\{q:\ \sum_y p(x|\ell)q(y|x) = \sum_y p(x|\ell)\hat q(y|x)\ \forall \ell,y\right\}
\end{equation*}
is indeed consistent in the sense that we can be asymptotically assured that the true $q^*$ lies very close to some point in $\hat \Theta$.

\vspace{.1in}
\begin{thm}
Fix any $\kappa>0$.  Let $N_{X,\ell},N_{Y,\ell}\rightarrow\infty$ in such a way that $N_{Y,\ell'}/\sum_{\ell}N_{Y,\ell} \geq \rho>0$ for each $\ell'$.  Let us assume that $q^*(y|x)>0$ for every $x,y$ and the rows of $p^*$ are linearly independent.  Then $\inf_{q\in \hat\Theta} \sup_{x,y}|q^*(y|x)-q(y|x)|\rightarrow 0$ in probability.
\end{thm}

\begin{proof}
We defer the proof to Appendix \ref{sec:proof}.
\end{proof}

We close by remarking that the assumption $q^*(y|x)>0$ is very likely to be unnecessary.  However, it substantially simplifies the proof.  We therefore leave a more complete theorem for future work.

\section{Empirical results}

Our motivation for this problem arose from looking at Allen Institute cell-type assignment of cells, performed using two different modalities.  The only thing connecting the two modalities was that each modality had a variety of samples drawn from a variety of sub-populations, and the same sub-populations were used for both modalities.  Thus our input was two tables: (technique I cluster $\times$ sub-population), and (technique II cluster $\times$ sub-population) (Figure \ref{fig:alleninput}).  Our output was a polytope $\hat \Theta$ of possible ways that technique I clusters could line up against the technique II clusters (Figure \ref{fig:allenoutput}).

Empirically, we see that this polytope appears to be very small, and tightly centered around our estimate for $\hat q$.  Thus $\hat q$ may indeed serve as a useful indication of how the different technique's clusters line up.  There are certain aspects of $q$, however, which are more uncertain than others.  We were able to measure this by sampling uniformly from the polytope $\hat \Theta$ and seeing which entries varied the most (cf. Appendix \ref{sec:dikin}).  Detecting these entries is useful because it suggests what further experiments would be helpful to more fully clarify the relationship between the two modalities.

\begin{figure}
\begin{minipage}[c]{.3\textwidth}
\includegraphics[width=.9\textheight,angle=270]{pics/alleninput}
\end{minipage}
\hfill
\begin{minipage}[l]{.4\textwidth}
\caption{Input: two tables.  The Allen Institute had access to various cre-line-based cell selection techniques.  Each technique pulls out a different group of cells.  Once the cells were selected, they were then either subjected to technique I (`facs') or technique II (`patch').  Technique I gives a very complete analysis of the gene expression of the cell.  Technique II gives a less complete analysis, but yields additional electrophysiological data that may be of interest.  The results of technique I was used to assign it to one of 116 categories, and the results of technique II was used to assign it to one of 14 categories.  The tables above show the distribution of category assignment for each selection technique.  The goal is to use these tables to be able to say something about how the 14 categories from Technique II might line up with the 116 categories from technique I.  \label{fig:alleninput}}
\end{minipage}

\end{figure}

\begin{figure}
\includegraphics[width=0.8\textwidth]{pics/allenavg}
\caption{\label{fig:allenoutput}In the same setup as Figure \ref{fig:alleninput}.  Using our method we produce a set of possible $q$s which are consistent with the data tables given.   Using our method we first produce a single table, $\hat q$, that suggests how the outputs of the two techniques might line up.  We are able to do this even though we were never able to observe a cell processed by both techniques.    We emphasize, however, that this table is not guaranteed to be a consistent estimator for the true value of $q^*$.  However, we can guarantee that asymptotically the true $q^*$ will lie very close to a certain set $\hat \Theta$, containing $\hat q$. Fortunately, this set appears to be very small indeed, and tightly centered around $\hat q$.  To see this, we used a Dikin sampler to sample uniformly from $\hat \Theta$.  We saw that almost every sample was very close to $\hat q$.  The second plot shows an example of such a sample.  The third plot is an average value of each entry of $q$ over many samples.  The final plot indicates the standard deviation of each entry over the samples.  Where this is high we have some substantial ambiguity, and so it may be important to perform further experiments to narrow down the correspondence.  \label{fig:allenavg}}
\end{figure}

Another way to look at $\hat \Theta$ is to try to understand its scope of the polytope along various directions.  In this effort, we selected a direction uniformly at random and found the two extremal vertices in this direction.  We then computed the distance between those vertices, as well as the distance between those vertices projected to the direction.  Doing this procedure many times and making a scatterplot yields Figure \ref{fig:allenwidths}.  This figure suggests that the polytope may be only about 8 units in diameter, although we caution that general results on polytope diameter estimation are not encouraging, cf. \citep{brieden1998approximation}.  However, considering that the magnitude of this diameter would be distributed over 1624 entries of the matrix, each of which lies in $[0,1]$, it would seem that the polytope may indeed be constrained to be very close to the output we showed in Figure \ref{fig:allenoutput}.

\begin{figure}
\includegraphics[width=0.8\textwidth]{pics/allenwidths}
\caption{How wide is $\hat \Theta$?  Each dot above corresponds to a randomly selected direction.  The horizontal position of the dot indicates the width of the polytope along that direction, and the vertical position indicates the distance between the two extremal vertices along that direction.\label{fig:allenwidths}}
\end{figure}

\section{Conclusions}

When joint measurement is impossible, it can be difficult to calibrate two methods against each other or understand how they may be related.  Here we show that a simple Markov assumption can make it possible to actually learn quite a lot.  Although the exact relationship may not be identifiable, a polytope of possible relationships can be identified, and this polytope may in fact be quite small indeed.  By exploring this polytope, we can understand what we know -- and what we don't know -- about the relationship between measurements.

The Markov assumption is of course not the only one that we could have used, and may not be valid in every case.  For example, it has been speculated that some cell types tend to die more often in one experimental modality than another, and these cells are not part of the data.  This violates our assumptions.  However, assuming this death rate can be roughly measured, it can be adjusted for, yielding a different but equally meaningful assumption about the data.  Moreover, if this method yields bizarre results, it may give useful clues as to exactly how cell death may happen differently in the two modalities.

Once we accept that what we're interested in may not be fully identifiable, any of a wide variety of assumptions can help us obtain practical bounds.  Although we may not be able to learn exactly what we want, we can learn a set of possibilities.  By probing this set carefully with uniform samplers and extremal tests, we can learn what the data actually has to say and what experiments we need to do to learn more.

\bibliographystyle{unsrt}
\bibliography{refs}

\appendix

\section{Dikin sampler}

\label{sec:dikin}

Consider a convex polytope $T=\{x:\ Ax\leq b\}$.  We have implemented a method for sampling from this polytope, based on the paper \citep{kannan2012random}.  This method makes use of the Dikin ellipsoids, $E(x)$.  For any $x$, these are defined by

\begin{itemize}
    \item Computing the distance from $x$ to each facet of the polytope, i.e. $d_i = b_i - \sum_j A_{ij} X_j$.
    \item Constructing $\tilde A$ as $\tilde A_{ij} = A_{ij} / d_i$.
    \item Define $E(x) = \{y:\ |\tilde A (X-y)|\leq 1\}$.  
\end{itemize}

We can use these ellpsoids to efficiently sample the polytope $T$.    At each step, we have some point $X\in T$, and we would to use this point to obtain a new sample $Y$, such that by iterating this process we asymptotically obtain samples which are uniform in $T$.  Here is how we use $X$ to get $Y$:

\begin{algorithm}[H]
 \KwData{A point $X\in T$}
 \KwResult{A point $Y\in T$}
 \vspace{.1in}
  Sample a proposal $\tilde Y$, uniformly from $E(X)$\;
    \eIf{$X \in E(\tilde Y)$}{
        Sample $U \sim \mathrm{Uniform}[0,1]$\;
        \eIf{$U \leq \mathrm{Vol}(E(X))/\mathrm{Vol}(E(\tilde Y)) \leq 1$}{
            Let $Y \gets \tilde Y$\;
        }{
            Let $Y \gets X$\;
        }
    }{
        Let $Y\gets X$\;
    }
 \caption{Dikin sampler step}
\end{algorithm}

It is easy to show that the stationary distribution of the Markov chain found by iterating these Dikin sampler steps is indeed uniform on $T$.  To ensure an numerically robust method in the face of high-dimensional and nearly degenerate matrices, we take the following approach to robustly sampling from the ellipsoid:

\begin{algorithm}[H]
 \KwData{An $n\times m$ matrix $\tilde A$}
 \KwResult{A point $X$ sampled uniformly from $\{x:\ |Ax|\leq 1\}$}
 \vspace{.1in}
  Sample $Z$ as an $n$-dimensional normal variables vector\;
  Let $X$ denote the solution to the least squares problem $\min_x|\tilde Ax - Z|$\;
  Normalize $X$ by $X \gets X/|\tilde AX|$\;
  Sample $U \sim \mathrm{Uniform}[0,1]$\;
  Scale $X$ by $X \gets X \times U^{1/m}$\;
 \caption{Ellipsoid sampler}
\end{algorithm}

\section{Proof of the theorem}

\label{sec:proof}

For the benefit of the reader, we here repeat the statement of our theorem in more explicit terms.

\begin{itemize}
\item Let $\UN{\cdot}$ denote the uniform norm (i.e.\ the maximum absolute value) and $\EN{\cdot}$ denote the Euclidean norm (i.e.\ the square root of the sum of the squares).  In the case of matrices, this Euclidean norm goes by the name of the Frobenius norm.  Recall that in this norm matrices satisfy a Cauchy-Schwarz like equality, $\EN{pq}\leq\EN{p}\EN{q}$.  Also recall that $\UN{a} \leq EN{a} \leq \sqrt{n}\UN{a}$ where $n$ is the number of entries in $a$.
\item Let $T_{a,b}$ denote the transition matrix polytope, i.e. the set of $a\times b$ matrices whose rows sum to 1 and whose entries are all positive.
\item Let $|\Omega_\ell|,|\Omega_X|,|\Omega_Y| \in \mathbb{N}$.
\item Let $p^*\in T_{|\Omega_\ell|,|\Omega_X|}$.
\item Let $q^*\in T_{|\Omega_X|,|\Omega_Y|}$.
\item We require the matrix $q^*$ has strictly positive entries, $q^*_{xy}\geq c>0$.
\item We require that the rows of $p^*$ are linearly independent.
\item Let $\hat p$ denote an empirical transition matrix drawn by obtaining $N_{X,\ell}$ samples for each row of $p^*$, i.e. we have samples $(\ell_1,X_1) \cdots (\ell_1,X_n)$ such that $\mathbb{P}(X_i=x) = p^*_{\ell_i,x}$, $N_{X,\ell}=\sum_{i=1}^n \mathbb{I}_{\ell_i=\ell}$, and $\hat p_{\ell x} = \sum_{i=1}^n \mathbb{I}_{X_i=x,\ell_i=\ell} / N_{X,\ell}$. 
\item Let $\hat h$ denote an empirical transition matrix drawn by obtainin $N_{Y,\ell}$ samples for each row of $h^*$.  
\end{itemize}

Now fix any $\kappa>0$.  Let
\begin{align}
\hat q = \argmax_q \left(\sum_\ell N_{Y,\ell}\sum_{y}\hat h(y|\ell)\log\left(\sum_{x}\hat p(x|\ell)q(y|x)\right) + \kappa \sum_{xy} \log q(y|x)\right)\label{eq:mindef}
\end{align}
and $\hat \Theta=\{q:\ \hat p\hat q=\hat pq\} \cap T_{|\Omega_X|,|\Omega_Y|}$.  
\vspace{.1in}
\begin{thm}
If $N_{X,\ell},N_{Y,\ell}\rightarrow \infty$ in such a way that $N_{Y,\ell'}/\sum_{\ell}N_{Y,\ell} \geq \rho>0$ for each $\ell'$, then $\inf_{q\in \hat\Theta} \UN{q^*-q}\rightarrow 0$ in probability.
\end{thm}

\begin{proof}
It is well-known that $\hat p\rightarrow p^*$ in probability (in both the uniform or the Euclidean norm, which are of course equivalent in this case).  It is easy to see that the same goes for $\hat p\hat q\rightarrow h^*$ (see Lemma \ref{lem:easyconsistency}).  Thus, intuitively, the difficulty is this: by allowing ourselves to ensure $\UN{\hat p - p^*},\EN{\hat p - p^*},\UN{\hat p\hat q - h^*},\EN{\hat p\hat q - h^*}$ sufficiently small, can we find some $\tilde q\in \hat \Theta$ so that $\UN{\tilde q - q^*}$ is arbitrarily small?  It turns out we can.

Recall that $c>0$ is the smallest value of $q^*_{xy}$.  Fix any $\epsilon<c, p^*,q^*$.  Let the right inverse of a matrix be defined by $a^\dagger \triangleq a^T (aa^T)^{-1}$.  Note that since $p^*$ has linearly independent rows, this is well-defined and continuous in a small neighborhood around $p^*$.  Let $M=\EN{(p^*)^\dagger}$.  Find $\delta$ small enough so that if $\UN{p-p^*}<\delta$ then $\EN{p^\dagger}<2M$.  Taking a further smaller $\delta$ if necessary, ensure that if $\UN{p^*-p}<\delta$ then $\EN{p^*-p}$ is less than $\epsilon / 4M\sqrt{|\Omega_X||\Omega_Y|}$.  
Now fix any $\hat p,\hat q$ with $\UN{\hat p-p^*}<\delta$ and $\EN{\hat p \hat q - p^*q^*}<\epsilon/4M$.  Take
\[
\tilde q = q^* + \hat p^\dagger \hat p (\hat q - q^*)
\] 
Then we make the following observations:
\begin{itemize}
  \item Let us compute $\EN{\tilde q -q^*}$. We have
   \begin{align*}
   \EN{\tilde q -q^*} &= \EN{\hat p^\dagger \hat p (\hat q - q^*)} \leq 2M \EN{\hat p \hat q - \hat p q^*} \\
        &\leq 2M \EN{\hat p \hat q - p^* q^*} + 2M\EN{(p^* - \hat p)q^*}\\
        &\leq 2M \frac{\epsilon}{4M} + \frac{2M\epsilon}{4M\sqrt{|\Omega_X||\Omega_Y|}}\sqrt{|\Omega_X||\Omega_Y|}\UN{q^*} \leq \epsilon 
   \end{align*}
  \item $\hat p \tilde q = \hat p q^* + \hat p \hat q - \hat p q^* = \hat p \hat q$
  \item The rows of $\tilde q$ sum to 1.  This is easy to see, because the rows of $q^*$ sum to 1 and the rows of $\hat q$ sum to 1, and so $\tilde q \mathbf{1} = q^* \mathbf{1} + \hat p^\dagger \hat p (\hat q - q^*) \mathbf{1} = \mathbf{1} + 0$ as desired.
  \item The entries of $\tilde q$ are positive.   Indeed, the the smallest value of $q^*$ is $c$, and we have already argued that $\UN{\tilde q - q^*}\leq \epsilon$.  Thus the smallest value of $\tilde q$ is at least $c-\epsilon$, and we have required $\epsilon<c$.  
\end{itemize}
Thus $\UN{\tilde q - q^*}<\epsilon$ and $\tilde q \in \hat \Theta$.

In conclusion, we see that by taking $\hat p$ sufficiently close to $p^*$ and $\hat p \hat q$ sufficiently close to $p^*q^*$, we can ensure that the set $\hat \Theta$ contains a close which is arbitrarily close to the true $q^*$.  Since $\hat p$ and $\hat p \hat q$ are themselves consistent estimators, this completes the proof.
\end{proof}

\begin{lem}
\label{lem:easyconsistency} If $N_{X,\ell},N_{Y,\ell}\rightarrow \infty$ in such a way that $N_{Y,\ell'}/\sum_{\ell}N_{Y,\ell} \geq \rho>0$ for each $\ell'$, then $\UN{p^*q^*-\hat p\hat q},\EN{p^*q^*-\hat p\hat q}\rightarrow 0$ in probability.
\end{lem}
\begin{proof}
Our first task is to make a short study of the continuity of KL divergences on categorical distributions when the probabilities are bounded away from zero.  Recall that we have insisted $q^*_{xy}\geq c>0$ for every $x,y$ -- and this also means $(pq^*)_{\ell y}$ for every $\ell y$, since each row of $p$ is itself a probability distribution.  Moreover, observe that the KL-divergence on $|\Omega_Y|$-dimensional distributions, $\kldiv{\hat r}{\tilde r} \triangleq \sum_{y} \hat r_y \log \hat r_y / \tilde r_y$, is \emph{uniformly} continuous on the space of such distributions whose minimum probability is greater than any fixed positive constant.   It follows that the map $h,p,q \mapsto D(h_\ell||(pq)_\ell)$ is also uniformly continuous on a space where $h$ and $q$ are strictly greater than some fixed positive constant.

With this in hand, the remainder of the proof follows naturally, using the well-known results that $\hat p \rightarrow p^*$ and $\hat h \rightarrow p^*q^*$ in probability.

Fix any $\epsilon,\pi$.  Let $\delta$ the modulus of continuity in the norm $\UN{\cdot}$ at level $\epsilon\rho$ for $h,p,q \mapsto \kldiv{h_\ell}{(pq)_\ell}$ when $h,q>c/2$.  Select $N$ large enough so that $\frac{1}{N_{Y,\ell}} \kappa |\Omega_X||\Omega_Y| \log \frac{1}{c} < \epsilon$ for each $\ell$ and so that with probability at least $\pi$ we have that $\hat h,\hat p$ so that $\UN{\hat h-p^*q^*},\UN{\hat p-p^*}\leq \delta,\UN{\hat h - p^*q^*}<c/2$. Then 
%
\begin{align*}
|\kldiv{\hat h_\ell}{(\hat p \hat q)_\ell} - \kldiv{h^*_\ell}{(\hat p \hat q)_\ell}| \leq \rho \epsilon \\
\kldiv{\hat h_\ell}{(\hat p q^*)_\ell} = |\kldiv{\hat h_\ell}{(\hat p q^*)_\ell} - \kldiv{(p^*q^*)_\ell}{(p^* q^*)_\ell}| \leq \rho\epsilon 
\end{align*}
%
Now, since $\hat q$ is defined as the maximizer of a certain quantity (Equation \ref{eq:mindef}), we may be sure that it is greater than the same quantity evaluated at $q=q^*$.  That is,
%
\begin{align*}
0 \leq& \sum_\ell N_{Y,\ell}\sum_{y}\hat h(y|\ell)\log\frac{\sum_{x}\hat p(x|\ell)\hat q(y|x)}{\sum_{x}\hat p(x|\ell)q^*(y|x)} + \kappa \sum_{xy} \log \frac{\hat q(y|x)}{q^*(y|x)}\\
 =& \sum_\ell N_{Y,\ell}\left(\kldiv{\hat h_\ell}{(\hat pq^*)_\ell} - \kldiv{\hat h_\ell}{(\hat p\hat q)_\ell}\right) + \kappa \sum_{xy} \log \frac{\hat q(y|x)}{q^*(y|x))}
\end{align*}
%
Applying our continuity results, it follows that
%
\begin{gather*}
 \sum_\ell N_{Y,\ell}\left(\kldiv{\hat h^*_\ell}{(\hat p\hat q)_\ell}\right)  \leq 2\left(\sum_\ell N_{Y,\ell}\right)\epsilon  + \kappa |\Omega_X||\Omega_Y| \log \frac{1}{c} 
\end{gather*}
%
Note that the left-hand summands are all positive.  So, in particular, it follows that for each $\ell'$, applying the uniformity condition $\rho$, we have
%
\begin{gather*}
\kldiv{\hat h^*_{\ell'}}{(\hat p\hat q)_{\ell'}}  \leq 2\epsilon  + \frac{1}{N_{Y,\ell'}} \kappa |\Omega_X||\Omega_Y| \log \frac{1}{c} \leq 3\epsilon
\end{gather*}

That is, we have shown convergence of probability in the KL sense: for any $\epsilon,\pi$ we can find $N$ high enough so that $\kldiv{\hat h^*_{\ell'}}{(\hat p\hat q)_{\ell'}}<\epsilon$ with at least probability $\pi$.  This, in turn yields convergence in probability in the Euclidean or uniform metrics by Pinsker's inequality. 
\end{proof}

\end{document}

### 2018-04-02 generativemodelsforscrna/outline.tex
corrected bug

@@
 \includegraphics[width=0.8\textwidth]{pics/Z1gn}
-\caption{\label{fig:Z1gn}}
+\caption{Colors indicate gene expression level. \label{fig:Z1gn}}
 \end{figure}
@@
 \includegraphics[width=0.8\textwidth]{pics/resids}
-\caption{\label{fig:resids}}
+\caption{A $p$-value residual plot.\label{fig:resids}}
 \end{figure}
@@
 If we restrict ourselves to those genes whose average probability of going off is less than 20\%, we obtain Figure \ref{fig:residslesszeros}.
+We do this in Figure \ref{fig:residslesszerosFA}.
 
@@
 \includegraphics[width=0.8\textwidth]{pics/residslesszeros}
-\caption{\label{fig:residslesszeros}}
+\caption{If we include only genes that express at least 80\% of the time, the $p$-value residual plot looks less structured.\label{fig:residslesszeros}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/residslesszerosFA}
+\caption{Using Factor Analysis for our model (with 16 factors), we see more significant underfitting.  (Here we include only genes that express at least 80\% of the time)\label{fig:residslesszerosFA}}
 \end{figure}

### 2018-04-04 trygve/paper.tex
more evaluations for mlm

@@
 
-\title{Identifying the unidentifiable: combining knowledge from destructive measurements}
+\title{Markov Link Method for combining knowledge from destructive measurements}
 \author{Jackson Loper}
@@
 We call this ``joint measurement.''  What can we do if joint measurement is unavailable?
-In this paper we investigate a simple assumption that can make it possible do obtain joint distributions even when joint measurement isn't possible.
+In this paper we propose the Markov Link Method: a simple assumption that can make it possible to obtain joint distributions linking tools together, even when joint measurement isn't possible.
 \end{abstract}
@@
 
-In this work we show that it is indeed sometimes possible to get useful and statistically rigorous bounds for these kinds of questions, even when joint measurements are unavailable.
+In this work we propose the Markov Link Method (MLM) as a way to get useful and statistically rigorous bounds on the relationship between two measurement tools even when joint measurements are unavailable, hopefully yielding insight for the kinds of difficulties illustrated above.
 In particular, we consider the case that we can obtain samples from a variety of subpopulations.
@@
 
-We apply our method to single cell RNA data.
-We show how different clusterings from different single cell RNA modalities can be connected.
+We apply MLM to the inferred cell-types of single neurons, as determined by two different methods for single-cell RNA sequencing.
+Technique I is known to have high fidelity, whereas technique II is less accurate.
+However, technique II also carries additional electrophysiological information which we would later like to use for other purposes.
+We show how the cell-type inferences from the two techniques can be connected.
 Our method yields results which are generally consistent with the scientist's understanding of the data and clusters, but also reveal potential gaps which could be important for further study.
@@
 \hfill{}\includegraphics[width=.3\textwidth]{pics/plate}\hfill{}
-\caption{We assume that we have two datasets, both governed by the same process.
+\caption{We assume that we have two datasets, drawn from different techniques applied to specimens from various subpopulations.
 In the top dataset, the results of technique II, denoted by $Y_i$, are hidden from us (we indicate this by making the circle gray).
@@
 Using observations of $X_i$ we can certainly learn the parameters $p$ which govern the relationship between $X_i$ and the subpopulation, $\ell_i$.
-However, if we believe that the relationship between $X$ and $Y$ is the same regardless of the population value $\ell$, we can also learn something about the parameters $q$ which govern the relationship between $X$ and $Y$.\label{fig:plate}}
+Likewise we can determine the relationship between $Y_i$ and $\ell_i$.
+However, if we believe that the relationship between $X$ and $Y$ is the same regardless of the population value $\ell$, we can also learn something about the parameters $q$ which govern the relationship \emph{between} $X$ and $Y$.\label{fig:plate}}
 \end{figure}
@@
 
-Our main assumption is that
+Our main assumption is a kind of Markov assumption, from which we obtain the name ``Markov Link Method'':
 
@@
 If this seems unlikely, it may make sense to collapse similar values of $\ell$ together.
-This assumption can be articulated by the plate diagram found in Figure \ref{fig:plate}.
+The Markov assumption can be visualized with the plate diagram found in Figure \ref{fig:plate}.
 
@@
 More generally, if two techniques have something that they both measure, we can restrict attention to that one common phenomenon.}  If the answer to all of these questions is yes, the assumption that $(X_i,Y_i)$ is drawn from $p(x_i|\ell_i)p(y_i|x_i)$ may apply.
+If some answer is no the method may still apply, but certain corrections may be necessary; we discuss these in our conclusions.
 
-We can now articulate our goal.
-Informally, our goal is to see what we can understand about the relationship between the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$, i.e.
+Under these key assumptions, the Markov Link Method gives us a way to uncover something about the relationship between the observations $X_i$ from one measurement technique and the observations $Y_i$ from the other measurement technique, i.e.
 the distribution $q^*$.
-Unfortunately, given the limitations of the data, this may not be possible.
-Indeed, it is easy to see that $q$ is dramatically unidentifiable in many cases.
+Unfortunately, given the limitations of the data, $q^*$ itself may be dramatically unidentifiable in many cases.
+Instead, the Markov Link Method yields a \emph{set} of possible values of $q$ which are consistent with the data.
+It can then be shown that the true value $q^*$ lies close to this set with high probability.
 
-For example, consider the case that $\ell$ lies in the set ${1,2}$, $X$ lies in the set ${1,\cdots 100}$ and $Y$ lies in the set ${1,2}$.
+To understand the limitations of the data more clearly, consider the case that $\ell$ lies in the set $\{1,2\}$, $X$ lies in the set $\{1,\cdots 100\}$ and $Y$ lies in the set $\{1,2\}$.
 As long as $m,n$ are large enough, we can effectively estimate $p^*$ and $h^*(Y|\ell)=\sum_x p^*(x|\ell)q(Y|x)$.
 However, this does not tell us very much about what $q$ might be.
-Indeed, $q$ lies in a 100-dimensional space, and the restriction that $q$ must satisfy $h^*(Y|\ell)=\sum_x p^*(x|\ell)q(Y|x)$ for each $\ell,Y$ actually only introduces 2 new constraints.
+Indeed, $q$ lies in a 100-dimensional space, and the restriction that $q$ must satisfy $h^*(Y|\ell)=\sum_x p^*(x|\ell)q(Y|x)$ for each $(\ell,Y)$ actually only introduces 2 new constraints.
 Thus there is a 98-dimensional space that $q$ may lie in, and we simply cannot know where in that space $q$ may lie.
-However, in practice, we find that the \emph{inequality constraints}, namely that $q(y|x)>0$ for every $x,y$, can actually force this apparently vast space to be tightly centered around a single point.
+However, in practice, we find that the \emph{inequality constraints}, namely that $q(y|x)\geq0$ for every $(x,y)$, can actually force this potentially vast space to be tightly centered around a single point.
 
-This suggests a somewhat less amibtious goal: to understand the \emph{set of values} of $q$ which are consistent with the data.  Specifically, we will seek to produce an estimator for the set
+In formal terms, the Markov Link Method produces a estimator for the set of possible values of $q$ which are consistent with what is observable, i.e.
 
@@
 
+Let us see now how MLM achieves this.
+
 \section{Algorithm}
 
-We begin by estimating
+The Markov Link Method begins by estimating
 
@@
 We take
-
+%
 \begin{equation*}
@@
 \end{equation*}
-
-Note in the second term we apply an entropic penalty to $q$.
-This is both necessary to ensure that $\hat q$ is uniquely defined in terms of $\hat q,\hat h$, as well as a useful regularization.
-This regularization pushes us slightly towards uniform distributions, which makes sense if we believe that $q(y|x)>0$  for each $x,y$.
-Other regularizations could also be used and would potentially enjoy similar theoretical guarantees to the ones we will show below.
+%
+In practice, we took the entropic regularizer $\kappa$ to be $0.01$; it is mostly useful for ensuring numerical stability of the algorithm.
 
@@
 \begin{equation*}
-\hat \Theta\triangleq \left\{q:\ \sum_y p(x|\ell)q(y|x) = \sum_y p(x|\ell)\hat q(y|x)\ \forall \ell,y\right\}
+\hat \Theta\triangleq \left\{q:\ \sum_y \hat p(x|\ell)q(y|x) = \sum_y \hat p(x|\ell)\hat q(y|x)\ \forall \ell,y\right\}
 \end{equation*}
@@
 
-Our motivation for this problem arose from looking at Allen Institute cell-type assignment of cells, performed using two different modalities.
-The only thing connecting the two modalities was that each modality had a variety of samples drawn from a variety of sub-populations, and the same sub-populations were used for both modalities.
-Thus our input was two tables: (technique I cluster $\times$ sub-population), and (technique II cluster $\times$ sub-population) (Figure \ref{fig:alleninput}).
-Our output was a polytope $\hat \Theta$ of possible ways that technique I clusters could line up against the technique II clusters (Figure \ref{fig:allenoutput}).
-
-Empirically, we see that this polytope appears to be very small, and tightly centered around our estimate for $\hat q$.
-Thus $\hat q$ may indeed serve as a useful indication of how the different technique's clusters line up.
-There are certain aspects of $q$, however, which are more uncertain than others.
-We were able to measure this by sampling uniformly from the polytope $\hat \Theta$ and seeing which entries varied the most (cf.
-Appendix \ref{sec:dikin}).
-Detecting these entries is useful because it suggests what further experiments would be helpful to more fully clarify the relationship between the two modalities.
-
 \begin{figure}
@@
 Technique II gives a less complete analysis, but yields additional electrophysiological data that may be of interest.
-The results of technique I was used to assign it to one of 116 categories, and the results of technique II was used to assign it to one of 14 categories.
-The tables above show the distribution of category assignment for each selection technique.
+The results of technique I was used to assign cells to one of 116 categories, and the results of technique II was used to assign cells to one of 14 categories.
+The tables above show the distribution of category assignment for each sub-population selection technique.
 The goal is to use these tables to be able to say something about how the 14 categories from Technique II might line up with the 116 categories from technique I.
@@
 \end{minipage}
-
 \end{figure}
 
-\begin{figure}
-\includegraphics[width=0.8\textwidth]{pics/allenavg}
-\caption{\label{fig:allenoutput}In the same setup as Figure \ref{fig:alleninput}.
-Using our method we produce a set of possible $q$s which are consistent with the data tables given.
-Using our method we first produce a single table, $\hat q$, that suggests how the outputs of the two techniques might line up.
-We are able to do this even though we were never able to observe a cell processed by both techniques.
-We emphasize, however, that this table is not guaranteed to be a consistent estimator for the true value of $q^*$.
-However, we can guarantee that asymptotically the true $q^*$ will lie very close to a certain set $\hat \Theta$, containing $\hat q$.
-Fortunately, this set appears to be very small indeed, and tightly centered around $\hat q$.
-To see this, we used a Dikin sampler to sample uniformly from $\hat \Theta$.
-We saw that almost every sample was very close to $\hat q$.
-The second plot shows an example of such a sample.
-The third plot is an average value of each entry of $q$ over many samples.
-The final plot indicates the standard deviation of each entry over the samples.
-Where this is high we have some substantial ambiguity, and so it may be important to perform further experiments to narrow down the correspondence.
-\label{fig:allenavg}}
-\end{figure}
+Our motivation for this problem arose from looking at Allen Institute cell-type assignment of cells, performed using two different experimental techniques (also called experimental ``modalities'').
+The only thing connecting the two modalities was that each modality had a variety of samples drawn from a variety of sub-populations, and the same sub-populations were used for both modalities.
+Thus our input was two tables: (technique I cell-type $\times$ sub-population), and (technique II cell-type $\times$ sub-population) (Figure \ref{fig:alleninput}).
+A key question immediately appears from these tables: does the way in which we have clustered cells into different types under technique I line up nicely with the way in which we have clustered cells under technique II?
+The Markov Link Method gives us a way of addressing this question that is completely agnostic about the underlying genetic markers that were used to produce the clusters.
+Our output was a set $\hat \Theta$ of possible ways that technique I cell-types could line up against the technique II cell-types.
+That is, each element of $\hat \Theta$ was itself a table of the form (technique I cell-type $\times$ technique II cell-type).
 
-Another way to look at $\hat \Theta$ is to try to understand its scope of the polytope along various directions.
-In this effort, we selected a direction uniformly at random and found the two extremal vertices in this direction.
+\subsection{Making sense of Markov Link Method output}
+
+The Markov Link Method yields a set (indeed, a convex polytope).
+Each element of this set gives a particular way that technique I might be related to technique II.
+It is not immediately easy to understand how to use or visualize the polytope $\hat \Theta$ of possible values for $q$.
+Ultimately we want to know something about the correspondence between the techniques, i.e.
+$q^*$ -- how can we make good use of a \emph{set} of possible corresponences?
+Here we suggest several main tools for analysis:
+
+\begin{itemize}
+    \item Rotationally Uniform eXtremal samples (RUX).
+Every point in the polytope is a convex combination of its vertices, so by understanding the extremal vertices of the polytope we understand much about what values are possible.
+Unfortunatelly, there are a truly vast number of vertices, and looking at all of them is quite impossible.
+Instead, we propose a method for drawing samples which are illustrative of the overall character of the polytope: Rotationally Uniform eXtremal samples.
+These samples are drawn simply by selecting a random direction in the polytope, and finding the point in the polytope which is furthest along that direction.
+With probability 1 this point is unique, and provides a useful notion of the polytope's extent.
+Obtaining such a sample simply requires solving a linear programming problem.
+We show 10 such samples in Figure \ref{fig:allenRUX}.
+
+    Each sample shows a possible way that the two different techniques might correspond; together, these samples reveal both what the data says about the correspondence and what we still need to know.
+In particular, some entries of correspondence are nearly the same for every sample, and others can vary more significantly.
+This suggests what possible experiments might be important to do next.
+Since each row always sums to one, any ambiguity must lie in trading off between different columns within a row.
+To resolve this ambiguity, we could find a subpopulation which we believe might include, say, half of those ambiguous columns, and then perform additional experiments in both modalities, drawing from that subpopulation.
+
+    \item Diameter estimation.
+One problem with RUX samples is that there may be some very special direction in which the polytope has tremendous width.
+That is, simply looking at ten or twenty RUX samples or even taking standard deviations over thousands of RUX samples, it may appear that the polytope is quite small.
+Yet, in fact, it may be that there is a very special direction such that if we travel along that direction we find a value of $q$ which is radically different from the typical RUX samples.
+To assess this, we attempt to show the overall diameter of the polytope is fairly limited, thus convincing ourselves that this is not a problem.
+In this effort, we obtained Paired Rotationally Uniform eXtremal samples (PRUX): we first choose a direction uniformly at random and then found the two extremal vertices on the polytope in that direction and in the opposite direction.
 We then computed the distance between those vertices, as well as the distance between those vertices projected to the direction.
 Doing this procedure many times and making a scatterplot yields Figure \ref{fig:allenwidths}.
-This figure suggests that the polytope may be only about 8 units in diameter, although we caution that general results on polytope diameter estimation are not encouraging, cf.
-\citep{brieden1998approximation}.
-However, considering that the magnitude of this diameter would be distributed over 1624 entries of the matrix, each of which lies in $[0,1]$, it would seem that the polytope may indeed be constrained to be very close to the output we showed in Figure \ref{fig:allenoutput}.
+This figure suggests that the polytope may be only about 8 units in diameter (although we caution that general results on polytope diameter estimation are not encouraging, cf.
+\citep{brieden1998approximation}).
+Considering that the magnitude of this diameter would be distributed over 1624 entries of the matrix, each of which lies in $[0,1]$, it would seem that the polytope may indeed be a fairly small set.
+Although there is certainly no guarantee that there isn't some very large-magnitude direction hiding within this polytope, it is encouraging that we do not find any extreme outliers in this figure.
+
+    \item Random uniform samples.
+The maximum entropy principle provides a quite different point of view: given that we can never distinguish where $q$ may lie inside the polytope $\hat \Theta$, we could treat all such values of $q$ as equally likely.
+In this vein, we apply a Dikin ellipsoid sampler to obtain samples drawn approximately uniformly from $\hat \Theta$.
+We also look at the standard deviation of each entry of $q$ across all these samples, which gives us another way to assess where the polytope has significant ambiguity.
+The results may be seen in \ref{fig:allenoutput}.
+\end{itemize}
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenRUX}
+\caption{10 Rotationally Uniform eXtremal samples from the polytope $\hat \Theta$.
+These extremal samples from $\hat \Theta$ have quite a lot in common.
+The common features are ones that we can feel confident are present in the true value of the joint distribution, $q^*$.
+Other entries have more variability; these suggest specific experiments which might help disambiguate how technique I and technique II are related through $q^*$.
+\label{fig:allenRUX}}
+\end{figure}
 
@@
 Each dot above corresponds to a randomly selected direction.
-The horizontal position of the dot indicates the width of the polytope along that direction, and the vertical position indicates the distance between the two extremal vertices along that direction.\label{fig:allenwidths}}
+The horizontal position of the dot indicates the width of the polytope along that direction, and the vertical position indicates the distance between the two extremal vertices along that direction.
+It appears that the overall diameter of the set may be less than 8 units.\label{fig:allenwidths}}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenavg}
+\caption{\label{fig:allenoutput}The first row shows the the MLM estimate of $\hat q$ (which is not a consistent estimator for $q^*$).
+The second row shows a sample drawn approximately uniform from $\hat \Theta$, obtained by 10000 iterations of the Dikin sampler.
+The third row shows the center of mass of $\hat \Theta$ and the final row shows the standard deviation of each entry (as estimated by the Dikin samples).
+Notice that these deviations are never more than 6\%, suggesting that the polytope $\hat \Theta$ is closely centered around its center of mass.\label{fig:allenavg}}
+\end{figure}
+
+Together, these approaches give us practical ways of making real use of the results of the Markov Link Method.
+More generally, they may be applied to many situations where a parameter of interest simply cannot be exactly identified but can be shown to lie near to some set.
+
+\subsection{Evaluation}
+
+It is difficult to evaluate the accuracy of the Markov Link Method, because ground-truth data is not available in the cases where it actually matters.
+Nonetheless, we can make a variety of attempts to see how the method fares.
+
+Parametric bootstrap provides one natural way to assess our confidence.
+That is, we simulate a surrogate dataset according to the estimated distribution (holding the number of samples for each group fixed), apply our method to the resulting surrogate dataset, and then obtain an estimate $\hat \Theta^{(i)}$.
+We do this 100 times.
+For each estimate, we can then measure whether the original $\hat q$ was is close to the estimated polytope, i.e.
+for each $i$ we can look at the Euclidean norm between $\hat q$ and the set $\hat \Theta^{(i)}$.
+Doing so for 100 bootstrap samples and taking a histogram of these distances yield Figure \ref{fig:bootstrap}.
+The results suggest our estimator has a fairly tight confidence around the true polytope $\Theta^*$ for this dataset.
+However, we note that theoretical properties of this are certainly lacking.
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenbootstrap}
+\caption{Bootstrap gives us some measure of confidence in the MLM estimate.
+None of the 100 bootstrap sample polytopes measured more than 2 units from from the original point estimate, $\hat q$.
+That is, for each sample polytope we could find some point in the polytope that was less than 2 units from $\hat q$.
+Spread out over the 1624 entries of the matrix $q$, each of which lies in $[0,1]$, we are encouraged insofar as 2 units does not seem to be a very great distance.
+\label{fig:bootstrap}}
+\end{figure}
+
+Nonparametric bootstrap is another approach.
+In this case, we simply sample the empirical distributions $\hat p,\hat h$ with replacement to obtain our surrogate datasets, and then apply the MLM method to each surrogate dataset.
+We evaluate the degree of variability amongst the MLM polytope estimates by looking at the extremal points of those polytopes.
+That is, we select a random direction and find the extremal vertex for each MLM estimate in that direction.
+We can then look at position of these extremal points projected along that direction.
+Looking over all the surrogates and subtracting the mean, we obtain a histogram of deviations.
+We can then apply this to various directions, which yields the variety of histograms seen in Figure \ref{fig:allenbootstrap}.
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenbootstrapnonparametric}
+\caption{Nonparametric bootstrap gives us another form of confidence.
+For each of nine randomly selected directions and for each bootstrap sample polytope, we compute how far the polytope extends in that direction.
+For each direction we can then plot a histogram of how far the various polytopes extend.
+We see that they do not vary dramatically between samples, which gives us confidence that our estimate is somewhat close to the truth.\label{fig:allenbootstrap}}
+\end{figure}
+
+Finally, we can look directly at whether MLM helps us model the data $Y$.
+In theory it seems possible that by pooling data between the modalities we could gain statistical power.
+This is particularly true in the case of the particular dataset we investigated, wherein obtaining samples for technique II is quite expensive and so there are far fewer available datapoints for the $Y$ data.
+Thus the empirical distribution of $Y|\ell$ (namely $\hat h$) may in fact be a very poor estimator.
+This suggests that perhaps we might consider the estimator $\tilde h(y|\ell)\triangleq \sum_x \hat p(x|\ell) \hat q(y|x)$ instead,  since it allows us to leverage knowledge gained from the plentiful samples of technique I.
+To test this estimator, we held out 10\% of the $Y$ data.
+We then applied the MLM method to obtain estimates $\hat p,\hat q$ and use those to produce $\tilde h$ as an estimator for the distributions of $Y|\ell$.
+We then compare $\tilde h$ and $\hat h$ by calculating the log-likelihood of each estimator on the held-out data.
+We found an average value of -1.1 per entry for the MLM estimator $\tilde h$, but the empirical distribution $\hat h$ yields a held-out log likelihood of $-\infty$ because there are examples in the testing set that are simply never seen in the training set.
+Even if we use pseudocounts to smooth the empirical distribution, we never obtain likelihoods as good as we get using the model which shared data between the two experimental modalities (See Figure \ref{fig:heldout}).
+Thus we see that sharing data between the modalities helps can help us estimate each modality more accurately, as well as enabling us to understand more about the connection between the modalities.
+
+\begin{figure}
+\includegraphics[width=0.8\textwidth]{pics/allenheldout}
+\caption{Here we compare two estimators of the distributions of $Y|\ell$.
+In the orange case, we use the Markov Link Method.
+In the blue case, we use the empirical distributions of $Y|\ell$, smoothed using pseudocounts.
+The horizontal axis indicates the size of these pseudocounts in log-scale.
+The MLM method performs better than the empirical distribution at any level of smoothing.
+\label{fig:heldout}}
 \end{figure}

### 2018-04-04 tryg/batteries.py

@@ -128,7 +128,10 @@ class TabCompleteGraph:
                                                       
 '''
 
-def batch_normalize_linear(name,momentsource,target_mean,target_stdev):
+def normalize_dense_affine(name,momentsource,target_mean,target_stdev):
+    '''
+    normalize a dense_affine layer based on data
+    '''
     GR=tf.get_default_graph()
     momentsource_tensor=GR.get_tensor_by_name(momentsource+":0")
     
@@ -146,22 +149,6 @@ def batch_normalize_linear(name,momentsource,target_mean,target_stdev):
 
     return assig1,assig2
 
-def set_limits_convex_combination(name,target,target_low,target_high):
-    GR=tf.get_default_graph()
-    
-    with tf.variable_scope(name):
-        bias1=GR.get_tensor_by_name(target+'_layer/bias1:0')
-        bias2=GR.get_tensor_by_name(target+'_layer/bias2:0')
-    
-        assig1 = tf.assign(bias1,target_low,name='rebias1')
-        
-        with tf.control_dependencies([assig1]):
-            assig2 = tf.assign(bias2,target_high,name='rebias2')
-
-        assig = tf.identity(assig2,name='batchnorm')
-
-    return assig
-
 
 '''
  _                           
@@ -174,6 +161,9 @@ def set_limits_convex_combination(name,target,target_low,target_high):
 '''
 
 def translate_initializer(init,dtype):
+    '''
+    Conventions for initializers
+    '''
     if isinstance(init,float) or isinstance(init,int):
         return tf.constant_initializer(init)
     elif isinstance(init,tuple):
@@ -183,51 +173,45 @@ def translate_initializer(init,dtype):
     else:
         return init
 
-
-
-def linear(name,inp,outsize,dtype,kernel_initializer='glorot',bias_initializer=0):
+def dense_affine(name,inp,outsize,dtype,kernel_initializer='glorot',bias_initializer=0,bias=True):
     with tf.variable_scope(name+"_layer"):
         inpsize=inp.shape[1]
-        bias_initializer=translate_initializer(bias_initializer,dtype)
+        if bias:
[…]

### 2018-04-04 trygve/paper.tex
changed caption

@@
 \includegraphics[width=0.8\textwidth]{pics/allenRUX}
-\caption{10 Rotationally Uniform eXtremal samples from the polytope $\hat \Theta$.
-These extremal samples from $\hat \Theta$ have quite a lot in common.
+\caption{Output: 10 Rotationally Uniform eXtremal samples from the set $\hat \Theta$.
+Each sample shows a different way that technique I might line up with technique II; all of them are consistent with the observed data.
+Note that these extremal samples from $\hat \Theta$ have quite a lot in common.
 The common features are ones that we can feel confident are present in the true value of the joint distribution, $q^*$.

### 2018-04-05 trygve/paper.tex
latest and greatest

@@
 we have samples $(\ell_1,X_1) \cdots (\ell_1,X_n)$ such that $\mathbb{P}(X_i=x) = p^*_{\ell_i,x}$, $N_{X,\ell}=\sum_{i=1}^n \mathbb{I}_{\ell_i=\ell}$, and $\hat p_{\ell x} = \sum_{i=1}^n \mathbb{I}_{X_i=x,\ell_i=\ell} / N_{X,\ell}$.
-\item Let $\hat h$ denote an empirical transition matrix drawn by obtainin $N_{Y,\ell}$ samples for each row of $h^*$.  
+\item Let $\hat h$ denote an empirical transition matrix drawn by obtainin $N_{Y,\ell}$ samples for each row of $h^*=p^*q^*$.  
 \end{itemize}
@@
 Our first task is to make a short study of the continuity of KL divergences on categorical distributions when the probabilities are bounded away from zero.
-Recall that we have insisted $q^*_{xy}\geq c>0$ for every $x,y$ -- and this also means $(pq^*)_{\ell y}$ for every $\ell y$, since each row of $p$ is itself a probability distribution.
+Recall that we have insisted $q^*_{xy}\geq c>0$ for every $x,y$ -- and this also means $(pq^*)_{\ell y}\geq c$ for every $\ell,y$, since each row of $p$ is itself a probability distribution.
 Moreover, observe that the KL-divergence on $|\Omega_Y|$-dimensional distributions, $\kldiv{\hat r}{\tilde r} \triangleq \sum_{y} \hat r_y \log \hat r_y / \tilde r_y$, is \emph{uniformly} continuous on the space of such distributions whose minimum probability is greater than any fixed positive constant.
@@
 
-With this in hand, the remainder of the proof follows naturally, using the well-known results that $\hat p \rightarrow p^*$ and $\hat h \rightarrow p^*q^*$ in probability.
+With this in hand, the remainder of the proof follows naturally, using the well-known results that empirical distributions are consistent, i.e.\ $\hat p \rightarrow p^*$ and $\hat h \rightarrow p^*q^*$ in probability.
 
 Fix any $\epsilon,\pi$.
-Let $\delta$ the modulus of continuity in the norm $\UN{\cdot}$ at level $\epsilon\rho$ for $h,p,q \mapsto \kldiv{h_\ell}{(pq)_\ell}$ when $h,q>c/2$.
+Let $\delta$ the modulus of continuity in the norm $\UN{\cdot}$ at level $\epsilon\rho$ for the map $h,p,q \mapsto \kldiv{h_\ell}{(pq)_\ell}$ restricted to the domain where $h,q>c/2$.
 Select $N$ large enough so that $\frac{1}{N_{Y,\ell}} \kappa |\Omega_X||\Omega_Y| \log \frac{1}{c} < \epsilon$ for each $\ell$ and so that with probability at least $\pi$ we have that $\hat h,\hat p$ so that $\UN{\hat h-p^*q^*},\UN{\hat p-p^*}\leq \delta,\UN{\hat h - p^*q^*}<c/2$.
-Then
+Then, with probability $\pi$, we must have
 %

### 2018-04-05 tryg/batteries.py

@@ -149,6 +149,27 @@ def normalize_dense_affine(name,momentsource,target_mean,target_stdev):
 
     return assig1,assig2
 
+def normalize_dense_affine_somewhat(name,momentsource,target_mean,target_stdev,alpha):
+    '''
+    normalize a dense_affine layer based on data
+    '''
+    GR=tf.get_default_graph()
+    momentsource_tensor=GR.get_tensor_by_name(momentsource+":0")
+    
+    with tf.variable_scope(name):
+        wts=GR.get_tensor_by_name(momentsource+'_layer/kernel:0')
+        bias=GR.get_tensor_by_name(momentsource+'_layer/bias:0')
+    
+        layshape1 = wts.shape[1]
+        if layshape1.value is None:
+            layshape1=tf.shape(wts)[1]
+
+        mn,vr=tf.nn.moments(momentsource_tensor,axes=[0],name='moments')
+        assig1 = tf.assign(wts,(1-alpha)*wts + alpha* target_stdev * wts / tf.reshape(tf.sqrt(vr),[-1,layshape1]),name='rekernel')
+        assig2 = tf.assign(bias,(1-alpha)*bias + alpha*(target_mean + bias-mn),name='rebias')
+
+    return assig1,assig2
+
 
 '''
  _                           

### 2018-04-05 tryg/batteries.py
elementwise monotone bias bug

@@ -211,10 +211,11 @@ def dense_affine(name,inp,outsize,dtype,kernel_initializer='glorot',bias_initial
             rez=tf.tensordot(inp,kernel,axes=[[1],[0]])
     return tf.identity(rez,name=name)
 
-def elementwise_monotone(name,inp,dtype,bias_initializer=0,bias=True,n_basis=4,
-        xscale=1,ysoftplusscale=1):
+def elementwise_monotone(name,inp,dtype,bias_initializer=0.0,bias=True,n_basis=4,
+        xscale=1.0,ysoftplusscale=1.0):
     '''
-    Unique parameterization for each guy in the second index
+    Different monotone function for each value of the second index
+    applied uniformly across the first index
     '''
 
     assert len(inp.shape)==2
@@ -224,14 +225,19 @@ def elementwise_monotone(name,inp,dtype,bias_initializer=0,bias=True,n_basis=4,
 
         inp=tf.reshape(inp,(tf.shape(inp)[0],shp,1))
 
-        bias = get_variable2('bias',(1,shp),dtype,bias_initializer)
+        if bias:
+            bias = get_variable2('bias',(1,shp),dtype,bias_initializer)
 
         shift = get_variable2('shift',(1,shp,n_basis),dtype,(-xscale,xscale))
         mult_almost = get_variable2('mult_almost',(1,shp,n_basis),dtype,(-1,1))
         mult = tf.softplus(ysoftplusscale+mult_almost,name='mult')
 
-        final = tf.reduce_sum(mult * tf.atan(inp+offset),axis=2,name='final')
-    return final
+        if bias:
+            final = tf.reduce_sum(mult * tf.atan(inp+offset),axis=2,name='final_before_bias')
+            final = tf.identify(bias+final,name='final')
+        else:
+            final = tf.reduce_sum(mult * tf.atan(inp+offset),axis=2,name='final')
+    return tf.identify(final,name)
 
 
 '''

### 2018-04-20 trygve/paper.tex
New examples for MLM

@@
 
-\title{Markov Link Method for combining knowledge from destructive measurements}
+\title{Markov Link Method for calibrating without joint measurement, including the case of destructive measurements}
 \author{Jackson Loper}
@@
 \begin{abstract}
-When many different measurement tools are applied, how can we best combine information from the different different tools?
-If we can apply multiple measurement tools to the same specimen, we can at least begin to understand how the tools they are related.
-We call this ``joint measurement.''  What can we do if joint measurement is unavailable?
-In this paper we propose the Markov Link Method: a simple assumption that can make it possible to obtain joint distributions linking tools together, even when joint measurement isn't possible.
+A proliferation of new experimental tools has left a serious gap: calibration.
+Two thermometers can be calibrated against each other by simply measuring the same bodies of water with both thermometers, but the problem is much harder for many modern tools.
+One common problem is that we do not have measurements from the same ``body of water'' for both tools.
+We propose the Markov Link Method (MLM) as a way to overcome this difficulty.
+This method produces consistent estimators that tightly bound the calibration, i.e.
+the conditional distribution of one tool's measurement given another tool's measurement.
+It achieves this without any measurement data from both tools applied to the same ``bodies of water.'' Moreover, MLM makes zero assumptions about what calibrations we might expect to see, instead applying a subpopulation-based conditional independence assumption.
+We evaluate MLM on a pair of single-cell RNA techniques, obtaining precise calibrations between the tools as well as more accurate models for each tool separately.
 \end{abstract}
 
-The modern setting is rife with experimental methodologies, and it can be very frustrating to understand how the output of these methods relate to one another.
-This is particularly difficult if joint measurement is unavailable, i.e.\ it is impossible or impractical to observe a single specimen with multiple methodologies.
-For example, after processing a neuron with a given single-cell rna sequencing technique, we cannot generally go back and process that same neuron with a different technique.
-This makes it seemingly impossible to calibrate the techniques against each other.
-By contrast, it is easy to calibrate thermometers against each other, because we can simply measure the same water with two different thermometers and see how the measurements relate.
-Without joint measurement, this isn't possible.
+The modern setting is rife with experimental measurement tools, and it can be very frustrating to understand how the output of these tools relate to one another.
+This problem is known as ``calibration'' or ``zeroing'' \cite{bipm2008international}.
+A calibration tells us what readings we should expect from one tool, given the reading we obtained from another tool.
+Calibration additionally must give uncertainty bounds for how much we can trust those expectations.
+Calibration between measurement tools allows us to combine experimental results from different labs and different methodologies into larger scientific theories.
 
-The absence joint measurement gives rise to a number of problems:
+Formally, a calibration is simply a conditional distribution.
+As input, this conditional distribution takes the measurement results from one tool on a particular specimen.
+As output, it yields a distribution on what a second tool would measure on the same specimen.
+One way to learn the calibration is to measure the same specimens under both tools.
+We call this ``joint measurement.''  Unfortunately, calibrations are often required even when joint measurement is unavailable.
+For example, if the measurement tool alters the specimen being measured, joint measurement is simply impossible.
+In other cases, it may be expensive or impractical.
+
+We here propose the Markov Link Method (MLM) to estimate calibrations between tools.
+The MLM can be trained without any joint measurement.
+The key idea is to use multiple subpopulations of specimens.
+If each subpopulation captures a different slice of the overall population, we can obtain tight bounds on the true calibration.
+This is true even if each subpopulation is highly heterogeneous.
+By integrating information from all the subpopulations we can make rigorous deductions about what the calibration might be.
+MLM also gives suggestions about which further subpopulations might be helpful to study in order to further refine our knowledge of the true calibration.
+The method can also be naturally extended to calibration distributions among many tools.
+
+Some example applications include:
 
 \begin{itemize}
+    \item Quality control for manufacturing.
+The surest way to test the reliability of a part is to construct a machine that pushes the part until it breaks.
+However, how can we test the reliability of the machine that performs the test?
+In each test run there will be some variability induced by the machine itself, which induces a measurement error  In practice, some kind of assumptions about part homogeneity are used to approximate this error (cf.\ \cite{de2005gauge}).
+However, if we have multiple testing machines we can use the MLM to obtain a calibration between the machines, even though we can never test the same part with both machines.
+This enables us to bound the overall measurement error.
 
-\item Different tools may teach us different things, and we need to combine tools
-to learn everything we can.
-Let $A,B$ denote two different quantities of interest about cells (e.g.
-morphology and gene expression).
-Let us say one tool enables us to measure $A$ and another tool to measure $B$.
-We would like
-to be able to say something about what morphologies are associated with what gene expressions.
+    \item Combining knowledge across experimental modalities: morphology and transcriptomics.
+There are different ways to think about the different types of cells in an organism.
+A traditional approach is to classify cells based on what they look like (cf.\ \cite{steinman1973identification,bloomfield1982physiological}).
+A more modern approach is to assay the cell's transcriptome (cf.\ \citep{tasic2017shared}).
+Unfortunately, modern high-resolution cell photography and single-cell sequencing technologies are both destructive.
+As a result, we can't get both kinds of data for the same specimens.
+For cells native to regions full of diverse cell-types, finding a correspondence is a real problem.
+The result is two completely independent classifications of cells, one for each way of looking at the cell.
+MLM allows us to estimate the relationship between those two classification systems, yielding a wholistic understanding of the different types of cells.
 
-\item Two labs may used different methods, but may wish to pool results.  Each lab looks at the other
-lab's data and asks themselves "what would their data have looked like if they had used our method?"  When
-they find they don't know the answer, it greatly complicates collaboration.
+    \item Cancer treatment efficacy prediction.
+Starting from in-vivo human cancers, many cell-lines have been cultured over the years.
+These cell cultures live indefinitely on plates.
+Many experiments have been performed to see how these cancer cells respond to treatment.
+However, if a treatment work on a particular cultured cell-line, what can we say about which kinds of cancer might respond well to that treatment?
+Coarse side-information such as original cancer location is often available for both in-vivo and cultured cells, but this is often a surprisingly weak signal.
+Cell transcriptomes provides much more specific information about the cancer, and thus, in theory, what treatments might be appropriate (cf.\ \cite{cieslik2018cancer}).
+However, we know that cultured cell-lines look quite different from in-vivo cells (cf.
+\cite{imamura2015comparison,haibe2013inconsistency}).
+Moreover, we have very little joint measurement between the human cells which originally gave rise to a cell culture and the cells that survived to beceome the cell line.
+The Markov Link method can leverage the common side-information together with separate transcriptome information to produce a fine-grained correspondence between in-vivo and cultured cells.
+This correspondence can used to propose new treatments for cancers.
 
-\item One may attempt to cluster the data.  It is simple to cluster the data from one technique, and cluster the data
-from another technique -- but how can we know how those clusters correspond to one another?
+    \item Text/image correspondence.
+Automatic image captioning is an ongoing effort in machine learning (cf.\ \cite{srivastava2018survey}).
+There are three types of data available to help develop such algorithms: text-only data, image-only data, and paired-text-and-image data.
+Obviously the last kind is the most useful for automatic image captioning, but there is much less of it.
+The Markov Link Method suggests one way to use the more plentiful text-only and image-only data.
+We can first apply classic machine learning techniques to get coarse labels for both kinds of data.
+Using this side-information to identify subpopulations, the MLM can then deduce a fine-grained correspondence between text and images by combining information from across all the subpopulations.
 
+    \item Replication crisis and lab effects.
+Replicating a published study is not always an easy thing to do.
+This difficulty is commonly attributed to selective publication bias, bad design, poor description of methods, and even outright fraud \cite{baker2016reproducibility}.
+A calibration would allow us to understand this problem in detail.
+If two labs perform identical experiments and get different data, that does not mean we need to throw out both datasets.
+Instead, we can use MLM to calibrate the tools.
+Once the tools are properly calibrated, we can combine both datasets.
+Unlike other tools to deal with lab or batch effects (e.g.
+\cite{crow2018characterizing,johnson2007adjusting}), MLM makes zero assumptions about what calibrations we might expect.
 \end{itemize}
-
-In this work we propose the Markov Link Method (MLM) as a way to get useful and statistically rigorous bounds on the relationship between two measurement tools even when joint measurements are unavailable, hopefully yielding insight for the kinds of difficulties illustrated above.
-In particular, we consider the case that we can obtain samples from a variety of subpopulations.
-If we can assume that the \emph{relationship} between the measurements is the same in each subpopulation, we can use these samples to learn something about that relationship -- even though we never see both measurements for any single member of the population.
 
@@
 Perhaps the closest work to this one would be \cite{makarov1982estimates}, which used two marginal distributions to get bounds on a property of the joint distribution (namely the distribution of the sum).
-We build on this approach, both by using subpopulations to considerably refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.
+We advance this approach to a more general-purpose technique, both by using many subpopulations to refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.
 
@@
 A key question immediately appears from these tables: does the way in which we have clustered cells into different types under technique I line up nicely with the way in which we have clustered cells under technique II?
-The Markov Link Method gives us a way of addressing this question that is completely agnostic about the underlying genetic markers that were used to produce the clusters.
+The Markov Link Method gives us a way of addressing this question that is completely agnostic about the underlying genetic markers that were used to produce the clusters (this in constrast to methods such as \cite{kiselev2018scmap} or \cite{crow2018characterizing} which assume both that the clusters arise from observables common to both techniques and that those observables are already at least marginally calibrated between the methods).
 Our output was a set $\hat \Theta$ of possible ways that technique I cell-types could line up against the technique II cell-types.

### 2018-04-20 tryg/batteries.py

@@ -235,11 +235,13 @@ def elementwise_monotone(name,inp,dtype,bias_initializer=0.0,bias=True,n_basis=4
         mult_almost = get_variable2('mult_almost',(1,shp,n_basis),dtype,yinit)
         mult = tf.nn.softplus(ysoftplusscale+mult_almost,name='mult')
 
+        final_unreduced = mult * tf.atan(inp+shift,name='final_unreduced')
+
         if bias:
-            final = tf.reduce_sum(mult * tf.atan(inp+shift),axis=2,name='final_before_bias')
+            final = tf.reduce_sum(final_unreduced,axis=2,name='final_before_bias')
             final = tf.identity(bias+final,name='final')
         else:
-            final = tf.reduce_sum(mult * tf.atan(inp+shift),axis=2,name='final')
+            final = tf.reduce_sum(final_unreduced,axis=2,name='final')
     return tf.identity(final,name)
 
 '''

### 2019-04-17 paper/paper.tex
First draft

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{mathrsfs}

\usepackage{graphicx}

\usepackage[algoruled]{algorithm2e}
\usepackage{listings}
\lstset{language=python}

\newcommand{\blind}[2]{#1}

\newtheorem{thm}{Theorem}
\newtheorem*{thm*}{Theorem}
\newtheorem{lemma}[thm]{Lemma}
\newtheorem*{lemma*}{Lemma}

\theoremstyle{definition}
\newtheorem{definition}[thm]{Definition}

\begin{document}

\title{Unpaired measurement integration using Markov Link Intervals}

\author{\blind{Jackson Loper$^{a,b}$\footnote{To whom correspondence should be addressed: email jl5116@columbia.edu.}, Trygve Bakken$^c$, Uygar Sumbul$^c$, \\ Gabe Murphy$^c$, Hongkui Zeng$^c$, David Blei$^{b,d}$, Liam Paninski$^{a,b}$ \\
\small $^a$Grossman Center for the Statistics of Mind, Zuckerman Mind-Brain Behavior Institute, \\ \small Department of Neuroscience, and Center for Theoretical Neuroscience, Columbia University, NY NY \\
\small $^b$Department of Statistics, Columbia University, NY NY \\
\small $^c$Allen Institute for Brain Science, Seattle WA \\
\small $^d$Department of Computer Science, Columbia University, NY NY}{[Authors blinded]}}

\maketitle

% https://academic.oup.com/biomet/article/106/1/145/5250868
% We consider nonparametric estimation of a covariance function on the unit square, given a sample of discretely observed fragments of functional data. When each sample path is observed only on a subinterval of length δ<1⁠, one has no statistical information on the unknown covariance outside a δ-band around the diagonal. The problem seems unidentifiable without parametric assumptions, but we show that nonparametric estimation is feasible under suitable smoothness and rank conditions on the unknown covariance. This remains true even when the observations are discrete, and we give precise deterministic conditions on how fine the observation grid needs to be relative to the rank and fragment length for identifiability to hold true. We show that our conditions translate the estimation problem to a low-rank matrix completion problem, construct a nonparametric estimator in this vein, and study its asymptotic properties. We illustrate the numerical performance of our method on real and simulated data.

\begin{abstract}
We consider two datasets, each measuring a different of observations on a common population.  The first dataset is constructed by drawing individuals randomly from this population and measuring a quantity about each individual.   The second dataset is constructed similarly, but a different quantity is measured.  ``Unpaired measurement integration'' attempts to use this data to infer the joint distribution of the two different quantities.  We overcome the unidentifiability problems in this task by measuring an additional quantity, similar to an instrumental variable.  This quantity is observed in both datasets and satisfies a conditional independence assumption.  We characterize identifiability in the presence of these variables and develop a consistent estimator.  The performance of the estimator is tested in simulations.  The method is applied to transcriptomic measurements to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.
\end{abstract}

\section{Introduction}

Unpaired measurement integration seeks to infer the joint distribution on two variables $(X,Y)$ using knowledge of the marginal distributions on $X$ and $Y$.  This problem is ill-posed, but practical algorithms have nonetheless been developed.  An early approach by Andrey Kolmogorov noted that the expectation for any linear function $f(X,Y)=\alpha X+\beta Y$ could be effectively bounded using only the marginal distributions \cite{makarov1982estimates}.  Bounds on arbitrary univariate functions $f(X,Y)$ have also been developed in the context of causal inference \cite{balke1994counterfactual,FanYan2017}.  More recently, the CycleGan algorithm has gained prominence in the computer graphics community for succesfully estimating a joint distribution between photographs ($X$) and watercolor paintings ($Y$) of the same subject; CycleGan accomplishes this unpaired measurement integration using two completely separate datasets, one of photographs and one of watercolor paintings \cite{zhu2017unpaired}.

To resolve the fundamental ill-posedness of unpaired dataset integration, some form of extra information is often employed.  We assume this side-information variable can be observed for each sample in both datasets.  The work of Fan \emph{et al} (2017) gives a detailed overview for this case \cite{FanYan2017}.  Fan characterizes the asymptotics of estimators for $\mathbb{E}[f(X,Y)]$, where $f$ can takes a wide variety of forms.  Estimator asymptotics are established under many different assumptions on the relationship between the measurements and the side-information

Inspired by problems in the study of cell populations, this paper considers a new kind of assumption about the side-information.  Let $L$ denote the side-information.  Suppose that $L$ and $Y$ are conditionally independent given $X$.  This assumption holds naturally whenever both $L$ and $Y$ can both be understood as independent noisy measurements of $X$.  We give a complete real-world example in Section \ref{sec:allen}.  The variable $L$ bears a formal similarity to an instrumental variable, but its function is different: instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.

Inference under the conditional independence assumption must overcome two obstacles: the datasets only allow approximate estimation for the marginal distributions, and the marginal distributions only give a partial picture of the joint.  These problems cannot be considered separately.  If we treat statistical estimates as exact truth and proceed with a purely algebraic analysis of the model, inference can go badly wrong.  It turns out that even infinitesimal estimation errors can result in substantially incorrect conclusions, as we show in Lemma \ref{lem:noobseq}.   This marks a stark contrast from the conditions considered in Fan \emph{et al} (2017), where additional statistical noise results in a graceful degradation of inference accuracy.

In this paper, we overcome both of these challanges for the discrete setting.  When $L,X,Y$ are categorical distributions it is possible to make a detailed study of what are known as the observational equivalence classes \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.  Each class corresponds to a set of parameters that give rise to the same distribution on the data we observe.  We can achieve consistent inference by taking a carefully chosen dilation of these sets.  We design a simple optimization algorithm to find extremal points in these dilated sets.  These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.    Performance is investigated through a variety of simulation studies.  Finally, the estimator is applied to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.

\newcommand{\DX}{\mathcal{D}^X}
\newcommand{\DY}{\mathcal{D}^Y}
\newcommand{\DV}{\mathcal{D}}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\pr}[1]{\mathrm{pr}\left(#1\right)}

%              _        _   _             
%  _ __   ___ | |_ __ _| |_(_) ___  _ __  
% | '_ \ / _ \| __/ _` | __| |/ _ \| '_ \ 
% | | | | (_) | || (_| | |_| | (_) | | | |
% |_| |_|\___/ \__\__,_|\__|_|\___/|_| |_|

\section{Formal problem statement and notation}

Consider a triple of discrete random variables variables, $(L,X,Y) \sim p$. We assume $p$ has support on $\{1\cdots N_L\}\times\{1\cdots N_X\}\times\{1\cdots N_Y\}$ and $L,Y$ are conditionally independent given $X$.  
% We further assume that $N_X > N_L$ (if $N_X\leq N_L$ it is straightforward to show that inference can almost always be performed with ordinary estimation methods).  
We parameterize $p$ as
\[
p(\ell,x,y;\pi,\alpha,\beta)=\pi_\ell \alpha_{\ell,x} \beta_{x,y}
\]
where $\pi$ lies strictly in the interior of the $N_L$-simplex, $\alpha$ is an $N_L \times N_X$ right-stochastic matrix, and $\beta$ is an $N_X \times N_Y$ right-stochastic matrix.  We will also find it useful to let $\gamma_{\ell y}$ denote $p(y|\ell)=\sum_x \alpha_{\ell,x} \beta_{x,y}$.

Suppose we are given independent samples from $p$, but some data is missing.  Each sample includes $L,X$ or $L,Y$ but no sample ever includes both $X$ and $Y$.  Can we still learn $p$?  Let $n,m>0$ and $(L_1,X_1,Y_1),(L_2,X_2,Y_2) \cdots (L_{n+m},X_{n+m},Y_{n+m})$ denote samples of independent copies of $(L,X,Y)$.  Let
%
\begin{align*}
\DX &= (L_1,X_1),\cdots (L_n,X_n) \\
\DY &= (L_{n+1},Y_{n+1}),\cdots (L_{n+m},Y_{n+m}) 
\end{align*}
%
Our task is to use the data $(\DX,\DY)$ to estimate the parameters of $p$, namely $(\pi,\alpha,\beta)$.  Estimating $\pi,\alpha$ from this data is fairly straightforward.  However, existing estimation theory is less helpful for the parameter $\beta$.  In this paper we develop theory and practice for estimating this parameter.

%            _   _                 _   _             
%   ___  ___| |_(_)_ __ ___   __ _| |_(_) ___  _ __  
%  / _ \/ __| __| | '_ ` _ \ / _` | __| |/ _ \| '_ \ 
% |  __/\__ \ |_| | | | | | | (_| | |_| | (_) | | | |
%  \___||___/\__|_|_| |_| |_|\__,_|\__|_|\___/|_| |_|

\section{Markov Link Intervals (MLI) for estimating $\beta$}

Correct inferences about $\beta$ require careful treatment of its identifiability.  Our first result is a negative one:

\begin{lemma}[Unidentifiability]  \label{lem:unident} When $N_L<N_X$ there is no generally consistent estimator to estimate $\beta$ from the data $(\DX,\DY)$.  
\end{lemma}

Unfortunately, the $N_L<N_X$ regime is precisely the one which motivated our present work, inspired by investigation of cell-types in neuroscience.  This result shows that perfect estimation of $\beta$ may not be possible in our case.

When a parameter cannot be consistently estimated, one common approach is to bound the unidentifiable parameter using  ``observational equivalence classes'' \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.  In our case, this equivalence class is defined as the set of all the different parameters $(\pi',\alpha',\beta')$ which would give rise to the true distribution on the observable variables $(\DX,\DY)$.  Even though $\beta$ is unidentifiable, we might hope to identify this set of possible parameters.  Our second result shows this approach is unlikely to succeed:

\begin{lemma}[Discontinuity of observational equivalence classes]  \label{lem:noobseq} Endow the space of observable distributions with the total variation metric.  Endow the space of sets of parameters $\pi,\alpha,\beta$ with the Pompeiu-Hausdorff metric.  The map from possible distributions on $\DX,\DY$ to observational equivalence classes of parameters is not continuous.   This mapping is sometimes discontinuous even around points where the rows of $\alpha$ are linearly independent and all parameters lie in the interior of the natural parameter space.  
\end{lemma}

Even infinitesimal error in the estimation of the observable distribution may thus yield large errors in our estimates of the observational equivalence class.

A partial resolution to this difficulty can be obtained if it is known that $\beta$ is strictly positive and the rows of $\alpha$ are linearly independent.  Under these conditions we can first estimate $\alpha$ and then estimate the observational equivalence class for $\beta$ for the observations $\DY$ under a fixed value for $\alpha$.  It turns out that the distance between this class and $\beta$ does converge to zero:

\begin{thm}[Consistency for positive $\beta$] Let the rows of $\alpha$ be linearly independent and let $\beta_{x,y}\geq c>0$.  Define $\gamma$ as the $N_L\times N_Y$ right-stochastic matrix
\[
\gamma_{\ell y} \triangleq \sum_{x} \alpha_{_\ell x} \beta_{x,y}
\]
Let $\hat \pi, \hat \alpha, \hat \gamma$ denote any consistent estimators for $\pi,\alpha,\gamma$.  Consider $\{\beta':\ \hat \alpha\beta'=\hat \gamma \}$, the equivalence class for the observations $\DY$ after fixing the estimate $\hat \alpha$.  The distance between the point $\beta$ and this set converges to zero in probability.  
\end{thm}

A more general solution can be obtained by seeking a small dilation of the observational equivalence classes.  Let $\LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta)$ denote the likelihood of the observed data under the model under the parameters $\hat \pi,\hat \alpha,\hat \beta$.  Let $\LL^*$ denote the maximum value of $\LL$ over all possible parameters.  Define
\[
\hat \Theta=\left\{\hat \pi, \hat \alpha,\hat \beta:\ \LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta) > \LL^* - \log(n+m)\right\}
\]
This log-scaled likelihood envelope has two desirable properties: it both concentrates around the true distribution on the observables and is asymptotically guaranteed to contain the true model parameters.
\begin{thm}[Consistency of $\hat \Theta$]\label{thm:consistency} Let $n,m\rightarrow\infty$ such that $|\log(n/m)|<M$.  Then:
\begin{enumerate}
    \item $\hat \Theta$ is asymptotically guaranteed to contain the true parameter values.  
    \item $\hat \Theta$ concentrates for $\pi,\alpha,\alpha\beta$. Let $\xi(\pi',\alpha',\beta')$ denote the maximum of the Euclidean differences $|\pi-\pi'|$, $|\alpha-\alpha'|$, $|\alpha\beta-\alpha'\beta'|$.  Then $\max_{\theta \in \hat \Theta} \xi(\theta)$ converges to zero in probability.   
\end{enumerate}
\end{thm}
This theorem indicates a practical solution for estimating $\beta$.  For each $x,y$, we define the \emph{Markov Link Interval} as 
\[
[\hat \beta^-_{x,y},\hat \beta^+_{x,y}] \triangleq \left[\min_{\pi',\alpha',\beta' \in \hat \Theta} \beta'_{x,y},\max_{\pi',\alpha',\beta' \in \hat \Theta}\beta'_{x,y}\right]
\]  
Theorem \ref{thm:consistency} shows that $\beta_{x,y}$ is asymptotically included in the interval $[\hat \beta^-_{x,y},\hat \beta^+_{x,y}]$.

To calculate the Markov Link Interval we must solve the associated optimization problems.  Due to the high dimensionality of the problem and the ill-conditioning of the matrices involved, we were unable to tune off-the-shelf optimization methods to reliably solve these problems.  We developed an alternative optimization procedure which relies on a minorization-maximization approach and requires no tuning parameters.  Appendix \ref{sec:mlmdetails} provides a detailed outline of the algorithm.\footnote{Code is published at https://github.com/jacksonloper/markov-link-method, including a tutorial ipython notebook that details every computation made in this paper.}

%      _                 _       _   _                 
%  ___(_)_ __ ___  _   _| | __ _| |_(_) ___  _ __  ___ 
% / __| | '_ ` _ \| | | | |/ _` | __| |/ _ \| '_ \/ __|
% \__ \ | | | | | | |_| | | (_| | |_| | (_) | | | \__ \
% |___/_|_| |_| |_|\__,_|_|\__,_|\__|_|\___/|_| |_|___/

\section{Simulations}

\begin{figure}\fbox{\begin{minipage}{\textwidth}
% \begin{tabular}{cc}
% \includegraphics[width=.5\textwidth]{simIII.png} 
% & \includegraphics[width=.5\textwidth]{simII.png} \\
% %
% \begin{minipage}{.5\textwidth}(a) Performance depends on the $(L,X)$ dependency\end{minipage} 
% & \begin{minipage}{.5\textwidth}(a) Performance depends on the $(X,Y)$ dependency\end{minipage} 
% \end{tabular}
\hfill{}\includegraphics[width=.9\textwidth]{simident.png}\hfill{}

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{Depending upon the exact model parameters, maximum likelihood estimation for $\beta$ may fail.}  We conducted a series of trials; in each trial we picked a sample size and ground-truth model parameters, simulated a dataset, calculated a maximum likelihood estimator for $\beta$, and compared it against the ground-truth.  Even when the sample size is effectively infinite, this estimator can fail.  Here we explore these failure modes.  To produce subfigure (a) each row of $\beta$ is drawn uniformly from the simplex; we explore how estimation error changes as we vary $\alpha$ to control the strength of the dependency between the variable $L$ and $X$.  When $L\perp X$, we see that maximum likelihood estimation fails significantly even with an essentially infinite number of samples.  To produce subfigure (b), we draw each row of $\alpha$ uniformly from the simplex and vary $\beta$ to control the dependency between $X$ and $Y$.  When $X\perp Y$ or $X=Y$ it is possible to get accurate estimates of $\beta$.  However, in a more typical case where the relationship between $X,Y$ is neither independent nor deterministic, we see that the maximum likelihood estimator fails to accurately determine $\beta$, even in the limit of essentially infinite data.  Both plots are consistent with Lemma \ref{lem:unident}, which states that there is no estimator for $\beta$ which will be consistent for every  ground-truth parameter. \label{fig:ident}}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

The previous section gives generic theoretical results about inference for the parameter $\beta$.  In this section we use simulations to get a more detailed understanding.

Our first simulations investigate Lemma \ref{lem:unident}, which states that there are cases where $N_L<N_X$ and $\beta$ cannot be identified even in the limit of infinite data.  We test this Lemma by exploring the performance of a maximum likelihood estimator for $\beta$ on different simulated datasets.  By varying the sample size and the ground-truth parameters of the model, we can explore this unidentifiability in detail:

\begin{itemize}
    \item Varying the sample size and the strength of the $L,X$ dependency.  Fix $N_X=N_Y=20$ and $N_L=15$.  We perform a series of trials.  In each trial we select $\pi$ and the rows of $\beta$ uniformly from the simplex.  The rows of $\alpha$ are chosen more carefully.  We will first generate $\tilde \alpha_1$ by sampling the rows uniformly from the simplex.  We then generate $\tau$ uniformly from the simplex and define $\tilde \alpha_2 = \mathbf{1}\tau^T$, i.e. every row of $\alpha_2$ is given by $\tau$.  Finally, we select $c\in[0,1]$ and define $\alpha=(1-c)(\tilde \alpha_1)+c\tilde \alpha_2$.  By varying $c$, we can control the mutual information between $L$ and $X$.  When $c=0$ the variable $L$ gives significant information about $X$; when $c=1$ we have that $L$ is independent of $X$.  Once $\pi,\alpha,\beta$ are selected, we select a sample size $n=m$, simulate a dataset, compute a maximum likelihood estimator $\hat \beta $ (see Appendix \ref{sec:mlmdetails} for our exact algorithm), and compute the average variation distance between $\beta$ and our estimate $\hat \beta$.  This average total variation distance is computed by averaging the row-by-row total variations between $\beta$ and  $\hat \beta$.  After performing a series of such trials, we visualize the results to obtain qualitative understanding of the performance of the maximium likelihood estimator $\hat \beta$.

    \item Varying the sample size and the strength of the $X,Y$ dependency.  Fix $N_X=N_Y=20$ and $N_L=15$.  We perform a series of trials.  For each trial, we select $\pi$ and the rows of $\alpha$ uniformly from the simplex.  To determine $\beta$, we first generate $\tau$ uniformly from the $N_Y$-simplex.  We then select $c\in[0,1]$ and define $\beta=c(\mathbf{1}\tau^T)+(1-c)I$.  By varying $c$, we can control the strength of dependency between $X$ and $Y$.  When $c=0$ we have that $X=Y$ (degenerate dependency); when $c=1$ we have that $X\perp Y$ (complete independence).  As above, for each trial we then compute a maximum likelihood estimator and compare the estimator to the ground-truth parameter.   
\end{itemize}

Figure \ref{fig:ident} shows the results.  We see that $\beta$ is much harder to estimate in some situations.  For example, the first simulation (Figure \ref{fig:ident}a) demonstrates that $\beta$ is harder to estimate when $L,X$ have low mutual information.  The second simulation (Figure \ref{fig:ident}b) shows a more complex picture; in some cases it is actually possible to get perfect estimation of $\beta$, even though $N_L<N_X$.  If $X,Y$ are independent or deterministically related, we see that perfect estimation becomes possible in the limit of infinite data.  However, in the more typical case where the relatioship between $X$ and $Y$ neither independent nor deterministic, it appears that consistent estimation is not possible.

We next investigate Theorem \ref{thm:consistency}, which implies that the Markov Link Intervals are asymptotically guaranteed to include the true parameter values for $\beta$.   Fix $N_X=N_Y=20$ and $N_L=15$.  We perform a series of trials.  In each trial we select $\pi$ and the rows of $\alpha,\beta$ uniformly from the simplex.  We then select a sample size $n=m$, simulate a dataset, and calculate the Markov Link Interval for the parameter $\beta_{1,1}$.  For each sample size $n$ we can plot a histogram of the differences between the upper bound $\hat \beta^{+}_{1,1}$ and the true parameter $\beta_{1,1}$.  A small positive value is ideal, indicating a tight bound on the true parameter.  Negative values are undesirable, since they indicate that the true parameter was in fact greater than the upper bound of the Markov Link Interval.  Large values are also undesirable since they indicate the bound was loose.  We can apply the same procedure to the differences between $\beta$ and $\hat \beta^{-}$.  By inspecting these histograms we obtian a qualitative sense for the performance of these intervals, complementing the theoretical guarantees of Theorem \ref{thm:consistency}.  The results are shown in Figure \ref{fig:simbound}.  We see that the true parameters are reliably bounded by the Markov Link Intervals, and these intervals shrink as we get more data.  However, even in the limit of essentially infinite data ($n=10^{10}$) we note that the width of the intervals does not vanish entirely.  This is consistent with the unidentifiability of $\beta$ demonstrated in Lemma \ref{lem:unident}.

These simulations are by no means exhaustive.  However, they yield a high-level overview of the estimation problem for this model.  First, estimation is easier when:
\begin{itemize}
    \item $L$ is highly informative about $X$.  
    \item $X,Y$ are either highly dependent or nearly independent
    \item Sample size is high
\end{itemize}
Second, even when estimation is difficult, the Markov Link Intervals give bounds which reliably contain the true parameter values.

\begin{figure}\fbox{\begin{minipage}{\textwidth}
% \begin{tabular}{cc}
% \includegraphics[width=.5\textwidth]{simIII.png} 
% & \includegraphics[width=.5\textwidth]{simII.png} \\
% %
% \begin{minipage}{.5\textwidth}(a) Performance depends on the $(L,X)$ dependency\end{minipage} 
% & \begin{minipage}{.5\textwidth}(a) Performance depends on the $(X,Y)$ dependency\end{minipage} 
% \end{tabular}
\hfill{}\includegraphics[width=.9\textwidth]{simbound.png}\hfill{}

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{The Markov Link Intervals are conservative bounds around the true parameter estimates.}  We conducted a series of trials; in each trial we picked a sample size and ground-truth model parameters, simulated a dataset, calculated the Markov Link Intervals $\hat \beta^{-},\hat \beta^{+}$, and compared them against the truth.  For each trial we calculated both an upper slack ($\hat \beta^{+}-\beta$) and a lower slack ($\hat \beta^{-}-\beta$).   For each sample size $n \in \{100,500,1000,10^4,10^5,10^{10}\}$, we computed a histogram of all the upper and lower slack values in all of the simulations performed with that sample size.  We observe first that for every $n$ the upper slack is always positive and the lower slack is always negative, indicating the true value was contained inside our bound for every simulation.  Moreover, as the number of samples increases, we see that we can obtain tighter bounds; the magnitude of the upper and lower slacks decreases.  However, even in the limit of essentially infinite data ($n=10^{10}$) we note that the slack does not converge to zero.  This is consistent with the unidentifiability of $\beta$ demonstrated in Lemma \ref{lem:unident}. \label{fig:simbound}}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

%        _ _            
%   __ _| | | ___ _ __  
%  / _` | | |/ _ \ '_ \ 
% | (_| | | |  __/ | | |
%  \__,_|_|_|\___|_| |_|

\section{Experiments for cell-types}
\label{sec:allen}

\begin{figure}\fbox{\begin{minipage}{\textwidth}
% \begin{tabular}{cc}
% \includegraphics[width=.5\textwidth]{simIII.png} 
% & \includegraphics[width=.5\textwidth]{simII.png} \\
% %
% \begin{minipage}{.5\textwidth}(a) Performance depends on the $(L,X)$ dependency\end{minipage} 
% & \begin{minipage}{.5\textwidth}(a) Performance depends on the $(X,Y)$ dependency\end{minipage} 
% \end{tabular}
\hfill{}\includegraphics[width=.9\textwidth]{allen.png}\hfill{}

\hfill{}\begin{minipage}{.9\textwidth}
\caption{\textbf{Cell-type science: maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
%
We performed unpaired measurement integration for two ways of measuring the `type' of mouse cells.  The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.  The second method yields a considerably less reliable estimate for the cell-type, $Y$.  For each cell we also know the value of $L$, which corresponds to another noisy indicator about the cell-type $X$.  In one dataset we can observe $(L,X)$ for many cells; in another dataset we can observe $(L,Y)$ many other cells.  Using this data we construct three estimates for $\beta$, the matrix of parameters which defines the relationship between $X$ and $Y$.  For each possible index $(x,y)$, $\hat \beta^{-}_{x,y}$ estimates a lower-bound estimate for $\beta_{x,y}$, $\hat \beta_{x,y}$ is a maximum likelihood estimator, and $\hat \beta^{+}$ estimates an upper bound.  Due to identifiability issues $\hat \beta_{x,y}$ is not guaranteed to be consistent.  However, Theorem \ref{thm:consistency} shows that the Markov Link Interval $[\hat \beta^{-}_{x,y},\hat \beta^{+}_{x,y}]$ is asymptotically guaranteed to include the truth.  Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ cell indicates the value of the corresponding parameter.  The values of $\hat \beta$ are essentially nonsensical in many cases.  On the other hand, the Markov Link Interval suggests plausible bounds for the parameters and also indicates further experiments which would help refine these bounds.
}
\end{minipage}\hfill{}
\end{minipage}}\end{figure}

We now apply Markov Link Intervals to study the scientific problem which motivated their development.

Every cell in a human body has the same DNA (to a first approximation), but some cells behave differently from others.  The biomolecular processes that drive this diversity are an area of active research \cite{wagner2018single,boudreau2018cell,he2018mechanical}.  Efforts such as the Human Cell Atlas project seek to map out a taxonomy of cell-types \cite{rozenblatt2017human}, thereby enabling a more systematic study of cellular diversity.

At least at a coarse level, the cell-type $X$ of a given cell can be reliably determined by applying RNA sequencing techniques \cite{okaty2011cell}.  However, these techniques only measure one property of the cell: RNA expression. To determine the relationship between cell type and other properties of the cell, alternative techniques are required.  ``Patch-seq'' is one such approach \cite{cadwell2016electrophysiological}; this method obtains RNA, electrophysiological, and morphological properties at the single-cell level.  Unfortunately, the richness of the data comes at the cost of a degraded RNA signal, leading to a noisy estimate of the cell-type.  We will denote this noisy estimate as $Y$.

For any cell it is thus impossible to measure a rich variety of cell properties \emph{and} reliably estimate the cell-type.  To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement integration.  This integration would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.  To facilitate this integration, we can construct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.  In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.  By varying the experimental parameters we can change the distribution on which cells are picked out.  Letting $L$ denote these corresponding experimental parameters, we see that $L$ can act as a noisy independent indicator about the cell-type.  Markov Link Intervals should therefore give asymptotically consistent bounds on the reliability of the noisy estimate $Y$.

The Allen Institute recently conducted a series of experiments using both traditional RNA sequencing as well as the Patch-seq technique \cite{tasic2017shared}.  The same Cre/Lox systems were used for both techniques, thereby allowing us to observe the indicator $L$ for each sample in both datasets.  For each sample in the traditional RNA sequencing dataset (``Standard'') the cell-type was determined.  For each sample in the Patch-seq dataset we also have an estimated cell-type, although the Patch cell-type estimates are only given at a coarse level of the taxonomy of cells.  For example, in the Standard dataset a distinction is made between cell-type `Lamp5 Egln 1' and cell-type `Lamp5 Pdlim5,' whereas in the Patch-seq dataset we group all GABAergic neuronal cell-types characterized by high expression of `Lamp5' into a single coarse `cell class.'

We applied the Markov Link Intervals to the Allen Institute data try to determine the reliability of the noisy cell-type estimators $Y$ arising from Patch-seq approach.  We can visualize a maximum likelihood estimator as well as the Markov Link Intervals through heatmaps (see Figure \ref{fig:allen})).  We see that the maximum likelihood estimator sometimes gives essentially nonsensical results.  For example, the maximum likelihood estimator suggests that a particular group of excitatory glutamatergic neurons from visual cortex are \emph{always} assigned to a GABAergic inhibitory type by the $Y$ variable.  This is quite unlikely, since even the noisy measurements of $Y$ can reliably determine between excitatory and inhibitory neurons \cite{cadwell2016electrophysiological,cadwell2017multimodal}.

The Markov Link Intervals yield plausible bounds on the true model parameters.  For example, the lower bounds of the Markov Link Interval suggests that a sizeable fraction of `Lamp5 Pdlim5' cells are indeed classified as `Lamp5' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as `Lamp5 Pdlim5', but instead only gives coarse class-level cell-type predictions, such as `Lamp5').  The upper bounds furthermore show that these cells are seldom classified as anything \emph{but} `Lamp5' cells or possibly the `unknown' type which the Patch method assigns to cells which it cannot identify.  For cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type; the intervals indicate that the Patch method does \emph{not} classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate  how they \emph{are} being classified.  At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.   These `Lamp5 Elgn3-1' cells might get perfectly classified or they may get poorly classified; the available data cannot tell us either way.

Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.  For example, the significant ambiguity for cells with standard type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.  If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which measure as Patch type `Pvalb,' this would show that the Standard `Lamp5 Egln 1' is not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells but \emph{only} cells with Patch type `Pvalb,' this would show the opposite.  If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.

%                       _           _                 
%   ___ ___  _ __   ___| |_   _ ___(_) ___  _ __  ___ 
%  / __/ _ \| '_ \ / __| | | | / __| |/ _ \| '_ \/ __|
% | (_| (_) | | | | (__| | |_| \__ \ | (_) | | | \__ \
%  \___\___/|_| |_|\___|_|\__,_|___/_|\___/|_| |_|___/

\section{Conclusions}

We here study `unpaired measurement integration,' the problem of estimating a joint distribution $(X,Y)$ from two separate datasets, one with samples of $X$ and another with samples of $Y$.  Rigorous statistical analysis of this ill-posed problem are complicated by statistical error and identifiability issues.  Here we characterize both issues in the discrete settings when we can observe a third variable $L$ for each sample in both dataset, under the assumption that $L$ and $Y$ are independent conditioned in $X$.  This assumption arises naturally if $L$ gives noisy measurement about $X$ and that noise is independent of $Y$.  Under this assumption, we construct ``Markov Link Intervals'' which give provably consistent bounds on the joint distribution.  Simulations bear out the consistency of these bounds and also suggest estimation is easiest performs best when $L$ is predictive of $X$ and the relationship between $X$ and $Y$ is either deterministic or completely independent.  An application of the method to cell-type populations demonstrates the method's practical utility for unpaired measurement integration problems.  One can imagine that these results could be extended to the continuous setting, though additional regularity conditions would be required.

\bibliographystyle{unsrt}
\bibliography{refs}

%     _    ____  ____  _____ _   _ ____ _____  ___ 
%    / \  |  _ \|  _ \| ____| \ | |  _ \_ _\ \/ / |
%   / _ \ | |_) | |_) |  _| |  \| | | | | | \  /| |
%  / ___ \|  __/|  __/| |___| |\  | |_| | | /  \|_|
% /_/   \_\_|   |_|   |_____|_| \_|____/___/_/\_(_)

\appendix

\section{Markov Link Interval details}
\label{sec:mlmdetails}

Let $n,m>0$ and $(L_1,X_1,Y_1),(L_2,X_2,Y_2) \cdots (L_{n+m},X_{n+m},Y_{n+m}) \sim p$, where $p(\ell,x,y)=\pi_\ell \alpha_{\ell,x} \beta_{x,y}$.  However, for $i\in 1\cdots n$ we cannot observe $Y_i$ and for $i\in n+1 \cdots n+m$ we cannot observe $X_i$.  Let $\LL(\theta)$ denote the likelihood of the observed data under the model under any particular choice of the parameters $\theta = \pi',\alpha',\beta'$.  Let $\LL^*$ denote the maximum value of $\LL$ over all possible parameters.  Define
\[
\hat \Theta=\left\{\theta:\ \LL(\theta) > \LL^* - \log(n + m)\right\}
\]
Our task is to calculate
\begin{align*}
[\hat \beta^-_{x,y},\hat \beta^+_{x,y}] &\triangleq \left[\min_{\pi',\alpha',\beta' \in \hat \Theta} \beta'_{x,y},\max_{\pi',\alpha',\beta' \in \hat \Theta}\beta'_{x,y}\right]
\end{align*}
Proceed as follows:

\begin{enumerate}
    \item Summarize the data into two matrices:
    \begin{align*}
    \DX_{\ell x} &= \#\{i\leq n: L_i=\ell,X_i=x\}\\
    \DY_{\ell y} &= \#\{i> n: L_i=\ell,Y_i=y\}
    \end{align*}

    \item Estimate the maximum likelihood.  The likelihood is given as
    \begin{align*}
    \LL(\pi,\alpha,\beta) = \sum_{\ell x} \DX_{\ell x} (\pi_\ell + \log \alpha_{\ell x}) + \sum_{\ell y} \DX_{\ell y} \left( \pi_\ell + \log \sum_x \alpha_{\ell x} \beta_{x,y}\right)
    \end{align*}
    We optimize this likelihood using Expectation Maximization.  Let
    \[
    \tau_{\ell x y} = \alpha_{\ell x}\beta_{x,y} / \sum_x \alpha_{\ell x'}\beta_{x',y}
    \]
    denote a posterior summary for the distribution of the unobserved values of the measurement $X$.  We then update the parameters by solving the surrogate problem
    \begin{align*}
    \max \left(\sum_{\ell x} \DX_{\ell x} (\pi_\ell + \log \alpha_{\ell x}) + \sum_{\ell y} \DX_{\ell y} \left( \tau_\ell + \sum_x \tau_{\ell x y}\log\alpha_{\ell x} \beta_{x,y}\right)\right)
    \end{align*}
    Note that this surrogate problem can be solved in closed form.  We proceed by iteratively updating our posterior summary then solving the the surrogate problem.  We terminate the algorithm when all parameters change less than $10^{-7}$ in a single iteration.

    \item Define 
    \begin{align*}
    \hat \Theta &= \left\{\theta:\ \LL(\theta) > \LL(\hat \theta) - \log(n+m)\right\}\\
    [\hat \beta^-_{x,y},\hat \beta^+_{x,y}] &\triangleq \left[\min_{\beta' \in \hat \Theta} \beta'_{x,y},\max_{\beta' \in \hat \Theta}\beta'_{x,y}\right]
    \end{align*}
    Let $x^*,y^*$ index a particular parameter for which we would like to compute $\hat \beta^-_{x^*,y^*},\hat \beta^+_{x^*,y^*}$.  We compute as follows:
    \begin{itemize}
        \item To compute $\min_{\beta' \in \hat \Theta} \beta'_{x^*,y^*}$.  The Lagrangian of this constrained problem may be written as
        \[
        V(\beta,\lambda) = \lambda \log \beta_{x^*,y^*} + L(\hat \alpha,\hat \beta)
        \]
        For any fixed $\lambda$, note that this problem has the same functional form as our original problem, and can be solved using the same algorithm as we used to find the maximium likelihood.  The updates for $\alpha$ are in fact identical.  The $\beta$ updates are only slightly different; combining the posterior summary and objective updates into one step, we obtain the updates
        \[
        \beta_{xy} \propto 
            \begin{cases} 
            {\beta}_{xy}\left(\frac{\lambda}{\beta_{xy}}  + \sum_{\ell}\alpha_{\ell x}\frac{\DX_{\ell y}}{\sum_{x}\alpha_{\ell x}{\beta}_{xy}}\right) & x=x^*,y=y^*\\
            {\beta}_{xy}\left(\sum_{\ell}\alpha_{\ell x}\frac{\DX_{\ell y}}{\sum_{x}\alpha_{\ell x}{\beta}_{xy}}\right) & \mathrm{otherwise}\\
            \end{cases}
        \]
        Let $\tilde \beta(\lambda,x^*,y^*)$ denote the optimizer of this inner problem.  To solve our overall problem, we need to find the correct value of $\lambda$, i.e. $\max \{\lambda:\ \tilde \beta(\lambda,x^*,y^*) \in \hat \Theta\}$.  We find this $\lambda$ using line-search.

        \item To compute $\min_{\beta' \in \hat \Theta^\epsilon} \beta'_{x^*,y^*}$.  This problem is equivalent to $\max_{\beta' \in \hat \Theta^\epsilon} \log \sum_{y\neq y^*} \beta'_{x^*,y}$.  This allows us to pose the problem just as we did above.  The resulting updates are given by:
        \[
        \beta_{xy} \propto 
            \begin{cases} 
            {\beta}_{xy}\left(\frac{\lambda}{\sum_{y'\neq y} \beta_{xy'}}  + \sum_{\ell}\alpha_{\ell x}\frac{\DX_{\ell y}}{\sum_{x}\alpha_{\ell x}{\beta}_{xy}}\right) & x=x^*,y\neq y^*\\
            {\beta}_{xy}\left(\sum_{\ell}\alpha_{\ell x}\frac{\DX_{\ell y}}{\sum_{x}\alpha_{\ell x}{\beta}_{xy}}\right) & \mathrm{otherwise}\\
            \end{cases}
        \]
    \end{itemize}

\end{enumerate}

%                         __     
%  _ __  _ __ ___   ___  / _|___ 
% | '_ \| '__/ _ \ / _ \| |_/ __|
% | |_) | | | (_) | (_) |  _\__ \
% | .__/|_|  \___/ \___/|_| |___/
% |_|

\section{Proofs}
\label{sec:proofs}

\begin{lemma*}[Unidentifiability]  When $N_L<N_X$ there is no generally consistent estimator to estimate $\beta$ from the data $(\DX,\DY)$. 
\end{lemma*}

\begin{proof}
    It suffices to show that there exist parameter choices $(\pi,\alpha,\beta)$ and $(\pi',\alpha',\beta')$ where $\beta\neq\beta'$ yet the distribution on $\DX,\DY$ remains the same.  In this case there would be not way to determine which of these paramer choices was correct, so no consistent estimator could be produced.  Note that the distribution on $\DX,\DY$ is entirely determined by $\pi$, $\alpha$, and $\gamma=\alpha\beta$.  Thus it suffices to find an example where $\alpha=\alpha',\beta\neq\beta',\alpha\beta=\alpha'\beta'$.  For a simple example, consider the case $N_L=1,N_X=N_Y=2$ with
    \[
    \alpha=\left(
        \begin{array}{cc}
        \frac{1}{4} & \frac{3}{4}\\
        \end{array}
    \right) \qquad
    \beta=\left(
        \begin{array}{cc}
        0.5 & 0.5\\
        0.1 & 0.9\\
        \end{array}
    \right) \qquad
    \beta'=\left(
        \begin{array}{cc}
        0.8 & 0.2\\
        0 & 1\\
        \end{array}
    \right) \qquad
    \]
\end{proof}

\begin{lemma*}[Discontinuity of observational equivalence classes]  Endow the space of observable distributions with the total variation metric.  Endow the space of sets of parameters $\pi,\alpha,\beta$ with the Pompeiu-Hausdorff metric.  The map from possible distributions on $\DX,\DY$ to observational equivalence classes of parameters is not continuous.   This mapping is sometimes discontinuous even around points where the rows of $\alpha$ are linearly independent and all parameters lie in the interior of the natural parameter space.  
\end{lemma*}

% sudo mount -v -t nfs4 -o nfsvers=4.1,rsize=1048576,wsize=1048576,hard,timeo=600,retrans=2 fs-d6b9599e.efs.us-east-1.amazonaws.com:/ groupstorage/

\begin{proof}
    We start by inspecting the two relevant spaces: the space of distributions on $\DX,\DY$ and the space of sets of parameters $\pi,\alpha,\beta$.
    \begin{itemize}
        \item The possible distributions on $\DX,\DY$.  Each distribution is parameterized by the triple $(\pi,\alpha,\beta)$, where $\pi$ lies in the probability simplex and $\alpha,\beta$ are right-stochastic matrices.  However, note that this parameterization is degenerate with respect to our chosen metric; any two points in this space, $(\pi,\alpha,\beta)$, $(\pi',\alpha',\beta')$, are considered equal as long as $\pi=\pi',\alpha=\alpha'$ and $\alpha\beta=\alpha'\beta'$.    In particular, the total variation distance between these two points is topologically equivalent to the $\mathscr{L}^1$ distance between for triples $(\pi,\alpha,\alpha \beta)$, $(\pi',\alpha',\alpha'\beta')$.  
        \item Sets of parameters.  Let $\Theta,\Theta'$ denote any two sets of parameters.  The Pompeiu-Hausdorff distance is the largest Euclidean distance from any point in either set to the closest point in the other set.
    \end{itemize}
    We are interested in the mapping from the first space to the second induced by observational equivalence classes, i.e.
    \[
    \pi,\alpha,\beta \mapsto \{(\pi,\alpha,\beta'):\ \alpha\beta=\alpha\beta'\}
    \]
    To show that this is discontinuous at some point $\pi,\alpha,\beta$ with respect to our chosen metrics, it suffices to find a sequence $\tilde \pi(\epsilon),\tilde \alpha(\epsilon),\tilde \beta(\epsilon)$ indexed by $\epsilon$ such that that 
    \begin{itemize}
        \item $\tilde \pi(\epsilon),\tilde \alpha(\epsilon),\tilde \alpha(\epsilon)\tilde\beta(\epsilon) \rightarrow \pi,\alpha,\alpha\beta$ in the $\mathscr{L}^1$ metric as $\epsilon\downarrow 0$
        \item $\Theta(\epsilon)\triangleq \{(\pi(\epsilon),\alpha(\epsilon),\beta'):\ \alpha(\epsilon)'\beta'=\gamma\}$ fails to converge to $\Theta(0)\triangleq \{(\pi,\alpha,\beta):\ \alpha\beta'=\gamma\}$ in the Pompeiu-Hausdorff metric.
    \end{itemize}
    We exhibit such a case.  Fix any $\pi$ strictly on the interior of the simplex.  Let $\Theta(0)=\{(\pi,\alpha,\beta'):\ \alpha\beta'=\alpha\beta\}$, where
    \[
    \alpha=\left(\begin{array}{ccc}
    \frac{1}{2} & \frac{1}{4} & \frac{1}{4}\\
    \frac{1}{3} & \frac{1}{3} & \frac{1}{3}
    \end{array}\right)
    \qquad
    \beta=\left(\begin{array}{cc}
    0 & 1\\
    \frac{1}{2} & \frac{1}{2}\\
    \frac{1}{2} & \frac{1}{2}
    \end{array}\right)
    \qquad
    \alpha\beta=\left(\begin{array}{cc}
    \frac{1}{4} & \frac{3}{4}\\
    \frac{1}{3} & \frac{2}{3}
    \end{array}\right)
    \]
    For each $\epsilon>0$ let $\Theta(\epsilon) =\{(\pi,\tilde \alpha(\epsilon),\beta'):\ \tilde \alpha(\epsilon)\beta'=\tilde \alpha(\epsilon) \tilde \beta\}$, where
    \[
    \tilde \alpha(\epsilon)=\left(\begin{array}{ccc}
    \frac{1}{2} & \frac{1}{4}-\frac{\epsilon}{2} & \frac{1}{4}+\frac{\epsilon}{2}\\
    \frac{1}{3} & \frac{1}{3}-\frac{\epsilon}{3} & \frac{1}{3}+\frac{\epsilon}{3}
    \end{array}\right)\qquad
    \tilde \beta=\left(\begin{array}{cc}
    0 & 1\\
    1 & 0\\
    0 & 1
    \end{array}\right)\qquad
    \tilde \alpha(\epsilon) \tilde \beta=\left(\begin{array}{cc}
    \frac{1}{4} -\frac{\epsilon}{2}& \frac{3}{4}+\frac{\epsilon}{2}\\
    \frac{1}{3}-\frac{\epsilon}{3} & \frac{2}{3}+\frac{\epsilon}{2}
    \end{array}\right)
    \]
    Now observe:
    \begin{itemize}
        \item $\pi,\tilde \alpha(\epsilon),\tilde \alpha(\epsilon)\beta \rightarrow \pi,\alpha,\alpha\beta$ in $\mathscr{L}^1$ metric as $\epsilon\downarrow 0$
        \item $\Theta(\epsilon)$ fails to converge to $\Theta(0)$.  Indeed consider any $\beta' \in S(\epsilon)$.  Algebraic manipulations yield that $\beta_{11}'=2(\beta_{21}'-1)\epsilon$ for every $\beta' \in S(\epsilon)$.  This is a negative value unless $\beta_{21}\geq 1$, ergo $\beta' \in S(\epsilon)$ is only a valid right-stochastic matrix if $\beta_{21}'=1$.  On the other hand, $\beta \in S(0)$ satisfies $\beta_{21}=1/2$.  It follows that the Pompeiu-Hausdorff from $\Theta(0)$ to $\Theta(\epsilon)$ is at least $1/2$ for every $\epsilon>0$.
    \end{itemize}

    This shows that the mapping is discontinuous at $\pi,\alpha,\beta$.  Furthermore note that the rows of $\alpha$ are linearly indendent.  Moreover, $\pi$ lies strictly on the interior of the simplex, as does every row of $\alpha$ and $\alpha\beta$.  Thus the distribution corresponding to $\pi,\alpha,\beta$ lies strictly on the interior of the natural parameter space.
\end{proof}

\begin{thm*}[Consistency for positive $\beta$]  Let the rows of $\alpha$ be linearly independent and let $\beta_{x,y}\geq c>0$.  Define $\gamma$ as the $N_L\times N_Y$ right-stochastic matrix
\[
\gamma_{\ell y} \triangleq \sum_{x} \alpha_{_\ell x} \beta_{x,y}
\]
Let $\hat \pi, \hat \alpha, \hat \gamma$ denote any consistent estimators for $\pi,\alpha,\gamma$.  Consider $\{\beta':\ \hat \alpha\beta'=\hat \gamma \}$, the equivalence class for the observations $\DY$ after fixing the estimate $\hat \alpha$.  The distance between the point $\beta$ and this set converges to zero in probability.  
\end{thm*}

\begin{proof}
Fix any $\alpha,\beta$.  It suffices to show that by ensuring $\left|\alpha'-\alpha\right|<\delta,\left|\gamma'-\alpha \beta\right|<\delta$ for $\delta$ sufficiently small we can ensure that $\beta$ is arbitrarily close to some valid right-stochastic matrix in the set $\{\beta':\ \alpha'\beta' = \gamma'\}$.  Let $\tilde{\beta}$ denote the orthogonal projection of $\beta$ to this set.  Note that since the rows of $\alpha$ are independent we can ensure the rows of $\alpha'$ are also independent by taking $\delta$ sufficiently small.  In particular, we can bound the spectral norm of the pseudoinverse of $\alpha'$, and thereby ensure that $\left|\tilde{\beta}_{x,y}-\beta_{x,y}\right|<\epsilon$ for every $x,y$ by taking $\delta$ sufficiently small. Finally, the positivity of $\beta$ allows to ensure that $\tilde \beta$ is a valid right-stochastic matrix.  Thus $\tilde \beta$ is indeed a point in the observational equivalence class which is arbitrarily close to $\beta$.  
\end{proof}

\begin{thm*}[Consistency of $\hat \Theta$] Let $n,m\rightarrow\infty$ such that $|\log(n/m)|<M$.  Then:
\begin{enumerate}
    \item $\hat \Theta$ is asymptotically guaranteed to contain the true parameter values.  
    \item $\hat \Theta$ concentrates for $\pi,\alpha,\alpha\beta$. Let $\xi(\pi',\alpha',\beta')$ denote the maximum of the Euclidean differences $|\pi-\pi'|$, $|\alpha-\alpha'|$, $|\alpha\beta-\alpha'\beta'|$.  Then $\max_{\theta \in \hat \Theta} \xi(\theta)$ converges to zero in probability.   
\end{enumerate}
\end{thm*}
\begin{proof} A corollary of Lemma \ref{lem:sqrtset}, below.  A minor correction is necessary to deal with the fact that we have two sets of independent samples instead of a single list of independent samples.  Fix any $n'>0,m'>0$.  Arrange the $\DX,\DY$ into $T$ independent groups of samples, each containing $n'>0$ samples from $\DX$ and $m'>0$ samples from $\DY$.  Drop samples at the end that cannot be fit into even groups.  Since we have required $|\log(n/m)|<M$, note that can obtain an arbitrarily large number of such groups in the limit.  We can now understand our data as a list of $T$ independent samples drawn from a distribution $p^\Pi$.  Here 
\[
p^\Pi(\ell_1,x_1\cdots \ell_{n'},x_{n'},\ell_{n'+1},y_{n'+1}\cdots \ell_{n'+m'},y_{n'+m'})
\]
is defined as the product of the corresponding distributions arising from $p$.

Now apply Lemma \ref{lem:sqrtset} to this grouped data in the limit as $T\rightarrow \infty$, where the set $S$ indicates the image of our parameterization of $p$. The lemma shows that the true parameter $\pi,\alpha,\beta$ is asymptotically guaranteed to lie in the set $\hat \Theta$; this proves the first claim of our theorem.  Second, the theorem shows that $\pi',\alpha',\beta'$ will be asymptotically excluded as long as $p^\Pi(\cdot; \pi',\alpha',\beta') \neq p^\Pi(\cdot; \pi,\alpha,\beta)$.  Observe that $p^\Pi(\cdot; \pi',\alpha',\beta') \neq p^\Pi(\cdot; \pi,\alpha,\beta)$ if and only if $\pi',\alpha',\beta'$ lies in the true observational equivalence class, i.e. $\pi=\pi',\alpha=\alpha',\alpha\beta=\alpha'\beta'$. Thus any set of parameters whose Kullback-Leibler divergence from the truth is bounded away from zero will asymptotically fail to intersect $\hat \Theta$.  To conclude our proof of the second claim, note that Pinsker's inequality lower-bounds this divergence by the Euclidean distance on the space of the observable distributions $\pi,\alpha,\alpha\beta$.
\end{proof}

\begin{lemma}  \label{lem:sqrtset} Let $(W_1 \cdots W_n)\ \mathrm{i.i.d} \sim p^*$, where $p^*$ is a probability mass function on a finite number of atoms.  Let $\LL(p)=\log\sum_i \log p_{W_i}$ denote the log likelihood for any $p$ on the simplex.  Let $S$ denote any subset of the simplex with $p^* \subset S$.  Let $\hat p = \arg \max_{p \in S} \LL(p)$ denote a maximum likelihood estimator.  Fix any $\delta>0$ and define 
\[
\hat \Theta = \{\theta :\ \LL(\theta) > \LL(\hat \theta) - \delta \log n\}
\]
Then:
\begin{enumerate}
    \item The probability that $p^* \in \hat \Theta$ is asymptotically unity.
    \item Let $D(p^*||p)$ denote the Kullback-Leibler divergence from $p^*$ to $p$ for any $p$ on the simplex.  Let $\Gamma(\epsilon)= \{p:\ D(p^*||p)>\epsilon\}$.  The probability that $\Gamma(\epsilon)$ and $\hat \Theta$ intersect asymptotically vanishes for any $\epsilon>0$.
\end{enumerate}
\end{lemma}
\begin{proof}  The proof is straightforward because we have required $p$ is a probability mass function with a finite number of atoms.  
\begin{enumerate}
    \item $p^* \in \hat \Theta$ with high probability. Since $\log n\rightarrow \infty$ it suffices to show that $\limsup_{n\rightarrow\infty} \mathbb{E}[\LL(\hat p)- \LL(p^*)]$ is bounded.  The only slight subtlety here is that the estimator $\hat p$ is required to lie in the arbitrary set $S$ -- otherwise the usual asymptotic theory would immediately give that $\LL(\hat p)- \LL(p^*)$ is asymptotically $\chi^2$ with $\#\{x:\ p(x;p^*)>0\}-1$ degrees of freedom.  However, since $\hat p$ arises from optimizing over a \emph{restricted} domain, it follows that $\LL(\hat p)- \LL(p^*)$ is actually stochastically dominated by this same $\chi^2$ distribution.  Thus the expectation in question is bounded.

    \item Incorrect parameters are excluded.  First observe that the result holds pointwise, i.e. for any $\epsilon>0,p \in \Gamma(\epsilon)$ we have that $p$ is asymptotically excluded from $\hat \Theta$.  To see this, observe that the central limit theorem yields, asymptotically, that
    \[
    V(p) \triangleq \frac{1}{\sqrt{n}} (\LL(p^*) - \LL(p) - nD(p^*||p)) \sim \mathcal N\left(0,\sigma^2_p\right)
    \]
    where $\sigma^2_p$ is the variance of the corresponding single-sample estimator for the KL divergence.   Thus   
    \begin{align*}
    \mathbb{P}\left(p \in \hat \Theta\right) 
        &=\mathbb{P}\left((\LL(\hat p) - \LL(p^*))+ (\LL(p^*) - \LL(p) -nD(p^*||p)) < \delta \log n - nD(p^*||p)\right)\\ 
        &=\mathbb{P}\left(V(p) < \delta \frac{\log n}{\sqrt n} - \sqrt{n}D(p^*||p) - (\LL(\hat p) - \LL(p^*))/\sqrt{n}\right) 
    \end{align*}
    This probability vanishes whenever $D(p^*||p)>0$, since the right-hand-side of the inequality inside the probability expression escapes to $-\infty$.  Thus every $p \in \Gamma$ is eventually excluded from $\hat \Theta$.  To conclude our proof of the second claim, note that $\hat \Theta$ is a convex set whose eccentricity is bounded by the ratio between the lowest and highest nonzero value in the empirical distribution of the $X$s; since there are a finite number of atoms this eccentricity is thus asymptotically bounded.  Moreover, note that $p \mapsto D(p^*||p)$ is convex and uniformly continuous in neighborhoods of $p^*$.   We can therefore find a grid of points $p\in \Gamma(\epsilon/2)$ whose convex hull contains $p^*$, such that any convex set which does not intersect those points must lie in the complement of $\Gamma(\epsilon)$.  By applying the pointwise result to this finite collection of points, we obtain our uniform result.

\end{enumerate}
\end{proof}

\end{document}

### 2019-04-18 paper/paper.tex
Typos

@@
 Unfortunately, the $N_L<N_X$ regime is precisely the one which motivated our present work, inspired by investigation of cell-types in neuroscience.
-This result shows that perfect estimation of $\beta$ may not be possible in our case.
+This result shows that consistent estimation of $\beta$ may not be possible in our case.
 
@@
 \hfill{}\begin{minipage}{.9\textwidth}
-\caption{\textbf{Cell-type science: maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
+\caption{\textbf{Cell-type science.  Maximum likelihood estimation of $\beta$ suggests unlikely results about cell biology; Markov Link Intervals indicate plausible bounds.}  \label{fig:allen}
 %
-We performed unpaired measurement integration for two ways of measuring the `type' of mouse cells.
+We performed unpaired measurement integration for two ways of measuring the type of cells from mouse visual cortex.
 The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.
@@
 However, Theorem \ref{thm:consistency} shows that the Markov Link Interval $[\hat \beta^{-}_{x,y},\hat \beta^{+}_{x,y}]$ is asymptotically guaranteed to include the truth.
-Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ cell indicates the value of the corresponding parameter.
-The values of $\hat \beta$ are essentially nonsensical in many cases.
+Above we show all three of these estimators for a selection of values of $(x,y)$; we visualize these values in a grid, where rows correspond to values of $x$, columns correspond to values of $y$, and the color of the square in the $(x,y)$ entry indicates the value of the corresponding parameter.
+The values of $\hat \beta$ are nonsensical in many cases.
 On the other hand, the Markov Link Interval suggests plausible bounds for the parameters and also indicates further experiments which would help refine these bounds.
@@
 This integration would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.
-To facilitate this integration, we can construct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.
+To facilitate this integration, one can conduct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.
 In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.
@@
 The Allen Institute recently conducted a series of experiments using both traditional RNA sequencing as well as the Patch-seq technique \cite{tasic2017shared}.
-The same Cre/Lox systems were used for both techniques, thereby allowing us to observe the indicator $L$ for each sample in both datasets.
-For each sample in the traditional RNA sequencing dataset (``Standard'') the cell-type was determined.
-For each sample in the Patch-seq dataset we also have an estimated cell-type, although the Patch cell-type estimates are only given at a coarse level of the taxonomy of cells.
-For example, in the Standard dataset a distinction is made between cell-type `Lamp5 Egln 1' and cell-type `Lamp5 Pdlim5,' whereas in the Patch-seq dataset we group all GABAergic neuronal cell-types characterized by high expression of `Lamp5' into a single coarse `cell class.'
+The same Cre/Lox systems were used for both techniques.
+For each sample in the traditional RNA sequencing dataset we observe both the cell-type ($X$) and the Cre/Lox indicator ($L$).
+For each sample in the Patch-seq dataset we observe a both noisy estimate of the cell-type ($Y$) and the Cre/Lox indicator ($L$).
+% Note that the Patch cell-type estimates are only given at a coarse level of the taxonomy of cells.
+For example, in the traditional RNA sequencing dataset a distinction is made between cell-type `Lamp5 Egln 1' and cell-type `Lamp5 Pdlim5,' whereas in the Patch-seq dataset we group all GABAergic neuronal cell-types characterized by high expression of `Lamp5' into a single coarse `cell class.'
 
 We applied the Markov Link Intervals to the Allen Institute data try to determine the reliability of the noisy cell-type estimators $Y$ arising from Patch-seq approach.
-We can visualize a maximum likelihood estimator as well as the Markov Link Intervals through heatmaps (see Figure \ref{fig:allen})).
-We see that the maximum likelihood estimator sometimes gives essentially nonsensical results.
+We can visualize a maximum likelihood estimator as well as the Markov Link Intervals through heatmaps (see Figure \ref{fig:allen}).
+We see that the maximum likelihood estimator sometimes gives nonsensical results.
 For example, the maximum likelihood estimator suggests that a particular group of excitatory glutamatergic neurons from visual cortex are \emph{always} assigned to a GABAergic inhibitory type by the $Y$ variable.
@@
 For example, the lower bounds of the Markov Link Interval suggests that a sizeable fraction of `Lamp5 Pdlim5' cells are indeed classified as `Lamp5' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as `Lamp5 Pdlim5', but instead only gives coarse class-level cell-type predictions, such as `Lamp5').
-The upper bounds furthermore show that these cells are seldom classified as anything \emph{but} `Lamp5' cells or possibly the `unknown' type which the Patch method assigns to cells which it cannot identify.
-For cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type; the intervals indicate that the Patch method does \emph{not} classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate  how they \emph{are} being classified.
+The upper bounds show that these cells are not commonly categorized into any group besides the `Lamp5' type and the the `unknown' type which the Patch method assigns to cells which it cannot identify.
+Consider another example: for cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type.
+The intervals indicate that the Patch method does not typically classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate how they are being classified.
 At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.
-These `Lamp5 Elgn3-1' cells might get perfectly classified or they may get poorly classified; the available data cannot tell us either way.
+It is possible that the Patch method perfectly classifies these `Lamp5 Elgn3-1' cells; it is possible that it always misclassifies them; the available data cannot tell us either way.
 
 Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.
-For example, the significant ambiguity for cells with standard type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.
-If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which measure as Patch type `Pvalb,' this would show that the Standard `Lamp5 Egln 1' is not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells but \emph{only} cells with Patch type `Pvalb,' this would show the opposite.
+For example, the significant parameter ambiguity for cells with type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.
+If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which Patch categorizes as `Pvalb,' this would show that `Lamp5 Egln 1' cells are not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells and also selects cells which are largely categorized as `Pvalb' by Patch, this would show the opposite.
 If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.

### 2019-05-10 paper/paper.tex
New structure!

@@
 
-\title{Unpaired measurement integration using Markov Link Intervals}
+\title{Measurement linkage using Markov Link Intervals}
 
@@
 \begin{abstract}
-We consider two datasets, each measuring a different of observations on a common population.
+We consider two datasets, each taking different kinds of measurements from samples of a common population.
 The first dataset is constructed by drawing individuals randomly from this population and measuring a quantity about each individual.
 The second dataset is constructed similarly, but a different quantity is measured.
-``Unpaired measurement integration'' attempts to use this data to infer the joint distribution of the two different quantities.
-We overcome the unidentifiability problems in this task by measuring an additional quantity, similar to an instrumental variable.
-This quantity is observed in both datasets and satisfies a conditional independence assumption.
-We characterize identifiability in the presence of these variables and develop a consistent estimator.
-The performance of the estimator is tested in simulations.
+``Unpaired measurement linkage'' attempts to use this data to infer the joint distribution of the two different quantities.
+We study this problem in the presence of a third quantity which is observed in both datasets and satisfies a conditional independence assumption.
+We characterize identifiability in this case and develop estimators for the joint distribution.
+The performance of the estimators is tested in simulations.
 The method is applied to transcriptomic measurements to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.
@@
 
-Unpaired measurement integration seeks to infer the joint distribution on two variables $(X,Y)$ using knowledge of the marginal distributions on $X$ and $Y$.
-This problem is ill-posed, but practical algorithms have nonetheless been developed.
-An early approach by Andrey Kolmogorov noted that the expectation for any linear function $f(X,Y)=\alpha X+\beta Y$ could be effectively bounded using only the marginal distributions \cite{makarov1982estimates}.
-Bounds on arbitrary univariate functions $f(X,Y)$ have also been developed in the context of causal inference \cite{balke1994counterfactual,FanYan2017}.
-More recently, the CycleGan algorithm has gained prominence in the computer graphics community for succesfully estimating a joint distribution between photographs ($X$) and watercolor paintings ($Y$) of the same subject; CycleGan accomplishes this unpaired measurement integration using two completely separate datasets, one of photographs and one of watercolor paintings \cite{zhu2017unpaired}.
+Many scientific experiments produce data that come from different measurement techniques.
+To use such data, we need to understand the link between the different types of measurements.
+For example, in a biological context, a transcriptomic-morphological link would answer a question like:  ``We have measured the transcriptome of this cell -- what might its morphology have been?''
 
-To resolve the fundamental ill-posedness of unpaired dataset integration, some form of extra information is often employed.
-We assume this side-information variable can be observed for each sample in both datasets.
-The work of Fan \emph{et al} (2017) gives a detailed overview for this case \cite{FanYan2017}.
-Fan characterizes the asymptotics of estimators for $\mathbb{E}[f(X,Y)]$, where $f$ can takes a wide variety of forms.
-Estimator asymptotics are established under many different assumptions on the relationship between the measurements and the side-information
+It is particularly challenging to estimate the link when multiple measurements of the same specimen are unavailable.
+For example, it is currently often difficult to observe both the RNA expression and morphological properties of a single cell.
+Rather, we typically have two datasets: one contains the transcriptomes of one group of cells and the other contains the morphology of another group of cells from the same population.
+However, the datasets are not completely disconnected: there is some information which is collected for every specimen in both datasets.
+For example, it may be possible to read optogenetic marker information for every cell in both datasets.
+The task is to leverage this information to uncover the link.
 
-Inspired by problems in the study of cell populations, this paper considers a new kind of assumption about the side-information.
-Let $L$ denote the side-information.
-Suppose that $L$ and $Y$ are conditionally independent given $X$.
-This assumption holds naturally whenever both $L$ and $Y$ can both be understood as independent noisy measurements of $X$.
-We give a complete real-world example in Section \ref{sec:allen}.
-The variable $L$ bears a formal similarity to an instrumental variable, but its function is different: instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.
+This \emph{unpaired measurement linkage} problem arises in many scientific and technological fields where it is difficult to take multiple types of measurements of the same specimen.
+Formally, unpaired measurement linkage seeks to estimate a distribution $p(\ell,x,y)$ even though we can never observe a complete sample $(L,X,Y)\sim p$.
+For every sample we can observe $(L,X)$ or we can observe $(L,Y)$ -- but we can never observe $(X,Y)$.
+In the cell biology example above, $X$ might correspond to a cell's transcriptome, $Y$ might correspond to the cell's morphology, and $L$ might correspond to an optogenetic marker that can be observed for every cell.
 
-Inference under the conditional independence assumption must overcome two obstacles: the datasets only allow approximate estimation for the marginal distributions, and the marginal distributions only give a partial picture of the joint.
-These problems cannot be considered separately.
-If we treat statistical estimates as exact truth and proceed with a purely algebraic analysis of the model, inference can go badly wrong.
-It turns out that even infinitesimal estimation errors can result in substantially incorrect conclusions, as we show in Lemma \ref{lem:noobseq}.
-This marks a stark contrast from the conditions considered in Fan \emph{et al} (2017), where additional statistical noise results in a graceful degradation of inference accuracy.
+The key challenge in unpaired measurement linkage is to identify good assumptions or inductive biases; without any additional assumptions the problem is always ill-posed.
+An overview of different possible assumptions can be found in Fan \emph{et al} \cite{FanYan2017}.
 
-In this paper, we overcome both of these challanges for the discrete setting.
-When $L,X,Y$ are categorical distributions it is possible to make a detailed study of what are known as the observational equivalence classes \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.
+In this paper, we seek to understand unpaired measurement linkage in the discrete setting under the assumption that $L$ and $Y$ are conditionally independent given $X$.
+This assumption is inspired by cell biology, and holds naturally whenever $L$ and $Y$ can be understood as independent noisy measurements of $X$.
+For example, in Section \ref{sec:allen} we will see an example where optogenetic markers ($L$) and a low-fidelity transcriptomic record ($Y$) act as independent noisy indicators about cell-type ($X$).
+Many other examples can be found in Section \ref{sec:conclusions}.
+Note that the side-information $L$ bears a formal similarity to an instrumental variable, but its function is different; instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.
+
+We investigate the identifiability issues and develop estimators which are asymptotically guaranteed to bound the true joint distribution.
+The key idea is to make a detailed study of the ``observational equivalence classes'' \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.
 Each class corresponds to a set of parameters that give rise to the same distribution on the data we observe.
-We can achieve consistent inference by taking a carefully chosen dilation of these sets.
-We design a simple optimization algorithm to find extremal points in these dilated sets.
+We show that estimating these sets is actually impossible, but a carefully-chosen dilation of these sets asymptotically contains the true model parameters and also concentrates on the observable distributions.
+We also design a simple optimization algorithm to find extremal points in these dilated sets.
 These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.
-Performance is investigated through a variety of simulation studies.
-Finally, the estimator is applied to uncover the relationship between two different technologies for measuring gene expression in mouse neurons.
+Finally, we interrogate the effectiveness of these intervals using simulations and real-world data.
+
+This work arises out of a literature that spans many fields and many decades.
+Perhaps the earliest approach to unpaired measurement integration comes from Andrey Kolmogorov, who conjectured an inequality for the cumulative distribution function of the sum of two random variables in terms of their marginal distributions \cite{makarov1982estimates}.
+Related inequalities have also been studied in the context of quantum physics \cite{chaves2014inferring}.
+More recently, bounds on arbitrary univariate functions $f(X,Y)$ have also been developed in the context of causal inference \cite{balke1994counterfactual}.
+Most recently, the CycleGan algorithm has gained prominence in the computer graphics community for successfully estimating a joint distribution between photographs ($X$) and watercolor paintings ($Y$) of the same subject; CycleGan accomplishes this unpaired measurement linkage using two completely separate datasets, one of photographs and one of watercolor paintings \cite{zhu2017unpaired}.
+This work connects most directly to the work of Fan \emph{et al} (2017), which gives an extensive study for unpaired measurement linkage problem under many different assumptions \cite{FanYan2017}.
+Under each assumption the asymptotics of estimators for $\mathbb{E}[f(X,Y)]$ are characterized, where $f$ can takes a wide variety of forms.
+Here we extend this work by considering similar questions under a new kind of assumption.
 
@@
 
-\section{Formal problem statement and notation}
+\section{Markov Link Intervals (MLI)}
 
@@
 %
-Our task is to use the data $(\DX,\DY)$ to estimate the parameters of $p$, namely $(\pi,\alpha,\beta)$.
-Estimating $\pi,\alpha$ from this data is fairly straightforward.
-However, existing estimation theory is less helpful for the parameter $\beta$.
-In this paper we develop theory and practice for estimating this parameter.
+Unpaired measurement linkage seeks to use the data $(\DX,\DY)$ to estimate the parameters of $p$, namely $(\pi,\alpha,\beta)$.  Observe that estimating $\pi,\alpha$ from this data is straightforward.
+
+Existing estimation theory is less helpful for the parameter $\beta$.  In this paper we develop theory and practice for estimating $\beta$.
 
@@
 
-\section{Markov Link Intervals (MLI) for estimating $\beta$}
+\subsection{Theory}
 
@@
 
-\begin{lemma}[Unidentifiability]  \label{lem:unident} When $N_L<N_X$ there is no generally consistent estimator to estimate $\beta$ from the data $(\DX,\DY)$.  
+\begin{lemma}[Unidentifiability]  \label{lem:unident} When $N_L<N_X$ there is no generally consistent estimator to infer $\beta$ from the data $(\DX,\DY)$.  
 \end{lemma}
@@
 Define
-\[
-\hat \Theta=\left\{\hat \pi, \hat \alpha,\hat \beta:\ \LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta) > \LL^* - \log(n+m)\right\}
-\]
+\begin{gather}
+\hat \Theta=\left\{\hat \pi, \hat \alpha,\hat \beta:\ \LL(\DX,\DY;\hat \pi,\hat \alpha,\hat \beta) > \LL^* - \log(n+m)\right\} \label{eq:magictheta}
+\end{gather}
 This log-scaled likelihood envelope has two desirable properties: it both concentrates around the true distribution on the observables and is asymptotically guaranteed to contain the true model parameters.
@@
 \end{thm}
-This theorem indicates a practical solution for estimating $\beta$.  For each $x,y$, we define the \emph{Markov Link Interval} as 
+
+\subsection{Practice}
+
+To estimate the parameters $\beta_{x,y} = p(y|x)$, we propose \emph{Markov Link Intervals}.  For each $x,y$ we define this interval as 
 \[
@@
 \]  
+where, as above, $\hat \Theta$ is the set of parameters whose likelihood is within $\log n+m$ of the maximum likelihood.
 Theorem \ref{thm:consistency} shows that $\beta_{x,y}$ is asymptotically included in the interval $[\hat \beta^-_{x,y},\hat \beta^+_{x,y}]$.
 
-To calculate the Markov Link Interval we must solve the associated optimization problems.
+To calculate the Markov Link Intervals from data, we must solve the associated constrained optimization problems.
 Due to the high dimensionality of the problem and the ill-conditioning of the matrices involved, we were unable to tune off-the-shelf optimization methods to reliably solve these problems.
-We developed an alternative optimization procedure which relies on a minorization-maximization approach and requires no tuning parameters.
+We developed an alternative optimization procedure which requires no tuning parameters.
+The key idea is that the Lagrangian of this problem can itself be understand as a penalized likelihood; for any fixed value of the Lagrange multiplier, the likelihood can be thus be optimized using an Expectation Maximization algorithm.
+The correct value of the Lagrange multiplier can be found with a simple line-search.
 Appendix \ref{sec:mlmdetails} provides a detailed outline of the algorithm.\footnote{Code is published at https://github.com/jacksonloper/markov-link-method, including a tutorial ipython notebook that details every computation made in this paper.}
@@
 This average total variation distance is computed by averaging the row-by-row total variations between $\beta$ and  $\hat \beta$.
-After performing a series of such trials, we visualize the results to obtain qualitative understanding of the performance of the maximium likelihood estimator $\hat \beta$.
+After performing a series of such trials, we visualize the results to obtain qualitative understanding of the performance of the maximum likelihood estimator $\hat \beta$.
 
@@
 If $X,Y$ are independent or deterministically related, we see that perfect estimation becomes possible in the limit of infinite data.
-However, in the more typical case where the relatioship between $X$ and $Y$ neither independent nor deterministic, it appears that consistent estimation is not possible.
+However, in the more typical case where the relationship between $X$ and $Y$ neither independent nor deterministic, it appears that consistent estimation is not possible.
 
@@
 %
-We performed unpaired measurement integration for two ways of measuring the type of cells from mouse visual cortex.
+We performed unpaired measurement linkage for two ways of measuring the type of cells from mouse visual cortex.
 The first method uses gene expression to obtain a definitive measurement of the cell-type, $X$.
@@
 For any cell it is thus impossible to measure a rich variety of cell properties \emph{and} reliably estimate the cell-type.
-To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement integration.
-This integration would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.
-To facilitate this integration, one can conduct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.
+To understand the relationship between cell-type and these other cell properties, we therefore need some form of unpaired measurement linkage.
+This linkage would allow us to determine the reliability of the noisy estimates $Y$, and thereby determine whether the relationship between cell-type and other cell properties can be reliably determined from the Patch-seq approach.
+To facilitate this linkage, one can conduct experiments which make restrictions on which cells are are picked out, using a Cre/Lox system \cite{tasic2017shared}.
 In each experiment, the probability that a particular cell will be picked out is entirely determined by the cell type $X$, and is thus statistically isolated from the noisy cell-type estimate $Y$.
@@
 The Markov Link Intervals yield plausible bounds on the true model parameters.
-For example, the lower bounds of the Markov Link Interval suggests that a sizeable fraction of `Lamp5 Pdlim5' cells are indeed classified as `Lamp5' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as `Lamp5 Pdlim5', but instead only gives coarse class-level cell-type predictions, such as `Lamp5').
-The upper bounds show that these cells are not commonly categorized into any group besides the `Lamp5' type and the the `unknown' type which the Patch method assigns to cells which it cannot identify.
-Consider another example: for cells of type `Sst Myh8-1',  Markov Link Intervals give us considerably less certainty regarding the related type.
-The intervals indicate that the Patch method does not typically classify these cells as `CR,' `L6b,' `Meis2,' `Sncg,' or `Vip' -- but it does not indicate how they are being classified.
+For example, the lower bounds of the Markov Link Interval suggests that a sizable fraction of ``Lamp5 Pdlim5'' cells are indeed classified as ``Lamp5'' cells by the Patch method (note that the Patch method does not give fine-grained cell-type predictions, such as ``Lamp5 Pdlim5'', but instead only gives coarse class-level cell-type predictions, such as ``Lamp5'').
+The upper bounds show that these cells are not commonly categorized into any group besides the `Lamp5' type and the the ``unknown'' type which the Patch method assigns to cells which it cannot identify.
+Consider another example: for cells of type ``Sst Myh8-1'',  Markov Link Intervals give us considerably less certainty regarding the related type.
+The intervals indicate that the Patch method does not typically classify these cells as ``CR,'' ``L6b,'' ``Meis2,'' ``Sncg,'' or ``Vip'' -- but it does not indicate how they are being classified.
 At the extreme end, we have cells of type `Lamp5 Elgn3-1,' for which the Markov Link Intervals are completely vacuous.
-It is possible that the Patch method perfectly classifies these `Lamp5 Elgn3-1' cells; it is possible that it always misclassifies them; the available data cannot tell us either way.
+It is possible that the Patch method perfectly classifies these ``Lamp5 Elgn3-1' cells; it is possible that it always misclassifies them; the available data cannot tell us either way.
 
 Parameters with large intervals suggest new experiments to perform in order to more closely determine the model parameters.
-For example, the significant parameter ambiguity for cells with type `Lamp5 Egln 1' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.
-If we could find a Cre/Lox procedure that obtained many `Lamp5 Egln-1' cells but no cells which Patch categorizes as `Pvalb,' this would show that `Lamp5 Egln 1' cells are not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells and also selects cells which are largely categorized as `Pvalb' by Patch, this would show the opposite.
+For example, the significant parameter ambiguity for cells with type ``Lamp5 Egln 1'' suggests we need a series of experiments with more distinct Cre/Lox experimental parameters that pick out cells of this kind.
+If we could find a Cre/Lox procedure that obtained many ``Lamp5 Egln-1'' cells but no cells which Patch categorizes as ``Pvalb,'' this would show that ``Lamp5 Egln 1'' cells are not associated with Patch type `Pvalb.'  On the other hand, if we could find Cre/Lox parameters that select many `Lamp5 Egln 1' cells and also selects cells which are largely categorized as `Pvalb' by Patch, this would show the opposite.
 If these additional experiments were performed, new Markov Link Intervals could be computed on the new data to determine what aspects of the link are still ambiguous.
@@
 
-We here study `unpaired measurement integration,' the problem of estimating a joint distribution $(X,Y)$ from two separate datasets, one with samples of $X$ and another with samples of $Y$.
+\label{sec:conclusions}
+
+We here study ``unpaired measurement linkage,'' the problem of estimating a joint distribution $(X,Y)$ from two separate datasets, one with samples of $X$ and another with samples of $Y$.
 Rigorous statistical analysis of this ill-posed problem are complicated by statistical error and identifiability issues.
 Here we characterize both issues in the discrete settings when we can observe a third variable $L$ for each sample in both dataset, under the assumption that $L$ and $Y$ are independent conditioned in $X$.
-This assumption arises naturally if $L$ gives noisy measurement about $X$ and that noise is independent of $Y$.
-Under this assumption, we construct ``Markov Link Intervals'' which give provably consistent bounds on the joint distribution.
+
+This assumption may hold under a variety of circumstances:
+
+\begin{itemize}
+    \item Quality control for manufacturing.
+Consider a manufacturing process that produces components; to evaluate production quality, they use two different destructive tests.
+The first kind of test is cheap, but its accuracy is unknown.
+A more established test is accurate but expensive.
+The manufacturer is interested in the link: given results from the cheap test ($X$), what might have been the results from the accurate test ($Y$)?
+To estimate the link, we can uses different batches of production ($L$).
+For example, $L$ might indicate the day on which a component was made.
+The assumption: the link doesn't depend on the day the component was made.
+This is reasonable if the tests are performed on different days.
+In this case, Markov Link Intervals can be used to calibrate the two kinds of tests.
+
+    \item Astronomy photography.
+Consider an astronomer taking photographs of stars over time; depending on availability, some pictures are taken using one camera and other pictures are taken using a different camera.
+To standardize the photos, the photographer is interested in the link: given a photo from one camera ($X$), what might the picture have looked like if it was taken with the other camera ($Y$)?
+To estimate the link, the MLM uses different astronomical targets ($L$).
+For example, $L$ might indicate which star is being photographed.
+The assumption: the difference between the cameras is the same for each star.
+This is reasonable as long as the camera settings are set to the same values for each star.
+In this case, Markov Link Intervals can be used to standardize photographs from different cameras.
+
+    \item Personalized medicine.
+Consider two sources of knowledge about how transcriptomes might help us personalize cancer medication.
+We can measure the transcriptome of cancer cells from inside humans, but experimenting on humans is difficult and morally fraught.
+We can also culture immortalized cells from human cancers and measure their transcriptomes;  experimenting on immortalized cells is cheap and straightforward.
+To generalize work on immortalized cells to human patients, we need the link: given a cultured cell's transcriptome ($X$), what can we expect about the corresponding in-vivo transcriptome ($Y$)?
+To estimate the link, we can use different types of cancer ($L$).
+The assumption: the transcriptomic effects of culturing the cells is the same for each cancer type.
+This may be reasonable as long as the cancer types are sufficiently similar.
+In this case, the Markov Link Intervals  can be used to help generalize results from immortalized cells to real human patients.
+
+    % \item Text/image correspondence.
+Consider the problem of automatic captioning.
+We can scrape images of people from the internet.
+We can also find descriptions of people from literature.
+Automatic captioning requires the link: given an image ($X$), what might its text description look like ($Y$)?
+To estimate the link, the MLM can use different subject material ($L$).
+For example, $L$ might indicate the climate of the location where the photograph was taken.
+The assumption: the link is the same for people from different climates.
+This is reasonable as long as the photograph itself indicates the important climate information, thereby creating an independence between $Y$ and $L$ after conditioning on $X$.
+In this case, the MLM can be used to perform automatic captioning.
+
+    \item Replication crisis and lab effects.
+We would like to understand the differences in how two labs perform experiments.
+These differences can be understood by looking at the link: given results from one lab ($X$), what results could we expect from another lab ($Y$)?
+To estimate the link, we can use different specimen batches ($L$).
+For example, $L$ might indicate a batch of mice, half of which were sent to one lab and half of which were sent to another lab.
+The assumption: the lab effects are the same for each batch of mice.
+    % This is reasonable as long as the order in which the batches are processed is the same for both labs.  
+    In this case, Markov Link Intervals can be used to discover ``lab effects'' that impede successful experimental replication.
+\end{itemize}
+
+When the conditional independence assumption holds, Markov Link Intervals  give provably consistent bounds on the joint distribution.
 Simulations bear out the consistency of these bounds and also suggest estimation is easiest performs best when $L$ is predictive of $X$ and the relationship between $X$ and $Y$ is either deterministic or completely independent.
-An application of the method to cell-type populations demonstrates the method's practical utility for unpaired measurement integration problems.
+An application of the method to cell-type populations demonstrates the method's practical utility for unpaired measurement linkage problems.
 One can imagine that these results could be extended to the continuous setting, though additional regularity conditions would be required.
@@
 \label{sec:mlmdetails}
+
+We here give a detailed explanation of how to compute the Markov Link Intervals from two unpaired datasets.
 
@@
 
-\begin{lemma*}[Unidentifiability]  When $N_L<N_X$ there is no generally consistent estimator to estimate $\beta$ from the data $(\DX,\DY)$. 
+\begin{lemma*}[Unidentifiability]  When $N_L<N_X$ there is no generally consistent estimator to infer $\beta$ from the data $(\DX,\DY)$. 
 \end{lemma*}
@@
     It suffices to show that there exist parameter choices $(\pi,\alpha,\beta)$ and $(\pi',\alpha',\beta')$ where $\beta\neq\beta'$ yet the distribution on $\DX,\DY$ remains the same.
-In this case there would be not way to determine which of these paramer choices was correct, so no consistent estimator could be produced.
+In this case there would be not way to determine which of these parameter choices was correct, so no consistent estimator could be produced.
 Note that the distribution on $\DX,\DY$ is entirely determined by $\pi$, $\alpha$, and $\gamma=\alpha\beta$.

### 2019-05-13 paper/paper.tex
More SCIENCE

@@
 For example, it may be possible to read optogenetic marker information for every cell in both datasets.
-The task is to leverage this information to uncover the link.
+The task is to leverage this common information to uncover the link.
+Figure \ref{fig:bigpicture} sketches an example of this difficult situation, which is common in biological problems \cite{gouwens2018classification}.
 
@@
 The key challenge in unpaired measurement linkage is to identify good assumptions or inductive biases; without any additional assumptions the problem is always ill-posed.
-An overview of different possible assumptions can be found in Fan \emph{et al} \cite{FanYan2017}.
+An overview of many typical assumptions can be found in Fan \emph{et al} \cite{FanYan2017}.
 
-In this paper, we seek to understand unpaired measurement linkage in the discrete setting under the assumption that $L$ and $Y$ are conditionally independent given $X$.
-This assumption is inspired by cell biology, and holds naturally whenever $L$ and $Y$ can be understood as independent noisy measurements of $X$.
-For example, in Section \ref{sec:allen} we will see an example where optogenetic markers ($L$) and a low-fidelity transcriptomic record ($Y$) act as independent noisy indicators about cell-type ($X$).
+In this paper, we investigate a new assumption: that $L$ and $Y$ are conditionally independent given $X$.
+We try to understand theory and practice for unpaired measurement linkage in the discrete setting under this assumption.
+This assumption is inspired by cell biology and holds naturally whenever $L$ and $Y$ can be understood as independent noisy measurements of $X$.
+For example, in Section \ref{sec:allen} we will see an example where optogenetic markers ($L$) and a low-fidelity transcriptomic measurement ($Y$) act as independent noisy indicators about cell-type ($X$).
 Many other examples can be found in Section \ref{sec:conclusions}.
-Note that the side-information $L$ bears a formal similarity to an instrumental variable, but its function is different; instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.
+Note that the side-information $L$ bears a formal similarity to an instrumental variable, but its function is different; instrumental variables resolve causal effects in the presence of unobserved confounders, whereas in our case $L$ permits estimation of joint distributions even though joint observations are unavailable.
+
+\begin{figure}\fbox{\begin{minipage}{\textwidth}
+\begin{tabular}{cc}
+\includegraphics[width=.5\textwidth]{bigpicture.png} 
+& \includegraphics[width=.5\textwidth]{graphicalmodel.png} \\
+%
+\begin{minipage}{.4\textwidth}(a) Setup: each specimen can be measured with one of two procedures,  but can't be measured with both.
+How do we study the relationship  between the procedures?\end{minipage}
+& \begin{minipage}{.3\textwidth}(b) Key modeling assumption: \\ $X$ separates $Y$ from $L$.\end{minipage} 
+\end{tabular}
+\caption{{\bf Markov Link Intervals use a conditional independence assumption to estimate joint distributions from marginal observations.}  In subfigure (a) we have a population of neurons.
+We select half of the neurons and determine what type of neuron they are using one kind of measurement ($X$).
+We use a noisier measurement for the second half of neurons and use it to guess their cell-types ($Y$).
+To reconcile the two resulting notions of cell-type, we represent them as two random variables on a common probability space.
+Markov Link Intervals (defined formally in Section \ref{sec:mlidef}) provides a way to estimate the joint distribution of these two random variables.
+In subfigure (b) we present a graphical model diagram articulating the assumption upon which our analysis is based.
+For each neuron in each group, this diagram considers the cell-type ($X$), the noisy cell-type guess ($Y$), and an optogentic marker ($L$) which can be observed for every cell.
+We assume that the optogenetic marker and the noisy estimate become independent if we condition upon the cell-type; this assumption allows us to estimate the joint distribution.
+\label{fig:bigpicture}}
+\end{minipage}
+}
+\end{figure}
 
@@
 We show that estimating these sets is actually impossible, but a carefully-chosen dilation of these sets asymptotically contains the true model parameters and also concentrates on the observable distributions.
-We also design a simple optimization algorithm to find extremal points in these dilated sets.
+We also design an optimization algorithm to find extremal points in these dilated sets.
 These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.
@@
 \section{Markov Link Intervals (MLI)}
+
+\label{sec:mainMLI}
 
@@
 \subsection{Practice}
+
+\label{sec:mlidef}
 
